### 2025-02-04 products/manuscript/Manuscript.qmd
This is the dataset for my project
I added a dataset along with some information about it for my final project.

---
title: "Yaqubi-MADA-projet"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(naniar)
```

**Authors**

-   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
-   Second Author$^{3, *}$
-   Third Autor$^{3, \dagger}$
-   Last Author$^{1, 2, \land}$

**Author affiliations**

1.  College of Public Health, University of Georgia, Athens, GA, USA.
2.  Another Affiliation, Athens, GA, USA.
3.  Yet another Affiliation.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some\@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

## General Background Information

Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease. It’s also what most people mean when they use the general term “heart disease.” It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart. This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath. As the leading cause of death in the U.S. and globally, CAD claimed 375,500 lives in the U.S. in 2021 and affects over 18 million American adults.

## Description of data and data source

This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses. Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).

This heart disease dataset is acquired from one o f the multispecialty hospitals in India. This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type). This dataset was obtained from mendeley data where it was first published in April of 2021.

This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>

## Questions/Hypotheses to be addressed

Do younger individuals (e.g., age \< 40) with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

For this analysis I will be using target which is the outcome of interest and is described as abcense or pressence of heart disease and some of the predictors of interest are age, serum cholesterol, resting BP, and old peak. The goal for this analysis is to be able observe if young patients with normal cholesterol but high BP/ST depression have higher CVD rates than expected.

The analysis will first isolate younger individuals with clinically normal cholesterol levels to examine their risk of cardiovascular disease (CVD) when other factors like high blood pressure or ST depression are present. Patients will be categorized based on these secondary risk factors, and descriptive analysis will compare CVD prevalence among groups, such as those with normal cholesterol but elevated blood pressure versus those with entirely normal profiles. This step aims to reveal potential associations between non-traditional risk factors and CVD in younger populations.

To validate these findings, statistical tests like chi-square analysis will assess significance, while logistic regression models will quantify the independent and combined effects of blood pressure and ST depression on CVD risk, adjusting for confounders like gender and fasting blood sugar. Visualization techniques, including bar plots and boxplots, will further illustrate risk patterns, making the findings more interpretable. By integrating these methods, the study aims to challenge traditional risk-assessment frameworks and support more nuanced screening approaches for younger individuals at risk of CVD.

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.

This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>

## Data import and cleaning

```{r}
cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))

head(cardio_df)

colnames(cardio_df)

dim(cardio_df)

summary(cardio_df)

glimpse(cardio_df)

gg_miss_var(cardio_df)

print(cardio_df)

view(cardio_df)

```

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-03-14 products/manuscript/Manuscript.qmd
created a different manuscript for the unfinished project codes
Create a new qmd file in the manuscript folder within the products folder to transfer all the codes for the unfinished final projects.

@@
 
+refernece: 
+1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings
+1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703
+2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean
+2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel
+3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/
+3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
+
 To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
@@
 
-print(cardio_df)
+# Transform variables into suitable types and label them according to the data dictionary to enhance readability.
+cardio_df <- cardio_df %>%
+  mutate(
+    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
+    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
+                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
+    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
+    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
+                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
+    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
+    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
+    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
+  )
 
-view(cardio_df)
+```
+
+# Part 2. Exploratory Data Analysis (EDA):
+```{r}
+# Calculate summary statistics for selected variables
+
+summary_stats <- cardio_df %>% 
+  summarise(
+    # Age
+    Age_min = min(age, na.rm = TRUE),
+    Age_max = max(age, na.rm = TRUE),
+    Age_mean = mean(age, na.rm = TRUE),
+    Age_sd = sd(age, na.rm = TRUE),
+
+    # Resting Blood Pressure
+    RestingBP_min = min(restingBP, na.rm = TRUE),
+    RestingBP_max = max(restingBP, na.rm = TRUE),
+    RestingBP_mean = mean(restingBP, na.rm = TRUE),
+    RestingBP_sd = sd(restingBP, na.rm = TRUE),
+
+    # Serum Cholesterol
+    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
+    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
+    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
+    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
+
+    # Maximum Heart Rate
+    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
+    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
+    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
+    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
+
+    # Oldpeak
+    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
+    Oldpeak_max = max(oldpeak, na.rm = TRUE),
+    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
+    Oldpeak_sd = sd(oldpeak, na.rm = TRUE),
+  )
+
+# Print the summary statistics
+print(summary_stats)
+
+# Histogram and density overlay of age distribution
+ggplot(cardio_df, aes(x = age)) +
+  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
+  geom_density(color = "darkblue", linewidth = 1) +
+  labs(title = "Age Distribution", x = "Age", y = "Density") +
+  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
+  theme_bw()
+
+# Histogram and density overlay of serum cholestrol (excluding zeros)
+ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
+  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
+  geom_density(color = "darkgreen", linewidth = 1) +
+  labs(title = "Serum Cholesterol Distribution", 
+       x = "Serum Cholesterol (mg/dL)", y = "Density") +
+  theme_minimal()
+
+# Histogram and density overlay of blood pressure (excluding zeros)
+ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
+  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
+  geom_density(color = "darkred", linewidth = 1) +
+  labs(title = "Resting Blood Pressure Distribution", 
+       x = "Resting Blood Pressure (mmHG)", y = "Density") +
+  theme_bw()
+
+```
+
+# Part 3, Statistical analysis: 
+```{r}
+
+# Filter data for younger individuals (age < 40) with normal cholesterol levels, elevated BP and elevated oldpeak
+cleaned_df <- cardio_df %>%
+  filter(age < 40, serumcholestrol > 0,serumcholestrol < 200) %>%  # Normal if serumcholestrol < 200 mg/dL
+  mutate(
+    elevated_BP = case_when(
+      restingBP < 120 ~ "Low BP",
+      between(restingBP, 120, 129) ~ "Elevated",  # Elevated blood pressure: restingBP between 120 and 129
+      restingBP > 129 ~ "High BP"
+    ),
+    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")  # Elevated ST depression: oldpeak > 1
+  ) %>% 
+  select(age, chestpain, restingBP, serumcholestrol, fastingbloodsugar, maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)  # Select variables of interest
+
+# Summarize the counts by blood pressure status, ST depression status, and CVD risk
+summary_table <- cleaned_df %>%
+  group_by(elevated_BP, elevated_oldpeak, target) %>%
+  summarise(count = n(), .groups = "drop")
+
+print(summary_table)
+
+# Visualize CVD risk by blood pressure and ST depression status
+ggplot(cleaned_df, aes(x = elevated_BP, fill = target)) +
+  geom_bar(position = "dodge") +
+  facet_wrap(~ elevated_oldpeak) +
+  labs(title = "CVD Risk by Blood Pressure and ST Depression Status\n(Individuals Age < 40 with Normal Cholesterol)",
+       x = "Blood Pressure Status",
+       y = "Count",
+       fill = "Heart Disease") +
+  theme_minimal()
 

### 2025-03-14 products/manuscript/Unfinished project codes.qmd
created a different manuscript for the unfinished project codes
Create a new qmd file in the manuscript folder within the products folder to transfer all the codes for the unfinished final projects.

--- 2025-03-14 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Yaqubi-MADA-projet"
-author: ""
-format:
-  docx:
-    toc: false
-    number-sections: true
-    highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
+title: "Final project"
+author: "Murtaza Yaqubi"
+format: html
+editor: visual
 ---
-
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
-Adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 
-**Authors**
-
--   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
--   Second Author$^{3, *}$
--   Third Autor$^{3, \dagger}$
--   Last Author$^{1, 2, \land}$
-
-**Author affiliations**
-
-1.  College of Public Health, University of Georgia, Athens, GA, USA.
-2.  Another Affiliation, Athens, GA, USA.
-3.  Yet another Affiliation.
-
-$*$ These authors contributed equally to this work.
-
-$\land$ Corresponding author: some\@email.com
-
-$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.
-
-{{< pagebreak >}}
-
-# Summary/Abstract
-
-*Write a summary of your project.*
-
-{{< pagebreak >}}
-
-# Introduction
-
 ## General Background Information
@@
 Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease.” It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
+It’s also what most people mean when they use the general term “heart disease”.
+It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
 This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
@@
 
-Do younger individuals (e.g., age \< 40) with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
+Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
 
-refernece: 
-1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings
-1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703
-2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean
-2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel
-3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/
-3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
+refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
-
-{{< pagebreak >}}
-
-# Methods
+## Methods
 
@@
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
-
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
-
 ## Data aquisition
@@
 
-## Data import and cleaning
+## Part 1, Data import and cleaning
 
@@
 
-head(cardio_df)
+head(cardio_df)  # View the column names
 
-colnames(cardio_df)
+colnames(cardio_df)  # View the column names
 
-dim(cardio_df)
+dim(cardio_df)  # View the dimensions of the data
 
-summary(cardio_df)
+summary(cardio_df)  # View the summary of the data
 
-glimpse(cardio_df)
+glimpse(cardio_df)  # Get a glimpse of the dat
 
-gg_miss_var(cardio_df)
+gg_miss_var(cardio_df)  # Ilustrate the number of NA's for each of the variables
 
@@
 
+print(cardio_df)
 ```
 
-# Part 2. Exploratory Data Analysis (EDA):
+## Part 2. Exploratory Data Analysis (EDA):
+
 ```{r}
@@
 
-# Part 3, Statistical analysis: 
+## Part 3, Statistical analysis:
+
 ```{r}
@@
 cleaned_df <- cardio_df %>%
-  filter(age < 40, serumcholestrol > 0,serumcholestrol < 200) %>%  # Normal if serumcholestrol < 200 mg/dL
+  filter(age < 50, serumcholestrol > 0,serumcholestrol < 200) %>%  # Normal if serumcholestrol < 200 mg/dL
   mutate(
-    elevated_BP = case_when(
-      restingBP < 120 ~ "Low BP",
-      between(restingBP, 120, 129) ~ "Elevated",  # Elevated blood pressure: restingBP between 120 and 129
-      restingBP > 129 ~ "High BP"
-    ),
+    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),  # Elevated blood pressure: restingBP between 120 and 129
     elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")  # Elevated ST depression: oldpeak > 1
@@
   facet_wrap(~ elevated_oldpeak) +
-  labs(title = "CVD Risk by Blood Pressure and ST Depression Status\n(Individuals Age < 40 with Normal Cholesterol)",
+  labs(title = "CVD Risk by Blood Pressure and ST Depression Status\n(Individuals Age < 50 with Normal Cholesterol)",
        x = "Blood Pressure Status",
@@
 ```
-
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
-
-## Statistical analysis
-
-*Explain anything related to your statistical analyses.*
-
-{{< pagebreak >}}
-
-# Results
-
-## Exploratory/Descriptive analysis
-
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
-```
-
-## Basic statistical analysis
-
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
-## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
-
-{{< pagebreak >}}
-
-# Discussion
-
-## Summary and Interpretation
-
-*Summarize what you did, what you found and what it means.*
-
-## Strengths and Limitations
-
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
-## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
-
-{{< pagebreak >}}
-
-# References

### 2025-04-05 products/manuscript/Manuscript.qmd
Part 4 of the final project

--- 2025-02-04 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Yaqubi-MADA-projet"
-author: ""
-format:
-  docx:
-    toc: false
-    number-sections: true
-    highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
+title: "Final project"
+author: "Murtaza Yaqubi"
+format: html
+editor: visual
 ---
-
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
-Adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
-```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-library(here)
-library(knitr)
-library(readr)
-library(dplyr)
-library(tidyverse)
-library(ggplot2)
-library(naniar)
-```
-
-**Authors**
-
--   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
--   Second Author$^{3, *}$
--   Third Autor$^{3, \dagger}$
--   Last Author$^{1, 2, \land}$
-
-**Author affiliations**
-
-1.  College of Public Health, University of Georgia, Athens, GA, USA.
-2.  Another Affiliation, Athens, GA, USA.
-3.  Yet another Affiliation.
-
-$*$ These authors contributed equally to this work.
-
-$\land$ Corresponding author: some\@email.com
-
-$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.
-
-{{< pagebreak >}}
-
-# Summary/Abstract
-
-*Write a summary of your project.*
-
-{{< pagebreak >}}
-
-# Introduction
 
@@
 
-Do younger individuals (e.g., age \< 40) with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
+Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
+refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
 
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+## Methods
 
-{{< pagebreak >}}
+In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present.
+The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).
 
-# Methods
+We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values.
+Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression.
+Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.
 
-For this analysis I will be using target which is the outcome of interest and is described as abcense or pressence of heart disease and some of the predictors of interest are age, serum cholesterol, resting BP, and old peak.
-The goal for this analysis is to be able observe if young patients with normal cholesterol but high BP/ST depression have higher CVD rates than expected.
+To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups.
+Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar.
+Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.
 
-The analysis will first isolate younger individuals with clinically normal cholesterol levels to examine their risk of cardiovascular disease (CVD) when other factors like high blood pressure or ST depression are present.
-Patients will be categorized based on these secondary risk factors, and descriptive analysis will compare CVD prevalence among groups, such as those with normal cholesterol but elevated blood pressure versus those with entirely normal profiles.
-This step aims to reveal potential associations between non-traditional risk factors and CVD in younger populations.
+In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods.
+These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE).
+By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.
 
-To validate these findings, statistical tests like chi-square analysis will assess significance, while logistic regression models will quantify the independent and combined effects of blood pressure and ST depression on CVD risk, adjusting for confounders like gender and fasting blood sugar.
-Visualization techniques, including bar plots and boxplots, will further illustrate risk patterns, making the findings more interpretable.
-By integrating these methods, the study aims to challenge traditional risk-assessment frameworks and support more nuanced screening approaches for younger individuals at risk of CVD.
+## **Conclusion**
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
-
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
+Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present.
+Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups.
+In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance.
+Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup.
+Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease.
+Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.
 
@@
 
-## Data import and cleaning
+# Load necessary libraries
 
 ```{r}
+library(knitr)
+library(readr)
+library(dplyr)
+library(ggplot2)
+library(ggcorrplot)
+library(vcd)
+library(caret)        # For data partitioning and model training
+library(randomForest) # For Random Forest modeling
+library(tidyverse)
+library(tidymodels)
+library(broom)
+library(here)           # For file paths
+library(pROC)
+library(naniar)         # For gg_miss_var()
+```
+
+# PART 1: DATA IMPORT AND CLEANING
+
+```{r}
+# Import the dataset
 cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
 
-head(cardio_df)
+# Basic exploration of the dataset
+head(cardio_df)           # View the first few rows
+colnames(cardio_df)       # Check column names
+dim(cardio_df)            # Check dimensions (should be 1025 rows x 14 columns)
+summary(cardio_df)        # Get a summary of the data
+glimpse(cardio_df)        # Glimpse at the data structure
 
-colnames(cardio_df)
+# Visualize missing values for each variable
+gg_miss_var(cardio_df)    # Display number of NAs
 
-dim(cardio_df)
+# Transform variables into suitable types and add descriptive labels
+cardio_df <- cardio_df %>%
+  mutate(
+    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
+    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
+                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
+    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
+    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
+                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
+    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
+    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
+    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
+  )
 
-summary(cardio_df)
+# Print the transformed dataset for verification
+print(cardio_df)
+```
 
-glimpse(cardio_df)
+# PART 2: EXPLORATORY DATA ANALYSIS (EDA)
 
-gg_miss_var(cardio_df)
+```{r}
+# Calculate summary statistics for selected variables
+summary_stats <- cardio_df %>% 
+  summarise(
+    # Age statistics
+    Age_min = min(age, na.rm = TRUE),
+    Age_max = max(age, na.rm = TRUE),
+    Age_mean = mean(age, na.rm = TRUE),
+    Age_sd = sd(age, na.rm = TRUE),
 
-print(cardio_df)
+    # Resting Blood Pressure statistics
+    RestingBP_min = min(restingBP, na.rm = TRUE),
+    RestingBP_max = max(restingBP, na.rm = TRUE),
+    RestingBP_mean = mean(restingBP, na.rm = TRUE),
+    RestingBP_sd = sd(restingBP, na.rm = TRUE),
 
-view(cardio_df)
+    # Serum Cholesterol statistics (excluding zeros)
+    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
+    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
+    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
+    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
+
+    # Maximum Heart Rate statistics
+    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
+    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
+    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
+    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
+
+    # Oldpeak statistics (excluding zeros)
+    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
+    Oldpeak_max = max(oldpeak, na.rm = TRUE),
+    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
+    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
+  )
+
+# Print summary statistics
+print(summary_stats)
+
+# Create histograms and density overlays for key variables
+
+# Age distribution
+ggplot(cardio_df, aes(x = age)) +
+  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
+  geom_density(color = "darkblue", linewidth = 1) +
+  labs(title = "Age Distribution", x = "Age", y = "Density") +
+  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
+  theme_bw()
+
+# Serum Cholesterol distribution (excluding zeros)
+ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
+  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
+  geom_density(color = "darkgreen", linewidth = 1) +
+  labs(title = "Serum Cholesterol Distribution", 
+       x = "Serum Cholesterol (mg/dL)", y = "Density") +
+  theme_minimal()
+
+# Resting Blood Pressure distribution (excluding zeros)
+ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
+  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
+  geom_density(color = "darkred", linewidth = 1) +
+  labs(title = "Resting Blood Pressure Distribution", 
+       x = "Resting Blood Pressure (mmHG)", y = "Density") +
+  theme_bw()
+
+# Boxplot for resting blood pressure by heart disease status
+ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
+  geom_boxplot() +
+  labs(title = "Resting Blood Pressure by Heart Disease Status",
+       x = "Heart Disease Status",
+       y = "Resting Blood Pressure (mmHg)") +
+  theme_minimal()
+
+# Violin plot for oldpeak by heart disease status
+ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
+  geom_violin(trim = FALSE) +
+  labs(title = "Oldpeak Distribution by Heart Disease Status",
+       x = "Heart Disease Status",
+       y = "Oldpeak") +
+  theme_minimal()
+
+# Create the correlation matrix for selected numeric variables
+numeric_vars <- cardio_df %>% 
+  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
+cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
+
+# Plot the correlation matrix with ggcorrplot
+ggcorrplot(cor_matrix, 
+           lab = TRUE,                         
+           lab_size = 3,                       
+           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
+           title = "Correlation Matrix", 
+           ggtheme = theme_minimal())       
+```
+
+# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
+
+```{r}
+# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
+cleaned_df <- cardio_df %>%
+  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
+  mutate(
+    # Create a binary variable for blood pressure:
+    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
+    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),
+
+    # Create a binary indicator for ST depression:
+    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
+    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
+  ) %>% 
+  # Select the variables we want to keep
+  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
+         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)
+
+# Print a sample of the cleaned dataset
+head(cleaned_df)
+
+# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
+summary_table <- cleaned_df %>%
+  group_by(elevated_BP, elevated_oldpeak, target) %>%
+  summarise(count = n(), .groups = "drop")
+print(summary_table)
+
+# Interaction Visualization: Heatmap
+interaction_counts <- cleaned_df %>%
+  group_by(elevated_BP, elevated_oldpeak, target) %>%
+  summarise(count = n(), .groups = "drop")
+
+heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
+  geom_tile(color = "white", size = 0.5) +
+  geom_text(aes(label = count), color = "black", size = 5) +
+  facet_wrap(~ target, ncol = 2) +
+  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
+  labs(title = "Interaction between Blood Pressure and ST Depression",
+       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
+       x = "Blood Pressure Status",
+       y = "ST Depression Status",
+       fill = "Count") +
+  theme_minimal(base_size = 14) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    strip.text = element_text(face = "bold"),
+    plot.title = element_text(face = "bold", hjust = 0.5),
+    plot.subtitle = element_text(hjust = 0.5)
+  )
+print(heatmap_plot)
+
+# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
+ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
+  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
+  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
+       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
+       x = "Blood Pressure Status",
+       y = "Count",
+       fill = "Heart Disease") +
+  theme_minimal(base_size = 14) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    strip.text = element_text(face = "bold"),
+    plot.title = element_text(face = "bold", hjust = 0.5),
+    plot.subtitle = element_text(hjust = 0.5),
+    legend.position = "bottom"
+  )
+```
+
+# PART 4a: INFERENTIAL STATISTICAL ANALYSIS
+
+# Chi-Square Tests
+
+```{r}
+# Create a contingency table for blood pressure status vs. heart disease
+table_bp_target <- table(cleaned_df$elevated_BP, cleaned_df$target)
+print(table_bp_target)
+chisq_test_bp <- chisq.test(table_bp_target)
+print(chisq_test_bp)
+
+# Create a contingency table for ST depression vs. heart disease
+table_oldpeak_target <- table(cleaned_df$elevated_oldpeak, cleaned_df$target)
+print(table_oldpeak_target)
+chisq_test_oldpeak <- chisq.test(table_oldpeak_target)
+print(chisq_test_oldpeak)
+```
+
+The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup.
+For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels.
+Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.
+
+# Logistic Regression Analysis
+
+```{r}
+# Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
+cleaned_df <- cleaned_df %>%
+  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
+
+# Fit a logistic regression model with elevated_BP and elevated_oldpeak as predictors
+model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
+             data = cleaned_df, family = binomial)
+summary(model)
+
+# Fit a logistic regression model including the interaction term
+model_interaction <- glm(target_binary ~ elevated_BP * elevated_oldpeak, 
+                         data = cleaned_df, family = binomial)
+summary(model_interaction)
+```
+
+In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero.
+However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own.
+When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance.
+These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.
+
+# Extract Odds Ratios (OR) from the logistic regression model with interaction
+
+```{r}
+or_values <- exp(coef(model_interaction))
+or_ci <- exp(confint(model_interaction))
+or_results <- data.frame(
+  Term = names(or_values),
+  OR = or_values,
+  CI_lower = or_ci[, 1],
+  CI_upper = or_ci[, 2]
+)
+print("Odds Ratios (with 95% CI):")
+print(or_results)
+
+# Tidy the model output and compute odds ratios and 95% CI
+tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
+print(tidy_model)
+```
+
+The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87).
+For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22).
+The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82).
+However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.
+
+# Fit a log-binomial model to estimate Relative Risks (RR)
+
+```{r}
+rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
+                data = cleaned_df, family = binomial(link = "log"))
+rr_values <- exp(coef(rr_model))
+rr_ci <- exp(confint(rr_model))
+rr_results <- data.frame(
+  Term = names(rr_values),
+  RR = rr_values,
+  CI_lower = rr_ci[, 1],
+  CI_upper = rr_ci[, 2]
+)
+print("Relative Risks (with 95% CI):")
+print(rr_results)
+```
+
+The baseline relative risk is about 0.23 (95% CI: 0.08–0.46).
+For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35).
+The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.
+
+# ROC Curve and AUC for the logistic regression model with interaction
+
+```{r}
+roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
+plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
+auc_value <- auc(roc_curve)
+print(paste("AUC:", auc_value))
+```
+
+The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing.
+It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.
+
+# Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)
+
+```{r}
+adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
+                      data = cleaned_df, family = binomial)
+summary(adjusted_model)
+
+# Extract Odds Ratios (OR) for the adjusted model
+adj_or_values <- exp(coef(adjusted_model))
+adj_or_ci <- exp(confint(adjusted_model))
+adj_or_results <- data.frame(
+  Term = names(adj_or_values),
+  OR = adj_or_values,
+  CI_lower = adj_or_ci[, 1],
+  CI_upper = adj_or_ci[, 2]
+)
+print("Adjusted Odds Ratios (with 95% CI):")
+print(adj_or_results)
+
+# Tidy the adjusted model output and compute odds ratios and 95% CI
+tidy_adjusted_model <- tidy(adjusted_model, exponentiate = TRUE, conf.int = TRUE)
+print(tidy_adjusted_model)
+```
+
+The adjusted logistic regression model produces unstable estimates.
+The intercept's odds ratio is about 0.07, indicating very low baseline odds.
+The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant.
+The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals.
+These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable.
+This interpretation is derived solely from your output.
+
+# PART 4b: MODEL BUILDING
+
+```{r}
+# Set seed for reproducibility
+rdmseed <- 123
+
+# Reset seed before modeling
+set.seed(rdmseed)
+
+cleaned_df <- cleaned_df %>%
+  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
+```
+
+# linear model
+
+```{r}
+# Fit a standard linear regression model using predictors
+lm_model <- linear_reg() %>%
+  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
+
+# Get predictions from linear model and calculate RMSE
+cleaned_df$pred_lm <- predict(lm_model, new_data = cleaned_df)$.pred
+rmse(cleaned_df, truth = target_binary, estimate = pred_lm)
+
+# # Plot observed vs. predicted for Linear Model
+ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
+  geom_point(color = "darkblue", alpha = 0.7, size = 3) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
+  scale_x_continuous(limits = c(0, 1)) +
+  scale_y_continuous(limits = c(0, 1))
 
@@
 
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
+The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average.
+This indicates limited accuracy for distinguishing between individuals with and without heart disease.
 
-## Statistical analysis
-
-*Explain anything related to your statistical analyses.*
-
-{{< pagebreak >}}
-
-# Results
-
-## Exploratory/Descriptive analysis
-
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
+# LASSO
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
+set.seed(rdmseed)
+
+# Prepare recipe for LASSO with normalization
+rec <- recipe(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df) %>%
+  step_dummy(elevated_BP, elevated_oldpeak) %>%   # Convert categorical predictors to dummies
+  step_normalize(all_numeric(), -all_outcomes())
+
+# Define LASSO model with a fixed penalty
+lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
+  set_engine("glmnet")
+
+# Create a workflow and fit the LASSO model
+wf <- workflow() %>%
+  add_recipe(rec) %>%
+  add_model(lasso_spec)
+
+lasso_fit <- fit(wf, data = cleaned_df)
+cleaned_df$pred_lasso <- predict(lasso_fit, new_data = cleaned_df)$.pred
+rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)
+
+# Plot observed vs. predicted for LASSO Model
+ggplot(cleaned_df, aes(x = target_binary, y = pred_lasso)) +
+  geom_point(color = "darkviolet", alpha = 0.7, size = 3) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
+  scale_x_continuous(limits = c(0, 1)) +
+  scale_y_continuous(limits = c(0, 1))
 ```
 
-## Basic statistical analysis
+The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average.
+This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+# Random Forest
 
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
+set.seed(rdmseed)
+
+# Create a workflow and fit the Random Forest model
+rf_model <- rand_forest() %>%
+  set_mode("regression") %>%
+  set_engine("ranger", seed = rdmseed)
+
+rf_fit <- rf_model %>%
+  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
+
+# Get predictions and evaluate model performance
+cleaned_df$pred_rf <- predict(rf_fit, cleaned_df)$.pred
+rmse(cleaned_df, truth = target_binary, estimate = pred_rf)
+
+# Plot observed vs. predicted for RF Model
+ggplot(cleaned_df, aes(x = target_binary, y = pred_rf)) +
+  geom_point(color = "darkgreen", alpha = 0.7, size = 3) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
+  scale_x_continuous(limits = c(0, 1)) +
+  scale_y_continuous(limits = c(0, 1))
+
 ```
 
-## Full analysis
+The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half.
+This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.
 
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
+# Tuning the LASSO model (without cross-validation)
 
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
+set.seed(rdmseed)
+
+# Create a grid of penalty values for LASSO
+lasso_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))
+
+# Define LASSO model with tunable penalty
+lasso_spec_tune <- linear_reg(penalty = tune()) %>%
+  set_engine("glmnet")
+
+# Create workflow and assign to lasso_tuning_workflow (new name)
+lasso_tuning_workflow <- workflow() %>%
+  add_model(lasso_spec_tune) %>%
+  add_recipe(rec)
+
+# Perform grid search without CV (apparent resampling)
+lasso_tune <- tune_grid(
+  lasso_tuning_workflow,
+  resamples = apparent(cleaned_df),
+  grid = lasso_grid,
+  metrics = metric_set(rmse)
+)
+
+# Extract tuning results
+lasso_tune_df <- as.data.frame(lasso_tune$.metrics)
+
+# Plot LASSO tuning results
+ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
+  geom_line(size = 1, color = "steelblue") +
+  geom_point(size = 2, color = "firebrick") +
+  scale_x_log10() +
+  labs(
+    x = "Penalty (log scale)",
+    y = "RMSE",
+    title = "LASSO Tuning (No Cross-validation)",
+    subtitle = "Using apparent resampling"
+  ) +
+  theme_minimal()
 ```
 
-{{< pagebreak >}}
+The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties.
+This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.
 
-# Discussion
+# Tuning for our Random Forest model (without cross-validation)
 
-## Summary and Interpretation
+```{r}
+# Create tuning grid for mtry and min_n
+rf_grid <- grid_regular(
+  mtry(range = c(1, 7)),
+  min_n(range = c(1, 21)),
+  levels = 7
+)
 
-*Summarize what you did, what you found and what it means.*
+# Define random forest model with tunable parameters
+rf_spec_tune <- rand_forest(
+  mtry = tune(),
+  min_n = tune(),
+  trees = 300
+) %>%
+  set_mode("regression") %>%
+  set_engine("ranger", seed = rdmseed)
 
-## Strengths and Limitations
+# Build workflow
+rf_wf_tune <- workflow() %>%
+  add_model(rf_spec_tune) %>%
+  add_recipe(rec)
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
+# Perform RF tuning without CV
+rf_tune <- tune_grid(
+  rf_wf_tune,
+  resamples = apparent(cleaned_df),
+  grid = rf_grid,
+  metrics = metric_set(rmse)
+)
 
-## Conclusions
+rf_tune_df <- as.data.frame(rf_tune$.metrics)
 
-*What are the main take-home messages?*
+# Plot RF tuning results
+ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
+  geom_tile() +
+  scale_fill_viridis_c() +
+  labs(
+    title = "Random Forest Tuning (No Cross-validation)",
+    x = "mtry",
+    y = "min_n",
+    fill = "RMSE"
+  ) +
+  theme_minimal()
+```
 
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
+The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739.
+This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used.
+In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.
 
-This paper [@leek2015] discusses types of analyses.
+# Cross-validation for our LASSO model tuning
 
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
+```{r}
+set.seed(rdmseed)
 
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+# Create 5-fold CV with 5 repeats
+cv_folds <- vfold_cv(cleaned_df, v = 5, repeats = 5)
 
-{{< pagebreak >}}
+# Tune LASSO using CV
+lasso_tune_cv <- tune_grid(
+  lasso_tuning_workflow,
+  resamples = cv_folds,
+  grid = lasso_grid,
+  metrics = metric_set(rmse)
+)
 
-# References
+autoplot(lasso_tune_cv)
+```
+
+Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases.
+This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.
+
+# Cross-validation tuning for our random forest model
+
+```{r}
+# Tune RF using CV
+rf_tune_cv <- tune_grid(
+  rf_wf_tune,
+  resamples = cv_folds,
+  grid = rf_grid,
+  metrics = metric_set(rmse)
+)
+autoplot(rf_tune_cv)
+
+```
+
+Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry).
+In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values.
+This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.

### 2025-04-05 products/manuscript/supplement/Supplementary-Material.qmd
Part 4 of the final project

---
title: "Supplement to Example Manuscript Template for a Data Analysis Project"
format:
  pdf:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

This shows some materials that could go into a supplementary file. Often you want/need references here too. You can use the same reference bib file for this and the main text (as done here) or have separate bib files.

For illustrative purposes, I'm doing the supplement as pdf. For this to work, you need a (La)TeX system installed. It's easy. Just follow [these steps](https://quarto.org/docs/output-formats/pdf-basics.html).

Of course you would choose the format based on needs.

I'm also using a different style for the references here. (vancouver vs apa in the main manuscript). Usually one would have the formatting of the references the same in those two documents, but I want to illustrate how easy it is to switch reference formatting styles, you just need to get the right CSL file and specify it in the YAML header. We could also have a seperate reference bibtext (`.bib`) file, but here we are using the same.

{{< pagebreak >}}

# Overview

A quick overview of what readers can find in the supplement.

# Code and file information

Explain here what each code/file is and does, and in which order (if any) users need to run thing to reproduce everything. Essentially, give a full set of instructions to re-generate everything.

{{< pagebreak >}}

# Additional Method Details

Often, the main manuscript only allows for an overview description of the methods. Use the supplement to describe all your methods, models and approaches in a lot of detail. Reference specific parts of your code as needed.

{{< pagebreak >}}

# Additional results

Show additional results here. Those can be some useful exploratory/descriptive figures or tables, or results from additional analyses that didn't make it into the main text.

## Example additional result

@tbl-resulttable1 shows an additional table summarizing a model fit.

```{r}
#| label: tbl-resulttable1
#| tbl-cap: "Another fit table."
#| echo: FALSE
resulttable1 = readRDS(here("results","tables","resulttable1.rds"))
knitr::kable(resulttable1)
```

@fig-result2 shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result2
#| fig-cap: "Height and weight."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight.png"))
```

{{< pagebreak >}}

# Discussion

Any additional discussion regarding the supplementary material/findings.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

{{< pagebreak >}}

# References

### 2025-04-05 code/eda-code/eda1.qmd
Changes

---
title: "Exploratory Data Analysis"
author: "Murtaza Yaqubi"
format: html
editor: visual
---

# Library/install libraries
```{r}
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(corrplot)
```

# EXPLORATORY DATA ANALYSIS (EDA)
```{r}
# Import the dataset
cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))

# Transform variables into suitable types and add descriptive labels
cardio_df <- cardio_df %>%
  mutate(
    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
  )

# Print the transformed dataset for verification
print(cardio_df)
```

# Calculate summary statistics for selected variables
```{r}
# Calculate summary statistics for selected variables
summary_stats <- cardio_df %>% 
  summarise(
    # Age statistics
    Age_min = min(age, na.rm = TRUE),
    Age_max = max(age, na.rm = TRUE),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd = sd(age, na.rm = TRUE),

    # Resting Blood Pressure statistics
    RestingBP_min = min(restingBP, na.rm = TRUE),
    RestingBP_max = max(restingBP, na.rm = TRUE),
    RestingBP_mean = mean(restingBP, na.rm = TRUE),
    RestingBP_sd = sd(restingBP, na.rm = TRUE),

    # Serum Cholesterol statistics (excluding zeros)
    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),

    # Maximum Heart Rate statistics
    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),

    # Oldpeak statistics (excluding zeros)
    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
    Oldpeak_max = max(oldpeak, na.rm = TRUE),
    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
  )

# Print summary statistics
print(summary_stats)

summary_df = skimr::skim(summary_stats)
print(summary_stats)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

# Age distribution
```{r}
p1 <- ggplot(cardio_df, aes(x = age)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(title = "Age Distribution", x = "Age", y = "Density") +
  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
  theme_bw()

plot(p1)
figure_file = here("results", "figures","Age distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

# Serum Cholesterol distribution
```{r}
p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
  geom_density(color = "darkgreen", linewidth = 1) +
  labs(title = "Serum Cholesterol Distribution", 
       x = "Serum Cholesterol (mg/dL)", y = "Density") +
  theme_minimal()

plot(p2)
figure_file = here("results", "figures","Serum Cholestrol.png")
ggsave(filename = figure_file, plot=p2) 
```

# Resting Blood Pressure distribution
```{r}
p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Resting Blood Pressure Distribution", 
       x = "Resting Blood Pressure (mmHG)", y = "Density") +
  theme_bw()

plot(p3)
figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
ggsave(filename = figure_file, plot=p3) 
```

# Boxplot for resting blood pressure by heart disease status
```{r}
p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
  geom_boxplot() +
  labs(title = "Resting Blood Pressure by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Resting Blood Pressure (mmHg)") +
  theme_minimal()

plot(p4)
figure_file = here("results", "figures","restingBP by heart disease status.png")
ggsave(filename = figure_file, plot=p4) 
```

# Violin plot for oldpeak by heart disease status
```{r}
p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
  geom_violin(trim = FALSE) +
  labs(title = "Oldpeak Distribution by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Oldpeak") +
  theme_minimal()

plot(p5)
figure_file = here("results", "figures","oldpeak by heart disease status.png")
ggsave(filename = figure_file, plot=p5)
```

# Create the correlation matrix for selected numeric variables
```{r}
numeric_vars <- cardio_df %>% 
  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Plot the correlation matrix with ggcorrplot
p6 <- ggcorrplot(cor_matrix, 
           lab = TRUE,                         
           lab_size = 3,                       
           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
           title = "Correlation Matrix", 
           ggtheme = theme_minimal())

plot(p6)
figure_file = here("results", "figures","correlation matrix.png")
ggsave(filename = figure_file, plot=p6) 
```

### 2025-04-05 code/processing-code/processed file.qmd
Changes

--- 2025-02-04 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Yaqubi-MADA-projet"
-author: ""
-format:
-  docx:
-    toc: false
-    number-sections: true
-    highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
+title: "Processing code"
+author: "Murtaza Yaqubi"
+format: html
+editor: visual
 ---
 
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
-Adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
-```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-library(here)
-library(knitr)
+# Load necessary libraries
+```{r}
 library(readr)
 library(dplyr)
+library(ggplot2)
 library(tidyverse)
-library(ggplot2)
-library(naniar)
+library(here)           # For file paths
+library(naniar)         # For gg_miss_var()
 ```
 
-**Authors**
+# PART 1: DATA IMPORT AND CLEANING
+```{r}
+# Import the dataset
+cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
 
--   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
--   Second Author$^{3, *}$
--   Third Autor$^{3, \dagger}$
--   Last Author$^{1, 2, \land}$
+# Basic exploration of the dataset
+head(cardio_df)           # View the first few rows
+colnames(cardio_df)       # Check column names
+dim(cardio_df)            # Check dimensions (should be 1025 rows x 14 columns)
+summary(cardio_df)        # Get a summary of the data
+glimpse(cardio_df)        # Glimpse at the data structure
 
-**Author affiliations**
+# Visualize missing values for each variable
+gg_miss_var(cardio_df)    # Display number of NAs
 
-1.  College of Public Health, University of Georgia, Athens, GA, USA.
-2.  Another Affiliation, Athens, GA, USA.
-3.  Yet another Affiliation.
+# Transform variables into suitable types and add descriptive labels
+cardio_df <- cardio_df %>%
+  mutate(
+    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
+    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
+                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
+    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
+    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
+                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
+    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
+    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
+    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
+  )
 
-$*$ These authors contributed equally to this work.
-
-$\land$ Corresponding author: some\@email.com
-
-$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.
-
-{{< pagebreak >}}
-
-# Summary/Abstract
-
-*Write a summary of your project.*
-
-{{< pagebreak >}}
-
-# Introduction
-
-## General Background Information
-
-Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease.” It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
-This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
-As the leading cause of death in the U.S.
-and globally, CAD claimed 375,500 lives in the U.S.
-in 2021 and affects over 18 million American adults.
-
-## Description of data and data source
-
-This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses.
-Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India.
-This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-This dataset was obtained from mendeley data where it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-## Questions/Hypotheses to be addressed
-
-Do younger individuals (e.g., age \< 40) with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
-
-{{< pagebreak >}}
-
-# Methods
-
-For this analysis I will be using target which is the outcome of interest and is described as abcense or pressence of heart disease and some of the predictors of interest are age, serum cholesterol, resting BP, and old peak.
-The goal for this analysis is to be able observe if young patients with normal cholesterol but high BP/ST depression have higher CVD rates than expected.
-
-The analysis will first isolate younger individuals with clinically normal cholesterol levels to examine their risk of cardiovascular disease (CVD) when other factors like high blood pressure or ST depression are present.
-Patients will be categorized based on these secondary risk factors, and descriptive analysis will compare CVD prevalence among groups, such as those with normal cholesterol but elevated blood pressure versus those with entirely normal profiles.
-This step aims to reveal potential associations between non-traditional risk factors and CVD in younger populations.
-
-To validate these findings, statistical tests like chi-square analysis will assess significance, while logistic regression models will quantify the independent and combined effects of blood pressure and ST depression on CVD risk, adjusting for confounders like gender and fasting blood sugar.
-Visualization techniques, including bar plots and boxplots, will further illustrate risk patterns, making the findings more interpretable.
-By integrating these methods, the study aims to challenge traditional risk-assessment frameworks and support more nuanced screening approaches for younger individuals at risk of CVD.
-
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
-
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
+# Print the transformed dataset for verification
+print(cardio_df)
 ```
 
-## Data aquisition
+# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
+```{r}
+# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
+cleaned_df <- cardio_df %>%
+  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
+  mutate(
+    # Create a binary variable for blood pressure:
+    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
+    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),
 
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.
+    # Create a binary indicator for ST depression:
+    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
+    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
+  ) %>% 
+  # Select the variables we want to keep
+  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
+         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)
 
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
+# Print a sample of the cleaned dataset
+head(cleaned_df)
 
-## Data import and cleaning
+# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
+summary_table <- cleaned_df %>%
+  group_by(elevated_BP, elevated_oldpeak, target) %>%
+  summarise(count = n(), .groups = "drop")
+print(summary_table)
 
-```{r}
-cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
-
-head(cardio_df)
-
-colnames(cardio_df)
-
-dim(cardio_df)
-
-summary(cardio_df)
-
-glimpse(cardio_df)
-
-gg_miss_var(cardio_df)
-
-print(cardio_df)
-
-view(cardio_df)
-
+summary_df = skimr::skim(summary_table)
+print(summary_table)
+# save to file
+summarytable_file = here("results", "tables", "summarytable2.rds")
+saveRDS(summary_df, file = summarytable_file)
 ```
 
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
-
-## Statistical analysis
-
-*Explain anything related to your statistical analyses.*
-
-{{< pagebreak >}}
-
-# Results
-
-## Exploratory/Descriptive analysis
-
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
-
+# Save the processed data
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
+save_data_location <- here::here("data/processed-data/cleaned_df.rds")
+saveRDS(cleaned_df, file = save_data_location)
 ```
-
-## Basic statistical analysis
-
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
-## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
-
-{{< pagebreak >}}
-
-# Discussion
-
-## Summary and Interpretation
-
-*Summarize what you did, what you found and what it means.*
-
-## Strengths and Limitations
-
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
-## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
-
-{{< pagebreak >}}
-
-# References

### 2025-04-05 products/manuscript/Manuscript.qmd
Changes

@@
 # Load necessary libraries
-
 ```{r}
@@
 # PART 1: DATA IMPORT AND CLEANING
-
 ```{r}
@@
 
-# Create histograms and density overlays for key variables
+summary_df = skimr::skim(summary_stats)
+print(summary_stats)
+# save to file
+summarytable_file = here("results", "tables", "summarytable.rds")
+saveRDS(summary_df, file = summarytable_file)
 
 # Age distribution
-ggplot(cardio_df, aes(x = age)) +
+p1 <- ggplot(cardio_df, aes(x = age)) +
   geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
@@
 
+plot(p1)
+figure_file = here("results", "figures","Age distribution.png")
+ggsave(filename = figure_file, plot=p1)
+
 # Serum Cholesterol distribution (excluding zeros)
-ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
+p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
   geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
@@
 
+plot(p2)
+figure_file = here("results", "figures","Serum Cholestrol.png")
+ggsave(filename = figure_file, plot=p2)
+
 # Resting Blood Pressure distribution (excluding zeros)
-ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
+p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
   geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
@@
 
+plot(p3)
+figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
+ggsave(filename = figure_file, plot=p3)
+
 # Boxplot for resting blood pressure by heart disease status
-ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
+p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
   geom_boxplot() +
@@
 
+plot(p4)
+figure_file = here("results", "figures","restingBP by heart disease status.png")
+ggsave(filename = figure_file, plot=p4)
+
 # Violin plot for oldpeak by heart disease status
-ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
+p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
   geom_violin(trim = FALSE) +
@@
   theme_minimal()
+
+plot(p5)
+figure_file = here("results", "figures","oldpeak by heart disease status.png")
+ggsave(filename = figure_file, plot=p5)
 
@@
 # Plot the correlation matrix with ggcorrplot
-ggcorrplot(cor_matrix, 
+p6 <- ggcorrplot(cor_matrix, 
            lab = TRUE,                         
@@
            ggtheme = theme_minimal())
+
+plot(p6)
+figure_file = here("results", "figures","correlation matrix.png")
+ggsave(filename = figure_file, plot=p6)     
 ```
@@
 
+# Save the processed data
+saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")
+
 # Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
@@
 
+summary_df = skimr::skim(summary_table)
+print(summary_table)
+# save to file
+summarytable_file = here("results", "tables", "summarytable2.rds")
+saveRDS(summary_df, file = summarytable_file)
+
 # Interaction Visualization: Heatmap
@@
 
-heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
+p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
   geom_tile(color = "white", size = 0.5) +
@@
   )
-print(heatmap_plot)
+
+plot(p7)
+figure_file = here("results", "figures","interaction heatmap.png")
+ggsave(filename = figure_file, plot=p7)
 
 # Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
-ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
+p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
   geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
@@
   )
+
+plot(p8)
+figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
+ggsave(filename = figure_file, plot=p8)
 ```

### 2025-04-05 code/analysis-code/statistical-analysis.R

@@ -1,56 +1,187 @@
-###############################
-# analysis script
-#
-#this script loads the processed, cleaned data, does a simple analysis
-#and saves the results to the results folder
-
-#load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
-library(here) #for data loading/saving
-
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processeddata.rds")
-
-#load data. 
-mydata <- readRDS(data_location)
-
-
-######################################
-#Data fitting/statistical analysis
-######################################
-
-############################
-#### First model fit
-# fit linear model using height as outcome, weight as predictor
-
-lmfit1 <- lm(Height ~ Weight, mydata)  
-
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
-
-#look at fit results
-print(lmtable1)
-
-# save fit results table  
-table_file1 = here("results", "tables", "resulttable1.rds")
-saveRDS(lmtable1, file = table_file1)
-
-############################
-#### Second model fit
-# fit linear model using height as outcome, weight and gender as predictor
-
-lmfit2 <- lm(Height ~ Weight + Gender, mydata)  
-
-# place results from fit into a data frame with the tidy function
-lmtable2 <- broom::tidy(lmfit2)
-
-#look at fit results
-print(lmtable2)
-
-# save fit results table  
-table_file2 = here("results", "tables", "resulttable2.rds")
-saveRDS(lmtable2, file = table_file2)
-
+---
+  title: "Statistical analysis"
+ author: "Murtaza Yaqubi"
+---
   
+# Load necessary packages
+
+library(here)
+library(dplyr)
+library(ggplot2)
+
+
+# Load data
+data_location <- here::here("data","processed-data","cleaned_df.rds")
+
+cleaned_df <- readRDS(data_location)
+
+
+
+# STATISTICAL ANALYSIS
+
+# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
+summary_table <- cleaned_df %>%
[…]

### 2025-04-06 products/manuscript/Manuscript.qmd
Finalizing changes

@@
 # Load necessary libraries
+
 ```{r}
@@
 # PART 1: DATA IMPORT AND CLEANING
+
 ```{r}
@@
 
+# save fit results table  
+table_file1 = here("results", "tables", "resulttable1.rds")
+saveRDS(model, file = table_file1)
+
 # Fit a logistic regression model including the interaction term
@@
 summary(model_interaction)
+
+# save fit results table  
+table_file1 = here("results", "tables", "resulttable2.rds")
+saveRDS(model_interaction, file = table_file1)
 ```
@@
 print(or_results)
+
+# save fit results table  
+table_file1 = here("results", "tables", "resulttable3.rds")
+saveRDS(or_results, file = table_file1)
 
@@
 print(rr_results)
+
+# save fit results table  
+table_file1 = here("results", "tables", "resulttable4.rds")
+saveRDS(rr_results, file = table_file1)
 ```
@@
 print(adj_or_results)
+
+# save fit results table  
+table_file1 = here("results", "tables", "resulttable5.rds")
+saveRDS(adj_or_results, file = table_file1)
 

### 2025-04-06 code/eda-code/eda.qmd
Made a few changes to the readme files

--- 2025-04-05 code/eda-code/eda1.qmd
@@
 # Library/install libraries
+
 ```{r}
@@
 # EXPLORATORY DATA ANALYSIS (EDA)
+
 ```{r}
@@
 # Calculate summary statistics for selected variables
+
 ```{r}
@@
 # save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
+summarytable_file = here("results", "tables", "summarytable1.rds")
 saveRDS(summary_df, file = summarytable_file)
@@
 # Age distribution
+
 ```{r}
@@
 # Serum Cholesterol distribution
+
 ```{r}
@@
 # Resting Blood Pressure distribution
+
 ```{r}
@@
 # Boxplot for resting blood pressure by heart disease status
+
 ```{r}
@@
 # Violin plot for oldpeak by heart disease status
+
 ```{r}
@@
 # Create the correlation matrix for selected numeric variables
+
 ```{r}

### 2025-04-06 code/processing-code/processed file.qmd
Made a few changes to the readme files

@@
 
-# PART 1: DATA IMPORT AND CLEANING
+# DATA IMPORT AND CLEANING
+
 ```{r}
@@
 
-# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
+# STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
+
 ```{r}
@@
 
-# Save the processed data
+# SAVE THE PROCESSED DATA
+
 ```{r}

### 2025-04-17 code/analysis-code/01_inferential_stats.qmd
Few minor changes

--- 2025-04-06 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Final project"
+title: "Inferential Statistics"
 author: "Murtaza Yaqubi"
@@
 ---
-
-## General Background Information
-
-Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease”.
-It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
-This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
-As the leading cause of death in the U.S.
-and globally, CAD claimed 375,500 lives in the U.S.
-in 2021 and affects over 18 million American adults.
-
-## Description of data and data source
-
-This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses.
-Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India.
-This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-This dataset was obtained from mendeley data where it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-## Questions/Hypotheses to be addressed
-
-Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
-
-refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
-
-## Methods
-
-In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present.
-The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).
-
-We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values.
-Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression.
-Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.
-
-To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups.
-Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar.
-Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.
-
-In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods.
-These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE).
-By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.
-
-## **Conclusion**
-
-Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present.
-Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups.
-In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance.
-Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup.
-Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease.
-Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.
-
-## Data aquisition
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-# Load necessary libraries
-
-```{r}
-library(knitr)
-library(readr)
-library(dplyr)
-library(ggplot2)
-library(ggcorrplot)
-library(vcd)
-library(caret)        # For data partitioning and model training
-library(randomForest) # For Random Forest modeling
-library(tidyverse)
-library(tidymodels)
-library(broom)
-library(here)           # For file paths
-library(pROC)
-library(naniar)         # For gg_miss_var()
-```
-
-# PART 1: DATA IMPORT AND CLEANING
-
-```{r}
-# Import the dataset
-cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
-
-# Basic exploration of the dataset
-head(cardio_df)           # View the first few rows
-colnames(cardio_df)       # Check column names
-dim(cardio_df)            # Check dimensions (should be 1025 rows x 14 columns)
-summary(cardio_df)        # Get a summary of the data
-glimpse(cardio_df)        # Glimpse at the data structure
-
-# Visualize missing values for each variable
-gg_miss_var(cardio_df)    # Display number of NAs
-
-# Transform variables into suitable types and add descriptive labels
-cardio_df <- cardio_df %>%
-  mutate(
-    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
-    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
-                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
-    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
-    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
-                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
-    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
-    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
-    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
-  )
-
-# Print the transformed dataset for verification
-print(cardio_df)
-```
-
-# PART 2: EXPLORATORY DATA ANALYSIS (EDA)
-
-```{r}
-# Calculate summary statistics for selected variables
-summary_stats <- cardio_df %>% 
-  summarise(
-    # Age statistics
-    Age_min = min(age, na.rm = TRUE),
-    Age_max = max(age, na.rm = TRUE),
-    Age_mean = mean(age, na.rm = TRUE),
-    Age_sd = sd(age, na.rm = TRUE),
-
-    # Resting Blood Pressure statistics
-    RestingBP_min = min(restingBP, na.rm = TRUE),
-    RestingBP_max = max(restingBP, na.rm = TRUE),
-    RestingBP_mean = mean(restingBP, na.rm = TRUE),
-    RestingBP_sd = sd(restingBP, na.rm = TRUE),
-
-    # Serum Cholesterol statistics (excluding zeros)
-    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
-
-    # Maximum Heart Rate statistics
-    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
-
-    # Oldpeak statistics (excluding zeros)
-    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
-    Oldpeak_max = max(oldpeak, na.rm = TRUE),
-    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
-    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
-  )
-
-# Print summary statistics
-print(summary_stats)
-
-summary_df = skimr::skim(summary_stats)
-print(summary_stats)
-# save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Age distribution
-p1 <- ggplot(cardio_df, aes(x = age)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
-  geom_density(color = "darkblue", linewidth = 1) +
-  labs(title = "Age Distribution", x = "Age", y = "Density") +
-  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
-  theme_bw()
-
-plot(p1)
-figure_file = here("results", "figures","Age distribution.png")
-ggsave(filename = figure_file, plot=p1)
-
-# Serum Cholesterol distribution (excluding zeros)
-p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
-  geom_density(color = "darkgreen", linewidth = 1) +
-  labs(title = "Serum Cholesterol Distribution", 
-       x = "Serum Cholesterol (mg/dL)", y = "Density") +
-  theme_minimal()
-
-plot(p2)
-figure_file = here("results", "figures","Serum Cholestrol.png")
-ggsave(filename = figure_file, plot=p2)
-
-# Resting Blood Pressure distribution (excluding zeros)
-p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
-  geom_density(color = "darkred", linewidth = 1) +
-  labs(title = "Resting Blood Pressure Distribution", 
-       x = "Resting Blood Pressure (mmHG)", y = "Density") +
-  theme_bw()
-
-plot(p3)
-figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
-ggsave(filename = figure_file, plot=p3)
-
-# Boxplot for resting blood pressure by heart disease status
-p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
-  geom_boxplot() +
-  labs(title = "Resting Blood Pressure by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Resting Blood Pressure (mmHg)") +
-  theme_minimal()
-
-plot(p4)
-figure_file = here("results", "figures","restingBP by heart disease status.png")
-ggsave(filename = figure_file, plot=p4)
-
-# Violin plot for oldpeak by heart disease status
-p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
-  geom_violin(trim = FALSE) +
-  labs(title = "Oldpeak Distribution by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Oldpeak") +
-  theme_minimal()
-
-plot(p5)
-figure_file = here("results", "figures","oldpeak by heart disease status.png")
-ggsave(filename = figure_file, plot=p5)
-
-# Create the correlation matrix for selected numeric variables
-numeric_vars <- cardio_df %>% 
-  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
-cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
-
-# Plot the correlation matrix with ggcorrplot
-p6 <- ggcorrplot(cor_matrix, 
-           lab = TRUE,                         
-           lab_size = 3,                       
-           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
-           title = "Correlation Matrix", 
-           ggtheme = theme_minimal())
-
-plot(p6)
-figure_file = here("results", "figures","correlation matrix.png")
-ggsave(filename = figure_file, plot=p6)     
-```
-
-# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
 
@@
 # Save the processed data
-saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")
+saveRDS(cleaned_df, file = here("data", "processed-data", "cleaned_df.rds"))
 
@@
 summary_df = skimr::skim(summary_table)
-print(summary_table)
+print(summary_df)
+
 # save to file
@@
 ggsave(filename = figure_file, plot=p8)
-```
-
-# PART 4a: INFERENTIAL STATISTICAL ANALYSIS
 
@@
 
-```{r}
 # Create a contingency table for blood pressure status vs. heart disease
@@
 print(chisq_test_oldpeak)
-```
-
-The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup.
-For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels.
-Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.
 
@@
 
-```{r}
 # Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
@@
 saveRDS(model_interaction, file = table_file1)
-```
-
-In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero.
-However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own.
-When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance.
-These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.
 
 # Extract Odds Ratios (OR) from the logistic regression model with interaction
-
-```{r}
 or_values <- exp(coef(model_interaction))
@@
 # save fit results table  
-table_file1 = here("results", "tables", "resulttable3.rds")
-saveRDS(or_results, file = table_file1)
+table_file2 = here("results", "tables", "table2_odds_ratios.rds")
+saveRDS(or_results, file = table_file2)
 
@@
 print(tidy_model)
-```
-
-The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87).
-For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22).
-The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82).
-However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.
 
 # Fit a log-binomial model to estimate Relative Risks (RR)
-
-```{r}
 rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
@@
 # save fit results table  
-table_file1 = here("results", "tables", "resulttable4.rds")
-saveRDS(rr_results, file = table_file1)
-```
-
-The baseline relative risk is about 0.23 (95% CI: 0.08–0.46).
-For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35).
-The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.
+table_file3 = here("results", "tables", "table3_relative_risks.rds")
+saveRDS(rr_results, file = table_file3)
 
 # ROC Curve and AUC for the logistic regression model with interaction
-
-```{r}
 roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
@@
 print(paste("AUC:", auc_value))
-```
 
-The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing.
-It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.
+# Save ROC plot
+roc_fig <- recordPlot()
+saveRDS(roc_fig, file = here("results", "figures", "figure1_roc_logistic_model.rds"))
 
 # Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)
-
-```{r}
 adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
@@
 # save fit results table  
-table_file1 = here("results", "tables", "resulttable5.rds")
-saveRDS(adj_or_results, file = table_file1)
+table_file4 = here("results", "tables", "table4_adjusted_odds_ratios.rds")
+saveRDS(adj_or_results, file = table_file4)
 
@@
 ```
-
-The adjusted logistic regression model produces unstable estimates.
-The intercept's odds ratio is about 0.07, indicating very low baseline odds.
-The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant.
-The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals.
-These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable.
-This interpretation is derived solely from your output.
-
-# PART 4b: MODEL BUILDING
-
-```{r}
-# Set seed for reproducibility
-rdmseed <- 123
-
-# Reset seed before modeling
-set.seed(rdmseed)
-
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-```
-
-# linear model
-
-```{r}
-# Fit a standard linear regression model using predictors
-lm_model <- linear_reg() %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions from linear model and calculate RMSE
-cleaned_df$pred_lm <- predict(lm_model, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lm)
-
-# # Plot observed vs. predicted for Linear Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
-  geom_point(color = "darkblue", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average.
-This indicates limited accuracy for distinguishing between individuals with and without heart disease.
-
-# LASSO
-
-```{r}
-set.seed(rdmseed)
-
-# Prepare recipe for LASSO with normalization
-rec <- recipe(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df) %>%
-  step_dummy(elevated_BP, elevated_oldpeak) %>%   # Convert categorical predictors to dummies
-  step_normalize(all_numeric(), -all_outcomes())
-
-# Define LASSO model with a fixed penalty
-lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
-  set_engine("glmnet")
-
-# Create a workflow and fit the LASSO model
-wf <- workflow() %>%
-  add_recipe(rec) %>%
-  add_model(lasso_spec)
-
-lasso_fit <- fit(wf, data = cleaned_df)
-cleaned_df$pred_lasso <- predict(lasso_fit, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)
-
-# Plot observed vs. predicted for LASSO Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lasso)) +
-  geom_point(color = "darkviolet", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-```
-
-The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average.
-This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.
-
-# Random Forest
-
-```{r}
-set.seed(rdmseed)
-
-# Create a workflow and fit the Random Forest model
-rf_model <- rand_forest() %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-rf_fit <- rf_model %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions and evaluate model performance
-cleaned_df$pred_rf <- predict(rf_fit, cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_rf)
-
-# Plot observed vs. predicted for RF Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_rf)) +
-  geom_point(color = "darkgreen", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half.
-This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.
-
-# Tuning the LASSO model (without cross-validation)
-
-```{r}
-set.seed(rdmseed)
-
-# Create a grid of penalty values for LASSO
-lasso_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))
-
-# Define LASSO model with tunable penalty
-lasso_spec_tune <- linear_reg(penalty = tune()) %>%
-  set_engine("glmnet")
-
-# Create workflow and assign to lasso_tuning_workflow (new name)
-lasso_tuning_workflow <- workflow() %>%
-  add_model(lasso_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform grid search without CV (apparent resampling)
-lasso_tune <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = apparent(cleaned_df),
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-# Extract tuning results
-lasso_tune_df <- as.data.frame(lasso_tune$.metrics)
-
-# Plot LASSO tuning results
-ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
-  geom_line(size = 1, color = "steelblue") +
-  geom_point(size = 2, color = "firebrick") +
-  scale_x_log10() +
-  labs(
-    x = "Penalty (log scale)",
-    y = "RMSE",
-    title = "LASSO Tuning (No Cross-validation)",
-    subtitle = "Using apparent resampling"
-  ) +
-  theme_minimal()
-```
-
-The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties.
-This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.
-
-# Tuning for our Random Forest model (without cross-validation)
-
-```{r}
-# Create tuning grid for mtry and min_n
-rf_grid <- grid_regular(
-  mtry(range = c(1, 7)),
-  min_n(range = c(1, 21)),
-  levels = 7
-)
-
-# Define random forest model with tunable parameters
-rf_spec_tune <- rand_forest(
-  mtry = tune(),
-  min_n = tune(),
-  trees = 300
-) %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-# Build workflow
-rf_wf_tune <- workflow() %>%
-  add_model(rf_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform RF tuning without CV
-rf_tune <- tune_grid(
-  rf_wf_tune,
-  resamples = apparent(cleaned_df),
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-
-rf_tune_df <- as.data.frame(rf_tune$.metrics)
-
-# Plot RF tuning results
-ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
-  geom_tile() +
-  scale_fill_viridis_c() +
-  labs(
-    title = "Random Forest Tuning (No Cross-validation)",
-    x = "mtry",
-    y = "min_n",
-    fill = "RMSE"
-  ) +
-  theme_minimal()
-```
-
-The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739.
-This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used.
-In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.
-
-# Cross-validation for our LASSO model tuning
-
-```{r}
-set.seed(rdmseed)
-
-# Create 5-fold CV with 5 repeats
-cv_folds <- vfold_cv(cleaned_df, v = 5, repeats = 5)
-
-# Tune LASSO using CV
-lasso_tune_cv <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = cv_folds,
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-autoplot(lasso_tune_cv)
-```
-
-Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases.
-This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.
-
-# Cross-validation tuning for our random forest model
-
-```{r}
-# Tune RF using CV
-rf_tune_cv <- tune_grid(
-  rf_wf_tune,
-  resamples = cv_folds,
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-autoplot(rf_tune_cv)
-
-```
-
-Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry).
-In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values.
-This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.

### 2025-04-17 code/analysis-code/02_modeling.qmd
Few minor changes

--- 2025-04-05 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Final project"
+title: "Model Building"
 author: "Murtaza Yaqubi"
@@
 ---
-
-## General Background Information
-
-Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease”.
-It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
-This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
-As the leading cause of death in the U.S.
-and globally, CAD claimed 375,500 lives in the U.S.
-in 2021 and affects over 18 million American adults.
-
-## Description of data and data source
-
-This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses.
-Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India.
-This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-This dataset was obtained from mendeley data where it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-## Questions/Hypotheses to be addressed
-
-Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
-
-refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
-
-## Methods
-
-In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present.
-The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).
-
-We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values.
-Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression.
-Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.
-
-To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups.
-Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar.
-Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.
-
-In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods.
-These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE).
-By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.
-
-## **Conclusion**
-
-Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present.
-Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups.
-In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance.
-Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup.
-Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease.
-Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.
-
-## Data aquisition
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-# Load necessary libraries
-```{r}
-library(knitr)
-library(readr)
-library(dplyr)
-library(ggplot2)
-library(ggcorrplot)
-library(vcd)
-library(caret)        # For data partitioning and model training
-library(randomForest) # For Random Forest modeling
-library(tidyverse)
-library(tidymodels)
-library(broom)
-library(here)           # For file paths
-library(pROC)
-library(naniar)         # For gg_miss_var()
-```
-
-# PART 1: DATA IMPORT AND CLEANING
-```{r}
-# Import the dataset
-cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
-
-# Basic exploration of the dataset
-head(cardio_df)           # View the first few rows
-colnames(cardio_df)       # Check column names
-dim(cardio_df)            # Check dimensions (should be 1025 rows x 14 columns)
-summary(cardio_df)        # Get a summary of the data
-glimpse(cardio_df)        # Glimpse at the data structure
-
-# Visualize missing values for each variable
-gg_miss_var(cardio_df)    # Display number of NAs
-
-# Transform variables into suitable types and add descriptive labels
-cardio_df <- cardio_df %>%
-  mutate(
-    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
-    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
-                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
-    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
-    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
-                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
-    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
-    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
-    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
-  )
-
-# Print the transformed dataset for verification
-print(cardio_df)
-```
-
-# PART 2: EXPLORATORY DATA ANALYSIS (EDA)
-
-```{r}
-# Calculate summary statistics for selected variables
-summary_stats <- cardio_df %>% 
-  summarise(
-    # Age statistics
-    Age_min = min(age, na.rm = TRUE),
-    Age_max = max(age, na.rm = TRUE),
-    Age_mean = mean(age, na.rm = TRUE),
-    Age_sd = sd(age, na.rm = TRUE),
-
-    # Resting Blood Pressure statistics
-    RestingBP_min = min(restingBP, na.rm = TRUE),
-    RestingBP_max = max(restingBP, na.rm = TRUE),
-    RestingBP_mean = mean(restingBP, na.rm = TRUE),
-    RestingBP_sd = sd(restingBP, na.rm = TRUE),
-
-    # Serum Cholesterol statistics (excluding zeros)
-    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
-
-    # Maximum Heart Rate statistics
-    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
-
-    # Oldpeak statistics (excluding zeros)
-    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
-    Oldpeak_max = max(oldpeak, na.rm = TRUE),
-    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
-    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
-  )
-
-# Print summary statistics
-print(summary_stats)
-
-summary_df = skimr::skim(summary_stats)
-print(summary_stats)
-# save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Age distribution
-p1 <- ggplot(cardio_df, aes(x = age)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
-  geom_density(color = "darkblue", linewidth = 1) +
-  labs(title = "Age Distribution", x = "Age", y = "Density") +
-  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
-  theme_bw()
-
-plot(p1)
-figure_file = here("results", "figures","Age distribution.png")
-ggsave(filename = figure_file, plot=p1)
-
-# Serum Cholesterol distribution (excluding zeros)
-p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
-  geom_density(color = "darkgreen", linewidth = 1) +
-  labs(title = "Serum Cholesterol Distribution", 
-       x = "Serum Cholesterol (mg/dL)", y = "Density") +
-  theme_minimal()
-
-plot(p2)
-figure_file = here("results", "figures","Serum Cholestrol.png")
-ggsave(filename = figure_file, plot=p2)
-
-# Resting Blood Pressure distribution (excluding zeros)
-p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
-  geom_density(color = "darkred", linewidth = 1) +
-  labs(title = "Resting Blood Pressure Distribution", 
-       x = "Resting Blood Pressure (mmHG)", y = "Density") +
-  theme_bw()
-
-plot(p3)
-figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
-ggsave(filename = figure_file, plot=p3)
-
-# Boxplot for resting blood pressure by heart disease status
-p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
-  geom_boxplot() +
-  labs(title = "Resting Blood Pressure by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Resting Blood Pressure (mmHg)") +
-  theme_minimal()
-
-plot(p4)
-figure_file = here("results", "figures","restingBP by heart disease status.png")
-ggsave(filename = figure_file, plot=p4)
-
-# Violin plot for oldpeak by heart disease status
-p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
-  geom_violin(trim = FALSE) +
-  labs(title = "Oldpeak Distribution by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Oldpeak") +
-  theme_minimal()
-
-plot(p5)
-figure_file = here("results", "figures","oldpeak by heart disease status.png")
-ggsave(filename = figure_file, plot=p5)
-
-# Create the correlation matrix for selected numeric variables
-numeric_vars <- cardio_df %>% 
-  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
-cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
-
-# Plot the correlation matrix with ggcorrplot
-p6 <- ggcorrplot(cor_matrix, 
-           lab = TRUE,                         
-           lab_size = 3,                       
-           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
-           title = "Correlation Matrix", 
-           ggtheme = theme_minimal())
-
-plot(p6)
-figure_file = here("results", "figures","correlation matrix.png")
-ggsave(filename = figure_file, plot=p6)     
-```
-
-# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
-
-```{r}
-# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
-cleaned_df <- cardio_df %>%
-  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
-  mutate(
-    # Create a binary variable for blood pressure:
-    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
-    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),
-
-    # Create a binary indicator for ST depression:
-    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
-    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
-  ) %>% 
-  # Select the variables we want to keep
-  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
-         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)
-
-# Print a sample of the cleaned dataset
-head(cleaned_df)
-
-# Save the processed data
-saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")
-
-# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
-summary_table <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-print(summary_table)
-
-summary_df = skimr::skim(summary_table)
-print(summary_table)
-# save to file
-summarytable_file = here("results", "tables", "summarytable2.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Interaction Visualization: Heatmap
-interaction_counts <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-
-p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
-  geom_tile(color = "white", size = 0.5) +
-  geom_text(aes(label = count), color = "black", size = 5) +
-  facet_wrap(~ target, ncol = 2) +
-  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
-  labs(title = "Interaction between Blood Pressure and ST Depression",
-       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
-       x = "Blood Pressure Status",
-       y = "ST Depression Status",
-       fill = "Count") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5)
-  )
-
-plot(p7)
-figure_file = here("results", "figures","interaction heatmap.png")
-ggsave(filename = figure_file, plot=p7)
-
-# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
-p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
-  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
-  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
-       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
-       x = "Blood Pressure Status",
-       y = "Count",
-       fill = "Heart Disease") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5),
-    legend.position = "bottom"
-  )
-
-plot(p8)
-figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
-ggsave(filename = figure_file, plot=p8)
-```
-
-# PART 4a: INFERENTIAL STATISTICAL ANALYSIS
-
-# Chi-Square Tests
-
-```{r}
-# Create a contingency table for blood pressure status vs. heart disease
-table_bp_target <- table(cleaned_df$elevated_BP, cleaned_df$target)
-print(table_bp_target)
-chisq_test_bp <- chisq.test(table_bp_target)
-print(chisq_test_bp)
-
-# Create a contingency table for ST depression vs. heart disease
-table_oldpeak_target <- table(cleaned_df$elevated_oldpeak, cleaned_df$target)
-print(table_oldpeak_target)
-chisq_test_oldpeak <- chisq.test(table_oldpeak_target)
-print(chisq_test_oldpeak)
-```
-
-The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup.
-For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels.
-Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.
-
-# Logistic Regression Analysis
-
-```{r}
-# Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-
-# Fit a logistic regression model with elevated_BP and elevated_oldpeak as predictors
-model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-             data = cleaned_df, family = binomial)
-summary(model)
-
-# Fit a logistic regression model including the interaction term
-model_interaction <- glm(target_binary ~ elevated_BP * elevated_oldpeak, 
-                         data = cleaned_df, family = binomial)
-summary(model_interaction)
-```
-
-In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero.
-However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own.
-When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance.
-These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.
-
-# Extract Odds Ratios (OR) from the logistic regression model with interaction
-
-```{r}
-or_values <- exp(coef(model_interaction))
-or_ci <- exp(confint(model_interaction))
-or_results <- data.frame(
-  Term = names(or_values),
-  OR = or_values,
-  CI_lower = or_ci[, 1],
-  CI_upper = or_ci[, 2]
-)
-print("Odds Ratios (with 95% CI):")
-print(or_results)
-
-# Tidy the model output and compute odds ratios and 95% CI
-tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_model)
-```
-
-The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87).
-For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22).
-The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82).
-However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.
-
-# Fit a log-binomial model to estimate Relative Risks (RR)
-
-```{r}
-rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-                data = cleaned_df, family = binomial(link = "log"))
-rr_values <- exp(coef(rr_model))
-rr_ci <- exp(confint(rr_model))
-rr_results <- data.frame(
-  Term = names(rr_values),
-  RR = rr_values,
-  CI_lower = rr_ci[, 1],
-  CI_upper = rr_ci[, 2]
-)
-print("Relative Risks (with 95% CI):")
-print(rr_results)
-```
-
-The baseline relative risk is about 0.23 (95% CI: 0.08–0.46).
-For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35).
-The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.
-
-# ROC Curve and AUC for the logistic regression model with interaction
-
-```{r}
-roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
-plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
-auc_value <- auc(roc_curve)
-print(paste("AUC:", auc_value))
-```
-
-The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing.
-It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.
-
-# Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)
-
-```{r}
-adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
-                      data = cleaned_df, family = binomial)
-summary(adjusted_model)
-
-# Extract Odds Ratios (OR) for the adjusted model
-adj_or_values <- exp(coef(adjusted_model))
-adj_or_ci <- exp(confint(adjusted_model))
-adj_or_results <- data.frame(
-  Term = names(adj_or_values),
-  OR = adj_or_values,
-  CI_lower = adj_or_ci[, 1],
-  CI_upper = adj_or_ci[, 2]
-)
-print("Adjusted Odds Ratios (with 95% CI):")
-print(adj_or_results)
-
-# Tidy the adjusted model output and compute odds ratios and 95% CI
-tidy_adjusted_model <- tidy(adjusted_model, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_adjusted_model)
-```
-
-The adjusted logistic regression model produces unstable estimates.
-The intercept's odds ratio is about 0.07, indicating very low baseline odds.
-The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant.
-The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals.
-These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable.
-This interpretation is derived solely from your output.
-
-# PART 4b: MODEL BUILDING
 
@@
   mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-```
 
-# linear model
-
-```{r}
 # Fit a standard linear regression model using predictors
@@
 
-# # Plot observed vs. predicted for Linear Model
+# Plot observed vs. predicted for Linear Model
 ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
@@
   scale_y_continuous(limits = c(0, 1))
-
-```
-
-The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average.
-This indicates limited accuracy for distinguishing between individuals with and without heart disease.
-
-# LASSO
-
-```{r}
-set.seed(rdmseed)
 
@@
   scale_y_continuous(limits = c(0, 1))
-```
-
-The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average.
-This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.
-
-# Random Forest
-
-```{r}
-set.seed(rdmseed)
 
@@
 
+# Save RMSEs from all models to a table
+rmse_df <- tibble(
+  Model = c("Linear", "LASSO", "Random Forest"),
+  RMSE = c(
+    rmse(cleaned_df, truth = target_binary, estimate = pred_lm)$.estimate,
+    rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)$.estimate,
+    rmse(cleaned_df, truth = target_binary, estimate = pred_rf)$.estimate
+  )
+)
+saveRDS(rmse_df, file = here("results", "tables", "table5_model_rmse_comparison.rds"))
+
 # Plot observed vs. predicted for RF Model
@@
   scale_y_continuous(limits = c(0, 1))
-
 ```
-
-The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half.
-This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.
-
-# Tuning the LASSO model (without cross-validation)
-
-```{r}
-set.seed(rdmseed)
-
-# Create a grid of penalty values for LASSO
-lasso_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))
-
-# Define LASSO model with tunable penalty
-lasso_spec_tune <- linear_reg(penalty = tune()) %>%
-  set_engine("glmnet")
-
-# Create workflow and assign to lasso_tuning_workflow (new name)
-lasso_tuning_workflow <- workflow() %>%
-  add_model(lasso_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform grid search without CV (apparent resampling)
-lasso_tune <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = apparent(cleaned_df),
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-# Extract tuning results
-lasso_tune_df <- as.data.frame(lasso_tune$.metrics)
-
-# Plot LASSO tuning results
-ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
-  geom_line(size = 1, color = "steelblue") +
-  geom_point(size = 2, color = "firebrick") +
-  scale_x_log10() +
-  labs(
-    x = "Penalty (log scale)",
-    y = "RMSE",
-    title = "LASSO Tuning (No Cross-validation)",
-    subtitle = "Using apparent resampling"
-  ) +
-  theme_minimal()
-```
-
-The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties.
-This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.
-
-# Tuning for our Random Forest model (without cross-validation)
-
-```{r}
-# Create tuning grid for mtry and min_n
-rf_grid <- grid_regular(
-  mtry(range = c(1, 7)),
-  min_n(range = c(1, 21)),
-  levels = 7
-)
-
-# Define random forest model with tunable parameters
-rf_spec_tune <- rand_forest(
-  mtry = tune(),
-  min_n = tune(),
-  trees = 300
-) %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-# Build workflow
-rf_wf_tune <- workflow() %>%
-  add_model(rf_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform RF tuning without CV
-rf_tune <- tune_grid(
-  rf_wf_tune,
-  resamples = apparent(cleaned_df),
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-
-rf_tune_df <- as.data.frame(rf_tune$.metrics)
-
-# Plot RF tuning results
-ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
-  geom_tile() +
-  scale_fill_viridis_c() +
-  labs(
-    title = "Random Forest Tuning (No Cross-validation)",
-    x = "mtry",
-    y = "min_n",
-    fill = "RMSE"
-  ) +
-  theme_minimal()
-```
-
-The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739.
-This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used.
-In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.
-
-# Cross-validation for our LASSO model tuning
-
-```{r}
-set.seed(rdmseed)
-
-# Create 5-fold CV with 5 repeats
-cv_folds <- vfold_cv(cleaned_df, v = 5, repeats = 5)
-
-# Tune LASSO using CV
-lasso_tune_cv <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = cv_folds,
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-autoplot(lasso_tune_cv)
-```
-
-Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases.
-This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.
-
-# Cross-validation tuning for our random forest model
-
-```{r}
-# Tune RF using CV
-rf_tune_cv <- tune_grid(
-  rf_wf_tune,
-  resamples = cv_folds,
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-autoplot(rf_tune_cv)
-
-```
-
-Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry).
-In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values.
-This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.

### 2025-04-17 code/analysis-code/03_model_tuning.qmd
Few minor changes

--- 2025-04-05 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Final project"
+title: "Model Tuning"
 author: "Murtaza Yaqubi"
@@
 
-## General Background Information
-
-Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease”.
-It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
-This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
-As the leading cause of death in the U.S.
-and globally, CAD claimed 375,500 lives in the U.S.
-in 2021 and affects over 18 million American adults.
-
-## Description of data and data source
-
-This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses.
-Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India.
-This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-This dataset was obtained from mendeley data where it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-## Questions/Hypotheses to be addressed
-
-Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
-
-refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
-
-## Methods
-
-In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present.
-The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).
-
-We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values.
-Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression.
-Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.
-
-To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups.
-Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar.
-Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.
-
-In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods.
-These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE).
-By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.
-
-## **Conclusion**
-
-Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present.
-Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups.
-In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance.
-Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup.
-Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease.
-Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.
-
-## Data aquisition
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-# Load necessary libraries
 ```{r}
-library(knitr)
-library(readr)
-library(dplyr)
-library(ggplot2)
-library(ggcorrplot)
-library(vcd)
-library(caret)        # For data partitioning and model training
-library(randomForest) # For Random Forest modeling
-library(tidyverse)
-library(tidymodels)
-library(broom)
-library(here)           # For file paths
-library(pROC)
-library(naniar)         # For gg_miss_var()
-```
-
-# PART 1: DATA IMPORT AND CLEANING
-```{r}
-# Import the dataset
-cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
-
-# Basic exploration of the dataset
-head(cardio_df)           # View the first few rows
-colnames(cardio_df)       # Check column names
-dim(cardio_df)            # Check dimensions (should be 1025 rows x 14 columns)
-summary(cardio_df)        # Get a summary of the data
-glimpse(cardio_df)        # Glimpse at the data structure
-
-# Visualize missing values for each variable
-gg_miss_var(cardio_df)    # Display number of NAs
-
-# Transform variables into suitable types and add descriptive labels
-cardio_df <- cardio_df %>%
-  mutate(
-    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
-    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
-                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
-    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
-    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
-                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
-    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
-    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
-    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
-  )
-
-# Print the transformed dataset for verification
-print(cardio_df)
-```
-
-# PART 2: EXPLORATORY DATA ANALYSIS (EDA)
-
-```{r}
-# Calculate summary statistics for selected variables
-summary_stats <- cardio_df %>% 
-  summarise(
-    # Age statistics
-    Age_min = min(age, na.rm = TRUE),
-    Age_max = max(age, na.rm = TRUE),
-    Age_mean = mean(age, na.rm = TRUE),
-    Age_sd = sd(age, na.rm = TRUE),
-
-    # Resting Blood Pressure statistics
-    RestingBP_min = min(restingBP, na.rm = TRUE),
-    RestingBP_max = max(restingBP, na.rm = TRUE),
-    RestingBP_mean = mean(restingBP, na.rm = TRUE),
-    RestingBP_sd = sd(restingBP, na.rm = TRUE),
-
-    # Serum Cholesterol statistics (excluding zeros)
-    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
-
-    # Maximum Heart Rate statistics
-    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
-
-    # Oldpeak statistics (excluding zeros)
-    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
-    Oldpeak_max = max(oldpeak, na.rm = TRUE),
-    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
-    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
-  )
-
-# Print summary statistics
-print(summary_stats)
-
-summary_df = skimr::skim(summary_stats)
-print(summary_stats)
-# save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Age distribution
-p1 <- ggplot(cardio_df, aes(x = age)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
-  geom_density(color = "darkblue", linewidth = 1) +
-  labs(title = "Age Distribution", x = "Age", y = "Density") +
-  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
-  theme_bw()
-
-plot(p1)
-figure_file = here("results", "figures","Age distribution.png")
-ggsave(filename = figure_file, plot=p1)
-
-# Serum Cholesterol distribution (excluding zeros)
-p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
-  geom_density(color = "darkgreen", linewidth = 1) +
-  labs(title = "Serum Cholesterol Distribution", 
-       x = "Serum Cholesterol (mg/dL)", y = "Density") +
-  theme_minimal()
-
-plot(p2)
-figure_file = here("results", "figures","Serum Cholestrol.png")
-ggsave(filename = figure_file, plot=p2)
-
-# Resting Blood Pressure distribution (excluding zeros)
-p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
-  geom_density(color = "darkred", linewidth = 1) +
-  labs(title = "Resting Blood Pressure Distribution", 
-       x = "Resting Blood Pressure (mmHG)", y = "Density") +
-  theme_bw()
-
-plot(p3)
-figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
-ggsave(filename = figure_file, plot=p3)
-
-# Boxplot for resting blood pressure by heart disease status
-p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
-  geom_boxplot() +
-  labs(title = "Resting Blood Pressure by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Resting Blood Pressure (mmHg)") +
-  theme_minimal()
-
-plot(p4)
-figure_file = here("results", "figures","restingBP by heart disease status.png")
-ggsave(filename = figure_file, plot=p4)
-
-# Violin plot for oldpeak by heart disease status
-p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
-  geom_violin(trim = FALSE) +
-  labs(title = "Oldpeak Distribution by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Oldpeak") +
-  theme_minimal()
-
-plot(p5)
-figure_file = here("results", "figures","oldpeak by heart disease status.png")
-ggsave(filename = figure_file, plot=p5)
-
-# Create the correlation matrix for selected numeric variables
-numeric_vars <- cardio_df %>% 
-  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
-cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
-
-# Plot the correlation matrix with ggcorrplot
-p6 <- ggcorrplot(cor_matrix, 
-           lab = TRUE,                         
-           lab_size = 3,                       
-           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
-           title = "Correlation Matrix", 
-           ggtheme = theme_minimal())
-
-plot(p6)
-figure_file = here("results", "figures","correlation matrix.png")
-ggsave(filename = figure_file, plot=p6)     
-```
-
-# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
-
-```{r}
-# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
-cleaned_df <- cardio_df %>%
-  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
-  mutate(
-    # Create a binary variable for blood pressure:
-    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
-    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),
-
-    # Create a binary indicator for ST depression:
-    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
-    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
-  ) %>% 
-  # Select the variables we want to keep
-  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
-         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)
-
-# Print a sample of the cleaned dataset
-head(cleaned_df)
-
-# Save the processed data
-saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")
-
-# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
-summary_table <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-print(summary_table)
-
-summary_df = skimr::skim(summary_table)
-print(summary_table)
-# save to file
-summarytable_file = here("results", "tables", "summarytable2.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Interaction Visualization: Heatmap
-interaction_counts <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-
-p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
-  geom_tile(color = "white", size = 0.5) +
-  geom_text(aes(label = count), color = "black", size = 5) +
-  facet_wrap(~ target, ncol = 2) +
-  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
-  labs(title = "Interaction between Blood Pressure and ST Depression",
-       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
-       x = "Blood Pressure Status",
-       y = "ST Depression Status",
-       fill = "Count") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5)
-  )
-
-plot(p7)
-figure_file = here("results", "figures","interaction heatmap.png")
-ggsave(filename = figure_file, plot=p7)
-
-# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
-p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
-  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
-  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
-       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
-       x = "Blood Pressure Status",
-       y = "Count",
-       fill = "Heart Disease") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5),
-    legend.position = "bottom"
-  )
-
-plot(p8)
-figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
-ggsave(filename = figure_file, plot=p8)
-```
-
-# PART 4a: INFERENTIAL STATISTICAL ANALYSIS
-
-# Chi-Square Tests
-
-```{r}
-# Create a contingency table for blood pressure status vs. heart disease
-table_bp_target <- table(cleaned_df$elevated_BP, cleaned_df$target)
-print(table_bp_target)
-chisq_test_bp <- chisq.test(table_bp_target)
-print(chisq_test_bp)
-
-# Create a contingency table for ST depression vs. heart disease
-table_oldpeak_target <- table(cleaned_df$elevated_oldpeak, cleaned_df$target)
-print(table_oldpeak_target)
-chisq_test_oldpeak <- chisq.test(table_oldpeak_target)
-print(chisq_test_oldpeak)
-```
-
-The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup.
-For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels.
-Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.
-
-# Logistic Regression Analysis
-
-```{r}
-# Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-
-# Fit a logistic regression model with elevated_BP and elevated_oldpeak as predictors
-model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-             data = cleaned_df, family = binomial)
-summary(model)
-
-# Fit a logistic regression model including the interaction term
-model_interaction <- glm(target_binary ~ elevated_BP * elevated_oldpeak, 
-                         data = cleaned_df, family = binomial)
-summary(model_interaction)
-```
-
-In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero.
-However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own.
-When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance.
-These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.
-
-# Extract Odds Ratios (OR) from the logistic regression model with interaction
-
-```{r}
-or_values <- exp(coef(model_interaction))
-or_ci <- exp(confint(model_interaction))
-or_results <- data.frame(
-  Term = names(or_values),
-  OR = or_values,
-  CI_lower = or_ci[, 1],
-  CI_upper = or_ci[, 2]
-)
-print("Odds Ratios (with 95% CI):")
-print(or_results)
-
-# Tidy the model output and compute odds ratios and 95% CI
-tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_model)
-```
-
-The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87).
-For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22).
-The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82).
-However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.
-
-# Fit a log-binomial model to estimate Relative Risks (RR)
-
-```{r}
-rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-                data = cleaned_df, family = binomial(link = "log"))
-rr_values <- exp(coef(rr_model))
-rr_ci <- exp(confint(rr_model))
-rr_results <- data.frame(
-  Term = names(rr_values),
-  RR = rr_values,
-  CI_lower = rr_ci[, 1],
-  CI_upper = rr_ci[, 2]
-)
-print("Relative Risks (with 95% CI):")
-print(rr_results)
-```
-
-The baseline relative risk is about 0.23 (95% CI: 0.08–0.46).
-For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35).
-The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.
-
-# ROC Curve and AUC for the logistic regression model with interaction
-
-```{r}
-roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
-plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
-auc_value <- auc(roc_curve)
-print(paste("AUC:", auc_value))
-```
-
-The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing.
-It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.
-
-# Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)
-
-```{r}
-adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
-                      data = cleaned_df, family = binomial)
-summary(adjusted_model)
-
-# Extract Odds Ratios (OR) for the adjusted model
-adj_or_values <- exp(coef(adjusted_model))
-adj_or_ci <- exp(confint(adjusted_model))
-adj_or_results <- data.frame(
-  Term = names(adj_or_values),
-  OR = adj_or_values,
-  CI_lower = adj_or_ci[, 1],
-  CI_upper = adj_or_ci[, 2]
-)
-print("Adjusted Odds Ratios (with 95% CI):")
-print(adj_or_results)
-
-# Tidy the adjusted model output and compute odds ratios and 95% CI
-tidy_adjusted_model <- tidy(adjusted_model, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_adjusted_model)
-```
-
-The adjusted logistic regression model produces unstable estimates.
-The intercept's odds ratio is about 0.07, indicating very low baseline odds.
-The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant.
-The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals.
-These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable.
-This interpretation is derived solely from your output.
-
-# PART 4b: MODEL BUILDING
-
-```{r}
-# Set seed for reproducibility
-rdmseed <- 123
-
-# Reset seed before modeling
-set.seed(rdmseed)
-
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-```
-
-# linear model
-
-```{r}
-# Fit a standard linear regression model using predictors
-lm_model <- linear_reg() %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions from linear model and calculate RMSE
-cleaned_df$pred_lm <- predict(lm_model, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lm)
-
-# # Plot observed vs. predicted for Linear Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
-  geom_point(color = "darkblue", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average.
-This indicates limited accuracy for distinguishing between individuals with and without heart disease.
-
-# LASSO
-
-```{r}
-set.seed(rdmseed)
-
-# Prepare recipe for LASSO with normalization
-rec <- recipe(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df) %>%
-  step_dummy(elevated_BP, elevated_oldpeak) %>%   # Convert categorical predictors to dummies
-  step_normalize(all_numeric(), -all_outcomes())
-
-# Define LASSO model with a fixed penalty
-lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
-  set_engine("glmnet")
-
-# Create a workflow and fit the LASSO model
-wf <- workflow() %>%
-  add_recipe(rec) %>%
-  add_model(lasso_spec)
-
-lasso_fit <- fit(wf, data = cleaned_df)
-cleaned_df$pred_lasso <- predict(lasso_fit, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)
-
-# Plot observed vs. predicted for LASSO Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lasso)) +
-  geom_point(color = "darkviolet", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-```
-
-The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average.
-This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.
-
-# Random Forest
-
-```{r}
-set.seed(rdmseed)
-
-# Create a workflow and fit the Random Forest model
-rf_model <- rand_forest() %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-rf_fit <- rf_model %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions and evaluate model performance
-cleaned_df$pred_rf <- predict(rf_fit, cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_rf)
-
-# Plot observed vs. predicted for RF Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_rf)) +
-  geom_point(color = "darkgreen", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half.
-This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.
-
 # Tuning the LASSO model (without cross-validation)
-
-```{r}
 set.seed(rdmseed)
@@
 
+# Identify best LASSO penalty
+best_lasso <- lasso_tune_df %>% filter(.estimate == min(.estimate))
+print("Best LASSO Penalty (No CV):")
+print(best_lasso)
+
 # Plot LASSO tuning results
@@
   theme_minimal()
-```
 
-The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties.
-This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.
+# Save LASSO tuning plot
+lasso_tune_plot <- last_plot()
+ggsave(filename = here("results", "figures", "figure2_lasso_tuning_plot.png"), plot = lasso_tune_plot)
 
 # Tuning for our Random Forest model (without cross-validation)
-
-```{r}
-# Create tuning grid for mtry and min_n
 rf_grid <- grid_regular(
@@
 
+# Identify best RF tuning parameters
+best_rf <- rf_tune_df %>% filter(.estimate == min(.estimate))
+print("Best Random Forest Parameters (No CV):")
+print(best_rf)
+
 # Plot RF tuning results
@@
   theme_minimal()
-```
 
-The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739.
-This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used.
-In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.
+# Save Random Forest tuning heatmap
+rf_tune_plot <- last_plot()
+ggsave(filename = here("results", "figures", "figure3_rf_tuning_heatmap.png"), plot = rf_tune_plot)
 
 # Cross-validation for our LASSO model tuning
-
-```{r}
 set.seed(rdmseed)
@@
 autoplot(lasso_tune_cv)
-```
-
-Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases.
-This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.
 
 # Cross-validation tuning for our random forest model
-
-```{r}
-# Tune RF using CV
 rf_tune_cv <- tune_grid(
@@
 )
+
 autoplot(rf_tune_cv)
-
 ```
-
-Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry).
-In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values.
-This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.

### 2025-04-17 code/eda-code/01_summary_stats.qmd
Few minor changes

---
title: "Summary statistics"
author: "Murtaza"
format: html
editor: visual
---

```{r}
# Calculate summary statistics for selected variables
summary_stats <- cardio_df %>% 
  summarise(
    Age_min = min(age, na.rm = TRUE),
    Age_max = max(age, na.rm = TRUE),
    Age_mean = mean(age, na.rm = TRUE),
    Age_sd = sd(age, na.rm = TRUE),

    RestingBP_min = min(restingBP, na.rm = TRUE),
    RestingBP_max = max(restingBP, na.rm = TRUE),
    RestingBP_mean = mean(restingBP, na.rm = TRUE),
    RestingBP_sd = sd(restingBP, na.rm = TRUE),

    SerumCholestrol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_max = max(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_mean = mean(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
    SerumCholestrol_sd = sd(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),

    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),

    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
    Oldpeak_max = max(oldpeak, na.rm = TRUE),
    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
  )

# Reshape into long format, then wide table
summary_table_formatted <- summary_stats %>%
  pivot_longer(cols = everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Variable, values_from = value)

# View formatted table
summary_table_formatted

# save to file
summarytable_file = here("results", "tables", "table1_summary_statistics.rds")
saveRDS(summary_table_formatted, file = summarytable_file)
```

### 2025-04-17 code/eda-code/02_visualizations.qmd
Few minor changes

--- 2025-04-05 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Final project"
+title: "Visualization"
 author: "Murtaza Yaqubi"
@@
 
-## General Background Information
-
-Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease”.
-It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
-This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
-As the leading cause of death in the U.S.
-and globally, CAD claimed 375,500 lives in the U.S.
-in 2021 and affects over 18 million American adults.
-
-## Description of data and data source
-
-This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses.
-Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India.
-This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-This dataset was obtained from mendeley data where it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-## Questions/Hypotheses to be addressed
-
-Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
-
-refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
-
-## Methods
-
-In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present.
-The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).
-
-We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values.
-Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression.
-Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.
-
-To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups.
-Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar.
-Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.
-
-In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods.
-These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE).
-By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.
-
-## **Conclusion**
-
-Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present.
-Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups.
-In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance.
-Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup.
-Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease.
-Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.
-
-## Data aquisition
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-# Load necessary libraries
 ```{r}
-library(knitr)
-library(readr)
-library(dplyr)
-library(ggplot2)
-library(ggcorrplot)
-library(vcd)
-library(caret)        # For data partitioning and model training
-library(randomForest) # For Random Forest modeling
-library(tidyverse)
-library(tidymodels)
-library(broom)
-library(here)           # For file paths
-library(pROC)
-library(naniar)         # For gg_miss_var()
-```
-
-# PART 1: DATA IMPORT AND CLEANING
-```{r}
-# Import the dataset
-cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
-
-# Basic exploration of the dataset
-head(cardio_df)           # View the first few rows
-colnames(cardio_df)       # Check column names
-dim(cardio_df)            # Check dimensions (should be 1025 rows x 14 columns)
-summary(cardio_df)        # Get a summary of the data
-glimpse(cardio_df)        # Glimpse at the data structure
-
-# Visualize missing values for each variable
-gg_miss_var(cardio_df)    # Display number of NAs
-
-# Transform variables into suitable types and add descriptive labels
-cardio_df <- cardio_df %>%
-  mutate(
-    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
-    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
-                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
-    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
-    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
-                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
-    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
-    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
-    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
-  )
-
-# Print the transformed dataset for verification
-print(cardio_df)
-```
-
-# PART 2: EXPLORATORY DATA ANALYSIS (EDA)
-
-```{r}
-# Calculate summary statistics for selected variables
-summary_stats <- cardio_df %>% 
-  summarise(
-    # Age statistics
-    Age_min = min(age, na.rm = TRUE),
-    Age_max = max(age, na.rm = TRUE),
-    Age_mean = mean(age, na.rm = TRUE),
-    Age_sd = sd(age, na.rm = TRUE),
-
-    # Resting Blood Pressure statistics
-    RestingBP_min = min(restingBP, na.rm = TRUE),
-    RestingBP_max = max(restingBP, na.rm = TRUE),
-    RestingBP_mean = mean(restingBP, na.rm = TRUE),
-    RestingBP_sd = sd(restingBP, na.rm = TRUE),
-
-    # Serum Cholesterol statistics (excluding zeros)
-    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
-
-    # Maximum Heart Rate statistics
-    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
-
-    # Oldpeak statistics (excluding zeros)
-    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
-    Oldpeak_max = max(oldpeak, na.rm = TRUE),
-    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
-    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
-  )
-
-# Print summary statistics
-print(summary_stats)
-
-summary_df = skimr::skim(summary_stats)
-print(summary_stats)
-# save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-
 # Age distribution
@@
 ```
-
-# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
-
-```{r}
-# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
-cleaned_df <- cardio_df %>%
-  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
-  mutate(
-    # Create a binary variable for blood pressure:
-    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
-    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),
-
-    # Create a binary indicator for ST depression:
-    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
-    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
-  ) %>% 
-  # Select the variables we want to keep
-  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
-         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)
-
-# Print a sample of the cleaned dataset
-head(cleaned_df)
-
-# Save the processed data
-saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")
-
-# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
-summary_table <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-print(summary_table)
-
-summary_df = skimr::skim(summary_table)
-print(summary_table)
-# save to file
-summarytable_file = here("results", "tables", "summarytable2.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Interaction Visualization: Heatmap
-interaction_counts <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-
-p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
-  geom_tile(color = "white", size = 0.5) +
-  geom_text(aes(label = count), color = "black", size = 5) +
-  facet_wrap(~ target, ncol = 2) +
-  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
-  labs(title = "Interaction between Blood Pressure and ST Depression",
-       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
-       x = "Blood Pressure Status",
-       y = "ST Depression Status",
-       fill = "Count") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5)
-  )
-
-plot(p7)
-figure_file = here("results", "figures","interaction heatmap.png")
-ggsave(filename = figure_file, plot=p7)
-
-# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
-p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
-  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
-  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
-       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
-       x = "Blood Pressure Status",
-       y = "Count",
-       fill = "Heart Disease") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5),
-    legend.position = "bottom"
-  )
-
-plot(p8)
-figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
-ggsave(filename = figure_file, plot=p8)
-```
-
-# PART 4a: INFERENTIAL STATISTICAL ANALYSIS
-
-# Chi-Square Tests
-
-```{r}
-# Create a contingency table for blood pressure status vs. heart disease
-table_bp_target <- table(cleaned_df$elevated_BP, cleaned_df$target)
-print(table_bp_target)
-chisq_test_bp <- chisq.test(table_bp_target)
-print(chisq_test_bp)
-
-# Create a contingency table for ST depression vs. heart disease
-table_oldpeak_target <- table(cleaned_df$elevated_oldpeak, cleaned_df$target)
-print(table_oldpeak_target)
-chisq_test_oldpeak <- chisq.test(table_oldpeak_target)
-print(chisq_test_oldpeak)
-```
-
-The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup.
-For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels.
-Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.
-
-# Logistic Regression Analysis
-
-```{r}
-# Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-
-# Fit a logistic regression model with elevated_BP and elevated_oldpeak as predictors
-model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-             data = cleaned_df, family = binomial)
-summary(model)
-
-# Fit a logistic regression model including the interaction term
-model_interaction <- glm(target_binary ~ elevated_BP * elevated_oldpeak, 
-                         data = cleaned_df, family = binomial)
-summary(model_interaction)
-```
-
-In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero.
-However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own.
-When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance.
-These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.
-
-# Extract Odds Ratios (OR) from the logistic regression model with interaction
-
-```{r}
-or_values <- exp(coef(model_interaction))
-or_ci <- exp(confint(model_interaction))
-or_results <- data.frame(
-  Term = names(or_values),
-  OR = or_values,
-  CI_lower = or_ci[, 1],
-  CI_upper = or_ci[, 2]
-)
-print("Odds Ratios (with 95% CI):")
-print(or_results)
-
-# Tidy the model output and compute odds ratios and 95% CI
-tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_model)
-```
-
-The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87).
-For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22).
-The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82).
-However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.
-
-# Fit a log-binomial model to estimate Relative Risks (RR)
-
-```{r}
-rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-                data = cleaned_df, family = binomial(link = "log"))
-rr_values <- exp(coef(rr_model))
-rr_ci <- exp(confint(rr_model))
-rr_results <- data.frame(
-  Term = names(rr_values),
-  RR = rr_values,
-  CI_lower = rr_ci[, 1],
-  CI_upper = rr_ci[, 2]
-)
-print("Relative Risks (with 95% CI):")
-print(rr_results)
-```
-
-The baseline relative risk is about 0.23 (95% CI: 0.08–0.46).
-For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35).
-The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.
-
-# ROC Curve and AUC for the logistic regression model with interaction
-
-```{r}
-roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
-plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
-auc_value <- auc(roc_curve)
-print(paste("AUC:", auc_value))
-```
-
-The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing.
-It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.
-
-# Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)
-
-```{r}
-adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
-                      data = cleaned_df, family = binomial)
-summary(adjusted_model)
-
-# Extract Odds Ratios (OR) for the adjusted model
-adj_or_values <- exp(coef(adjusted_model))
-adj_or_ci <- exp(confint(adjusted_model))
-adj_or_results <- data.frame(
-  Term = names(adj_or_values),
-  OR = adj_or_values,
-  CI_lower = adj_or_ci[, 1],
-  CI_upper = adj_or_ci[, 2]
-)
-print("Adjusted Odds Ratios (with 95% CI):")
-print(adj_or_results)
-
-# Tidy the adjusted model output and compute odds ratios and 95% CI
-tidy_adjusted_model <- tidy(adjusted_model, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_adjusted_model)
-```
-
-The adjusted logistic regression model produces unstable estimates.
-The intercept's odds ratio is about 0.07, indicating very low baseline odds.
-The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant.
-The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals.
-These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable.
-This interpretation is derived solely from your output.
-
-# PART 4b: MODEL BUILDING
-
-```{r}
-# Set seed for reproducibility
-rdmseed <- 123
-
-# Reset seed before modeling
-set.seed(rdmseed)
-
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-```
-
-# linear model
-
-```{r}
-# Fit a standard linear regression model using predictors
-lm_model <- linear_reg() %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions from linear model and calculate RMSE
-cleaned_df$pred_lm <- predict(lm_model, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lm)
-
-# # Plot observed vs. predicted for Linear Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
-  geom_point(color = "darkblue", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average.
-This indicates limited accuracy for distinguishing between individuals with and without heart disease.
-
-# LASSO
-
-```{r}
-set.seed(rdmseed)
-
-# Prepare recipe for LASSO with normalization
-rec <- recipe(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df) %>%
-  step_dummy(elevated_BP, elevated_oldpeak) %>%   # Convert categorical predictors to dummies
-  step_normalize(all_numeric(), -all_outcomes())
-
-# Define LASSO model with a fixed penalty
-lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
-  set_engine("glmnet")
-
-# Create a workflow and fit the LASSO model
-wf <- workflow() %>%
-  add_recipe(rec) %>%
-  add_model(lasso_spec)
-
-lasso_fit <- fit(wf, data = cleaned_df)
-cleaned_df$pred_lasso <- predict(lasso_fit, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)
-
-# Plot observed vs. predicted for LASSO Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lasso)) +
-  geom_point(color = "darkviolet", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-```
-
-The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average.
-This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.
-
-# Random Forest
-
-```{r}
-set.seed(rdmseed)
-
-# Create a workflow and fit the Random Forest model
-rf_model <- rand_forest() %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-rf_fit <- rf_model %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions and evaluate model performance
-cleaned_df$pred_rf <- predict(rf_fit, cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_rf)
-
-# Plot observed vs. predicted for RF Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_rf)) +
-  geom_point(color = "darkgreen", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half.
-This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.
-
-# Tuning the LASSO model (without cross-validation)
-
-```{r}
-set.seed(rdmseed)
-
-# Create a grid of penalty values for LASSO
-lasso_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))
-
-# Define LASSO model with tunable penalty
-lasso_spec_tune <- linear_reg(penalty = tune()) %>%
-  set_engine("glmnet")
-
-# Create workflow and assign to lasso_tuning_workflow (new name)
-lasso_tuning_workflow <- workflow() %>%
-  add_model(lasso_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform grid search without CV (apparent resampling)
-lasso_tune <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = apparent(cleaned_df),
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-# Extract tuning results
-lasso_tune_df <- as.data.frame(lasso_tune$.metrics)
-
-# Plot LASSO tuning results
-ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
-  geom_line(size = 1, color = "steelblue") +
-  geom_point(size = 2, color = "firebrick") +
-  scale_x_log10() +
-  labs(
-    x = "Penalty (log scale)",
-    y = "RMSE",
-    title = "LASSO Tuning (No Cross-validation)",
-    subtitle = "Using apparent resampling"
-  ) +
-  theme_minimal()
-```
-
-The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties.
-This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.
-
-# Tuning for our Random Forest model (without cross-validation)
-
-```{r}
-# Create tuning grid for mtry and min_n
-rf_grid <- grid_regular(
-  mtry(range = c(1, 7)),
-  min_n(range = c(1, 21)),
-  levels = 7
-)
-
-# Define random forest model with tunable parameters
-rf_spec_tune <- rand_forest(
-  mtry = tune(),
-  min_n = tune(),
-  trees = 300
-) %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-# Build workflow
-rf_wf_tune <- workflow() %>%
-  add_model(rf_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform RF tuning without CV
-rf_tune <- tune_grid(
-  rf_wf_tune,
-  resamples = apparent(cleaned_df),
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-
-rf_tune_df <- as.data.frame(rf_tune$.metrics)
-
-# Plot RF tuning results
-ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
-  geom_tile() +
-  scale_fill_viridis_c() +
-  labs(
-    title = "Random Forest Tuning (No Cross-validation)",
-    x = "mtry",
-    y = "min_n",
-    fill = "RMSE"
-  ) +
-  theme_minimal()
-```
-
-The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739.
-This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used.
-In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.
-
-# Cross-validation for our LASSO model tuning
-
-```{r}
-set.seed(rdmseed)
-
-# Create 5-fold CV with 5 repeats
-cv_folds <- vfold_cv(cleaned_df, v = 5, repeats = 5)
-
-# Tune LASSO using CV
-lasso_tune_cv <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = cv_folds,
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-autoplot(lasso_tune_cv)
-```
-
-Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases.
-This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.
-
-# Cross-validation tuning for our random forest model
-
-```{r}
-# Tune RF using CV
-rf_tune_cv <- tune_grid(
-  rf_wf_tune,
-  resamples = cv_folds,
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-autoplot(rf_tune_cv)
-
-```
-
-Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry).
-In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values.
-This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.

### 2025-04-17 code/full_code/full_code.qmd
Few minor changes

--- 2025-04-06 products/manuscript/Manuscript.qmd
@@
 ---
-
-## General Background Information
-
-Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease”.
-It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
-This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
-As the leading cause of death in the U.S.
-and globally, CAD claimed 375,500 lives in the U.S.
-in 2021 and affects over 18 million American adults.
-
-## Description of data and data source
-
-This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses.
-Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India.
-This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-This dataset was obtained from mendeley data where it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
-
-## Questions/Hypotheses to be addressed
-
-Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
-
-refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
-
-## Methods
-
-In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present.
-The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).
-
-We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values.
-Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression.
-Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.
-
-To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups.
-Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar.
-Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.
-
-In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods.
-These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE).
-By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.
-
-## **Conclusion**
-
-Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present.
-Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups.
-In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance.
-Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup.
-Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease.
-Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.
-
-## Data aquisition
-
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.
-
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
 
@@
   summarise(
-    # Age statistics
     Age_min = min(age, na.rm = TRUE),
@@
 
-    # Resting Blood Pressure statistics
     RestingBP_min = min(restingBP, na.rm = TRUE),
@@
 
-    # Serum Cholesterol statistics (excluding zeros)
-    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
+    SerumCholestrol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
+    SerumCholestrol_max = max(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
+    SerumCholestrol_mean = mean(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
+    SerumCholestrol_sd = sd(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
 
-    # Maximum Heart Rate statistics
     MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
@@
 
-    # Oldpeak statistics (excluding zeros)
     Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
@@
 
-# Print summary statistics
-print(summary_stats)
+# Reshape into long format, then wide table
+summary_table_formatted <- summary_stats %>%
+  pivot_longer(cols = everything(), names_to = "stat", values_to = "value") %>%
+  separate(stat, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%
+  pivot_wider(names_from = Variable, values_from = value)
 
-summary_df = skimr::skim(summary_stats)
-print(summary_stats)
+# View formatted table
+summary_table_formatted
+
 # save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
+summarytable_file = here("results", "tables", "table1_summary_statistics.rds")
+saveRDS(summary_table_formatted, file = summarytable_file)
 
@@
 
-The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup.
-For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels.
-Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.
-
 # Logistic Regression Analysis
@@
 
-In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero.
-However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own.
-When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance.
-These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.
-
 # Extract Odds Ratios (OR) from the logistic regression model with interaction
@@
 ```
-
-The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87).
-For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22).
-The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82).
-However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.
 
@@
 # save fit results table  
-table_file1 = here("results", "tables", "resulttable4.rds")
-saveRDS(rr_results, file = table_file1)
+table_file3 = here("results", "tables", "table3_relative_risks.rds")
+saveRDS(rr_results, file = table_file3)
 ```
-
-The baseline relative risk is about 0.23 (95% CI: 0.08–0.46).
-For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35).
-The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.
 
@@
 print(paste("AUC:", auc_value))
+
+# Save ROC plot
+roc_fig <- recordPlot()
+saveRDS(roc_fig, file = here("results", "figures", "figure1_roc_logistic_model.rds"))
+
 ```
-
-The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing.
-It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.
 
@@
 ```
-
-The adjusted logistic regression model produces unstable estimates.
-The intercept's odds ratio is about 0.07, indicating very low baseline odds.
-The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant.
-The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals.
-These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable.
-This interpretation is derived solely from your output.
 
@@
 
-The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average.
-This indicates limited accuracy for distinguishing between individuals with and without heart disease.
-
 # LASSO
@@
 
-The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average.
-This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.
-
 # Random Forest
@@
 ```
-
-The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half.
-This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.
 
@@
   theme_minimal()
+
+# Save LASSO tuning plot
+lasso_tune_plot <- last_plot()
+ggsave(filename = here("results", "figures", "figure2_lasso_tuning_plot.png"), plot = lasso_tune_plot)
 ```
-
-The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties.
-This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.
 
@@
   theme_minimal()
+
+# Save Random Forest tuning heatmap
+rf_tune_plot <- last_plot()
+ggsave(filename = here("results", "figures", "figure3_rf_tuning_heatmap.png"), plot = rf_tune_plot)
+
 ```
-
-The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739.
-This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used.
-In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.
 
@@
 
-Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases.
-This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.
-
 # Cross-validation tuning for our random forest model
@@
 ```
-
-Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry).
-In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values.
-This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.

### 2025-04-17 code/processing-code/01_data_cleaning.qmd
Few minor changes

--- 2025-04-17 code/full_code/full_code.qmd
@@
 ---
-title: "Final project"
+title: "Data Processing"
 author: "Murtaza Yaqubi"
-format: html
+format:
+  html:
+    code-fold: true      # Makes code chunks collapsible
+    code-tools: true     # Adds copy/download buttons
+    echo: true           # THIS is what ensures code is shown
 editor: visual
@@
 
+```{r}
 # Load necessary libraries
-
-```{r}
 library(knitr)
@@
 library(naniar)         # For gg_miss_var()
-```
 
-# PART 1: DATA IMPORT AND CLEANING
-
-```{r}
 # Import the dataset
@@
 ```
-
-# PART 2: EXPLORATORY DATA ANALYSIS (EDA)
-
-```{r}
-# Calculate summary statistics for selected variables
-summary_stats <- cardio_df %>% 
-  summarise(
-    Age_min = min(age, na.rm = TRUE),
-    Age_max = max(age, na.rm = TRUE),
-    Age_mean = mean(age, na.rm = TRUE),
-    Age_sd = sd(age, na.rm = TRUE),
-
-    RestingBP_min = min(restingBP, na.rm = TRUE),
-    RestingBP_max = max(restingBP, na.rm = TRUE),
-    RestingBP_mean = mean(restingBP, na.rm = TRUE),
-    RestingBP_sd = sd(restingBP, na.rm = TRUE),
-
-    SerumCholestrol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_max = max(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_mean = mean(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_sd = sd(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-
-    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
-
-    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
-    Oldpeak_max = max(oldpeak, na.rm = TRUE),
-    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
-    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
-  )
-
-# Reshape into long format, then wide table
-summary_table_formatted <- summary_stats %>%
-  pivot_longer(cols = everything(), names_to = "stat", values_to = "value") %>%
-  separate(stat, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") %>%
-  pivot_wider(names_from = Variable, values_from = value)
-
-# View formatted table
-summary_table_formatted
-
-# save to file
-summarytable_file = here("results", "tables", "table1_summary_statistics.rds")
-saveRDS(summary_table_formatted, file = summarytable_file)
-
-# Age distribution
-p1 <- ggplot(cardio_df, aes(x = age)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
-  geom_density(color = "darkblue", linewidth = 1) +
-  labs(title = "Age Distribution", x = "Age", y = "Density") +
-  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
-  theme_bw()
-
-plot(p1)
-figure_file = here("results", "figures","Age distribution.png")
-ggsave(filename = figure_file, plot=p1)
-
-# Serum Cholesterol distribution (excluding zeros)
-p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
-  geom_density(color = "darkgreen", linewidth = 1) +
-  labs(title = "Serum Cholesterol Distribution", 
-       x = "Serum Cholesterol (mg/dL)", y = "Density") +
-  theme_minimal()
-
-plot(p2)
-figure_file = here("results", "figures","Serum Cholestrol.png")
-ggsave(filename = figure_file, plot=p2)
-
-# Resting Blood Pressure distribution (excluding zeros)
-p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
-  geom_density(color = "darkred", linewidth = 1) +
-  labs(title = "Resting Blood Pressure Distribution", 
-       x = "Resting Blood Pressure (mmHG)", y = "Density") +
-  theme_bw()
-
-plot(p3)
-figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
-ggsave(filename = figure_file, plot=p3)
-
-# Boxplot for resting blood pressure by heart disease status
-p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
-  geom_boxplot() +
-  labs(title = "Resting Blood Pressure by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Resting Blood Pressure (mmHg)") +
-  theme_minimal()
-
-plot(p4)
-figure_file = here("results", "figures","restingBP by heart disease status.png")
-ggsave(filename = figure_file, plot=p4)
-
-# Violin plot for oldpeak by heart disease status
-p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
-  geom_violin(trim = FALSE) +
-  labs(title = "Oldpeak Distribution by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Oldpeak") +
-  theme_minimal()
-
-plot(p5)
-figure_file = here("results", "figures","oldpeak by heart disease status.png")
-ggsave(filename = figure_file, plot=p5)
-
-# Create the correlation matrix for selected numeric variables
-numeric_vars <- cardio_df %>% 
-  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
-cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
-
-# Plot the correlation matrix with ggcorrplot
-p6 <- ggcorrplot(cor_matrix, 
-           lab = TRUE,                         
-           lab_size = 3,                       
-           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
-           title = "Correlation Matrix", 
-           ggtheme = theme_minimal())
-
-plot(p6)
-figure_file = here("results", "figures","correlation matrix.png")
-ggsave(filename = figure_file, plot=p6)     
-```
-
-# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
-
-```{r}
-# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
-cleaned_df <- cardio_df %>%
-  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
-  mutate(
-    # Create a binary variable for blood pressure:
-    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
-    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),
-
-    # Create a binary indicator for ST depression:
-    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
-    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
-  ) %>% 
-  # Select the variables we want to keep
-  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
-         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)
-
-# Print a sample of the cleaned dataset
-head(cleaned_df)
-
-# Save the processed data
-saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")
-
-# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
-summary_table <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-print(summary_table)
-
-summary_df = skimr::skim(summary_table)
-print(summary_table)
-# save to file
-summarytable_file = here("results", "tables", "summarytable2.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Interaction Visualization: Heatmap
-interaction_counts <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-
-p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
-  geom_tile(color = "white", size = 0.5) +
-  geom_text(aes(label = count), color = "black", size = 5) +
-  facet_wrap(~ target, ncol = 2) +
-  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
-  labs(title = "Interaction between Blood Pressure and ST Depression",
-       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
-       x = "Blood Pressure Status",
-       y = "ST Depression Status",
-       fill = "Count") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5)
-  )
-
-plot(p7)
-figure_file = here("results", "figures","interaction heatmap.png")
-ggsave(filename = figure_file, plot=p7)
-
-# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
-p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
-  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
-  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
-       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
-       x = "Blood Pressure Status",
-       y = "Count",
-       fill = "Heart Disease") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5),
-    legend.position = "bottom"
-  )
-
-plot(p8)
-figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
-ggsave(filename = figure_file, plot=p8)
-```
-
-# PART 4a: INFERENTIAL STATISTICAL ANALYSIS
-
-# Chi-Square Tests
-
-```{r}
-# Create a contingency table for blood pressure status vs. heart disease
-table_bp_target <- table(cleaned_df$elevated_BP, cleaned_df$target)
-print(table_bp_target)
-chisq_test_bp <- chisq.test(table_bp_target)
-print(chisq_test_bp)
-
-# Create a contingency table for ST depression vs. heart disease
-table_oldpeak_target <- table(cleaned_df$elevated_oldpeak, cleaned_df$target)
-print(table_oldpeak_target)
-chisq_test_oldpeak <- chisq.test(table_oldpeak_target)
-print(chisq_test_oldpeak)
-```
-
-# Logistic Regression Analysis
-
-```{r}
-# Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-
-# Fit a logistic regression model with elevated_BP and elevated_oldpeak as predictors
-model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-             data = cleaned_df, family = binomial)
-summary(model)
-
-# save fit results table  
-table_file1 = here("results", "tables", "resulttable1.rds")
-saveRDS(model, file = table_file1)
-
-# Fit a logistic regression model including the interaction term
-model_interaction <- glm(target_binary ~ elevated_BP * elevated_oldpeak, 
-                         data = cleaned_df, family = binomial)
-summary(model_interaction)
-
-# save fit results table  
-table_file1 = here("results", "tables", "resulttable2.rds")
-saveRDS(model_interaction, file = table_file1)
-```
-
-# Extract Odds Ratios (OR) from the logistic regression model with interaction
-
-```{r}
-or_values <- exp(coef(model_interaction))
-or_ci <- exp(confint(model_interaction))
-or_results <- data.frame(
-  Term = names(or_values),
-  OR = or_values,
-  CI_lower = or_ci[, 1],
-  CI_upper = or_ci[, 2]
-)
-print("Odds Ratios (with 95% CI):")
-print(or_results)
-
-# save fit results table  
-table_file2 = here("results", "tables", "table2_odds_ratios.rds")
-saveRDS(or_results, file = table_file2)
-
-# Tidy the model output and compute odds ratios and 95% CI
-tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_model)
-```
-
-# Fit a log-binomial model to estimate Relative Risks (RR)
-
-```{r}
-rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-                data = cleaned_df, family = binomial(link = "log"))
-rr_values <- exp(coef(rr_model))
-rr_ci <- exp(confint(rr_model))
-rr_results <- data.frame(
-  Term = names(rr_values),
-  RR = rr_values,
-  CI_lower = rr_ci[, 1],
-  CI_upper = rr_ci[, 2]
-)
-print("Relative Risks (with 95% CI):")
-print(rr_results)
-
-# save fit results table  
-table_file3 = here("results", "tables", "table3_relative_risks.rds")
-saveRDS(rr_results, file = table_file3)
-```
-
-# ROC Curve and AUC for the logistic regression model with interaction
-
-```{r}
-roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
-plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
-auc_value <- auc(roc_curve)
-print(paste("AUC:", auc_value))
-
-# Save ROC plot
-roc_fig <- recordPlot()
-saveRDS(roc_fig, file = here("results", "figures", "figure1_roc_logistic_model.rds"))
-
-```
-
-# Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)
-
-```{r}
-adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
-                      data = cleaned_df, family = binomial)
-summary(adjusted_model)
-
-# Extract Odds Ratios (OR) for the adjusted model
-adj_or_values <- exp(coef(adjusted_model))
-adj_or_ci <- exp(confint(adjusted_model))
-adj_or_results <- data.frame(
-  Term = names(adj_or_values),
-  OR = adj_or_values,
-  CI_lower = adj_or_ci[, 1],
-  CI_upper = adj_or_ci[, 2]
-)
-print("Adjusted Odds Ratios (with 95% CI):")
-print(adj_or_results)
-
-# save fit results table  
-table_file4 = here("results", "tables", "table4_adjusted_odds_ratios.rds")
-saveRDS(adj_or_results, file = table_file4)
-
-# Tidy the adjusted model output and compute odds ratios and 95% CI
-tidy_adjusted_model <- tidy(adjusted_model, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_adjusted_model)
-```
-
-# PART 4b: MODEL BUILDING
-
-```{r}
-# Set seed for reproducibility
-rdmseed <- 123
-
-# Reset seed before modeling
-set.seed(rdmseed)
-
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-```
-
-# linear model
-
-```{r}
-# Fit a standard linear regression model using predictors
-lm_model <- linear_reg() %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions from linear model and calculate RMSE
-cleaned_df$pred_lm <- predict(lm_model, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lm)
-
-# # Plot observed vs. predicted for Linear Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
-  geom_point(color = "darkblue", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-# LASSO
-
-```{r}
-set.seed(rdmseed)
-
-# Prepare recipe for LASSO with normalization
-rec <- recipe(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df) %>%
-  step_dummy(elevated_BP, elevated_oldpeak) %>%   # Convert categorical predictors to dummies
-  step_normalize(all_numeric(), -all_outcomes())
-
-# Define LASSO model with a fixed penalty
-lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
-  set_engine("glmnet")
-
-# Create a workflow and fit the LASSO model
-wf <- workflow() %>%
-  add_recipe(rec) %>%
-  add_model(lasso_spec)
-
-lasso_fit <- fit(wf, data = cleaned_df)
-cleaned_df$pred_lasso <- predict(lasso_fit, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)
-
-# Plot observed vs. predicted for LASSO Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lasso)) +
-  geom_point(color = "darkviolet", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-```
-
-# Random Forest
-
-```{r}
-set.seed(rdmseed)
-
-# Create a workflow and fit the Random Forest model
-rf_model <- rand_forest() %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-rf_fit <- rf_model %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions and evaluate model performance
-cleaned_df$pred_rf <- predict(rf_fit, cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_rf)
-
-# Save RMSEs from all models to a table
-rmse_df <- tibble(
-  Model = c("Linear", "LASSO", "Random Forest"),
-  RMSE = c(
-    rmse(cleaned_df, truth = target_binary, estimate = pred_lm)$.estimate,
-    rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)$.estimate,
-    rmse(cleaned_df, truth = target_binary, estimate = pred_rf)$.estimate
-  )
-)
-saveRDS(rmse_df, file = here("results", "tables", "table5_model_rmse_comparison.rds"))
-
-# Plot observed vs. predicted for RF Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_rf)) +
-  geom_point(color = "darkgreen", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-# Tuning the LASSO model (without cross-validation)
-
-```{r}
-set.seed(rdmseed)
-
-# Create a grid of penalty values for LASSO
-lasso_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))
-
-# Define LASSO model with tunable penalty
-lasso_spec_tune <- linear_reg(penalty = tune()) %>%
-  set_engine("glmnet")
-
-# Create workflow and assign to lasso_tuning_workflow (new name)
-lasso_tuning_workflow <- workflow() %>%
-  add_model(lasso_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform grid search without CV (apparent resampling)
-lasso_tune <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = apparent(cleaned_df),
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-# Extract tuning results
-lasso_tune_df <- as.data.frame(lasso_tune$.metrics)
-
-# Identify best LASSO penalty
-best_lasso <- lasso_tune_df %>% filter(.estimate == min(.estimate))
-print("Best LASSO Penalty (No CV):")
-print(best_lasso)
-
-# Plot LASSO tuning results
-ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
-  geom_line(size = 1, color = "steelblue") +
-  geom_point(size = 2, color = "firebrick") +
-  scale_x_log10() +
-  labs(
-    x = "Penalty (log scale)",
-    y = "RMSE",
-    title = "LASSO Tuning (No Cross-validation)",
-    subtitle = "Using apparent resampling"
-  ) +
-  theme_minimal()
-
-# Save LASSO tuning plot
-lasso_tune_plot <- last_plot()
-ggsave(filename = here("results", "figures", "figure2_lasso_tuning_plot.png"), plot = lasso_tune_plot)
-```
-
-# Tuning for our Random Forest model (without cross-validation)
-
-```{r}
-# Create tuning grid for mtry and min_n
-rf_grid <- grid_regular(
-  mtry(range = c(1, 7)),
-  min_n(range = c(1, 21)),
-  levels = 7
-)
-
-# Define random forest model with tunable parameters
-rf_spec_tune <- rand_forest(
-  mtry = tune(),
-  min_n = tune(),
-  trees = 300
-) %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-# Build workflow
-rf_wf_tune <- workflow() %>%
-  add_model(rf_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform RF tuning without CV
-rf_tune <- tune_grid(
-  rf_wf_tune,
-  resamples = apparent(cleaned_df),
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-
-rf_tune_df <- as.data.frame(rf_tune$.metrics)
-
-# Identify best RF tuning parameters
-best_rf <- rf_tune_df %>% filter(.estimate == min(.estimate))
-print("Best Random Forest Parameters (No CV):")
-print(best_rf)
-
-# Plot RF tuning results
-ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
-  geom_tile() +
-  scale_fill_viridis_c() +
-  labs(
-    title = "Random Forest Tuning (No Cross-validation)",
-    x = "mtry",
-    y = "min_n",
-    fill = "RMSE"
-  ) +
-  theme_minimal()
-
-# Save Random Forest tuning heatmap
-rf_tune_plot <- last_plot()
-ggsave(filename = here("results", "figures", "figure3_rf_tuning_heatmap.png"), plot = rf_tune_plot)
-
-```
-
-# Cross-validation for our LASSO model tuning
-
-```{r}
-set.seed(rdmseed)
-
-# Create 5-fold CV with 5 repeats
-cv_folds <- vfold_cv(cleaned_df, v = 5, repeats = 5)
-
-# Tune LASSO using CV
-lasso_tune_cv <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = cv_folds,
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-autoplot(lasso_tune_cv)
-```
-
-# Cross-validation tuning for our random forest model
-
-```{r}
-# Tune RF using CV
-rf_tune_cv <- tune_grid(
-  rf_wf_tune,
-  resamples = cv_folds,
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-autoplot(rf_tune_cv)
-
-```

### 2025-04-17 products/manuscript/Manuscript.qmd
Few minor changes

--- 2025-04-05 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Final project"
+title: "Early Cardiovascular Disease Risk Among Younger Individuals with Normal Cholesterol: The Role of Elevated Blood Pressure and ST Depression"
 author: "Murtaza Yaqubi"
-format: html
-editor: visual
+format: docx
 ---
 
-## General Background Information
+# Abstract
 
-Coronary artery disease (CAD), also known as coronary heart disease or ischemic heart disease.
-It’s also what most people mean when they use the general term “heart disease”.
-It is a common and serious condition where plaque buildup in the coronary arteries restricts blood flow to the heart.
-This gradual narrowing, caused by atherosclerosis, often develops over years without symptoms until a heart attack occurs, making CAD a "silent killer." Symptoms, when present, include chest pain and shortness of breath.
-As the leading cause of death in the U.S.
-and globally, CAD claimed 375,500 lives in the U.S.
-in 2021 and affects over 18 million American adults.
+This study examines whether younger individuals with clinically normal cholesterol levels are at an increased risk of cardiovascular disease (CVD) when they also present with elevated blood pressure or ST depression.
+Using a dataset of anonymized patient records from a multispecialty hospital in India, we assess associations between these secondary risk factors and heart disease through descriptive statistics, inferential tests, and predictive modeling.
+The findings, based on chi-square tests, logistic regression, and machine learning approaches, suggest that these additional risk factors do not significantly alter the risk of CVD in this population.
 
-## Description of data and data source
+# Introduction
 
-This dataset contains anonymized patient records related to cardiovascular disease (CVD) risk factors and diagnoses.
-Each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
+Coronary artery disease (CAD), often referred to as heart disease, is a leading cause of death worldwide.
+It is characterized by the gradual narrowing of the coronary arteries due to plaque buildup, which can result in myocardial infarction, often without prior symptoms.
+This "silent killer" claims millions of lives each year, highlighting the importance of early detection and prevention.
 
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India.
-This dataset consists of 1000 subjects so each row represents a patient, with features like age, gender, blood pressure, cholesterol levels, and clinical indicators (e.g., ST depression, chest pain type).
-This dataset was obtained from mendeley data where it was first published in April of 2021.
+This study utilizes a dataset comprising anonymized patient records related to cardiovascular risk factors and diagnoses.
+Each record includes variables such as age, gender, blood pressure, cholesterol levels, and various clinical indicators like ST depression and chest pain type.
+The data, collected from a multispecialty hospital in India and published on Mendeley Data in April 2021, offers an opportunity to examine risk beyond conventional markers.
 
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
+The central research question is whether individuals under 50 with normal cholesterol levels (defined as <200 mg/dL) remain at risk for CVD if they also present with elevated blood pressure or ST depression.
+The goal is to challenge traditional frameworks and explore whether secondary risk factors contribute meaningfully to heart disease risk in this population.
 
-## Questions/Hypotheses to be addressed
+# Methods
 
-Do individuals younger than 50 with normal cholesterol levels (within "healthy" range) still face CVD risk if they have elevated blood pressure or ST depression (oldpeak)?
+## Study Population and Objective
 
-refernece: 1- https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings 1- https://www.mayoclinic.org/diseases-conditions/prehypertension/symptoms-causes/syc-20376703 2- https://my.clevelandclinic.org/health/articles/11920-cholesterol-numbers-what-do-they-mean 2- https://www.hopkinsmedicine.org/health/treatment-tests-and-therapies/lipid-panel 3- https://pmc.ncbi.nlm.nih.gov/articles/PMC6376358/ 3- https://www.ahajournals.org/doi/pdf/10.1161/01.STR.25.9.1820
+The study focused on individuals under 50 years of age with clinically normal cholesterol levels (<200 mg/dL).
+These thresholds were chosen based on clinical relevance.
+The dataset includes anonymized demographic, diagnostic, and physiological data on more than 1,000 patients.
 
-## Methods
+The primary outcome, target, indicates presence (1) or absence (0) of heart disease.
+Key predictors included age, resting blood pressure, serum cholesterol, ST depression (oldpeak), gender, and fasting blood sugar.
 
-In this study, we will investigate whether younger individuals with clinically normal serum cholesterol levels are at an increased risk of cardiovascular disease (CVD) when additional risk factors—specifically elevated blood pressure and ST depression—are present.
-The analysis will be based on a cardiovascular disease dataset where the primary outcome, “target,” indicates the absence or presence of heart disease, and key predictors include age, serum cholesterol, resting blood pressure, and ST depression (oldpeak).
+## Risk Variable Derivation
 
-We will begin by restricting our study population to patients under 50 years of age with serum cholesterol levels below 200 mg/dL, a threshold that reflects clinically normal values.
-Within this subset, patients will be categorized according to secondary risk factors: those with a resting blood pressure between 120 and 129 mmHg will be designated as having elevated blood pressure, and those with an oldpeak value greater than 1 will be classified as exhibiting elevated ST depression.
-Descriptive analyses will then be performed to compare CVD prevalence across these risk groups, providing initial insights into potential associations between these non-traditional risk factors and CVD in a younger population.
+Two dichotomous predictors were engineered:
 
-To validate these observations, chi-square tests will be applied to assess the statistical significance of differences in CVD prevalence among the defined groups.
-Furthermore, logistic regression models will be used to quantify the independent and combined effects of elevated blood pressure and ST depression on CVD risk, with additional adjustments made for potential confounders such as gender and fasting blood sugar.
-Model performance will be evaluated using outputs such as odds ratios with 95% confidence intervals, relative risks, and receiver operating characteristic (ROC) curves with corresponding area under the curve (AUC) values.
+- **Elevated Blood Pressure**: Defined as resting systolic BP between 120–129 mmHg
+- **Elevated ST Depression**: Defined as oldpeak > 1.0
 
-In addition to these inferential techniques, we will develop predictive models employing linear regression, LASSO regression, and random forest methods.
-These models will be optimized through grid search and cross-validation procedures, and their predictive accuracy will be measured using the root mean squared error (RMSE).
-By integrating these methodologies, readers can expect a comprehensive evaluation of CVD risk among younger individuals, ultimately challenging conventional risk-assessment frameworks and supporting more nuanced screening approaches for early detection of cardiovascular disease.
+All categorical variables were converted to factors for modeling.
 
-## **Conclusion**
+## Statistical Procedures
 
-Based solely on the outputs obtained from our analysis, there is little evidence to support the hypothesis that younger individuals with normal cholesterol levels face an increased risk of heart disease when secondary risk factors, namely elevated blood pressure and ST depression are present.
-Chi-square tests revealed no significant differences in the prevalence of cardiovascular disease across the risk groups.
-In the logistic regression models, the estimated odds ratio for elevated blood pressure was approximately 2.29 (95% CI: 0.55–12.00) and for ST depression about 1.83 (95% CI: 0.07–27.22), though neither effect reached statistical significance.
-Similarly, the relative risk estimates were around 1.66 (95% CI: 0.73–5.19) for elevated blood pressure and 1.21 (95% CI: 0.61–2.35) for ST depression, further suggesting that these factors do not significantly impact the risk of heart disease in this subgroup.
-Additionally, predictive models including linear regression, LASSO, and random forest, produced modest performance (with RMSE values near 0.47–0.48 and an AUC of about 0.60), reinforcing the conclusion that these secondary risk factors do not markedly improve prediction of cardiovascular disease.
-Overall, the findings do not provide strong evidence that the presence of elevated blood pressure or ST depression significantly increases CVD risk among younger individuals with normal cholesterol levels.
+Descriptive statistics summarized clinical variables, while histograms, boxplots, and correlation matrices described distributions and associations.
+A filtered dataset containing only younger individuals with normal cholesterol was used in all analyses.
 
-## Data aquisition
+Chi-square tests evaluated whether elevated blood pressure or ST depression was associated with higher CVD prevalence.
+Logistic regression models with and without interaction terms were fit, along with a log-binomial model for estimating relative risk.
+Models were further adjusted for gender and fasting blood sugar.
 
-This heart disease dataset is acquired from one o f the multispecialty hospitals in India and it was first published in April of 2021.
+## Predictive Modeling
 
-This dataset can be accessed at <https://data.mendeley.com/datasets/dzz48mvjht/1>
+Three predictive models—linear regression, LASSO regression, and random forest—were trained using the filtered dataset.
+Model performance was assessed using RMSE.
+Tuning procedures for LASSO (penalty parameter) and random forest (mtry and min_n) were performed via grid search and cross-validation.
+ROC analysis and AUC were used to evaluate classification performance.
 
-# Load necessary libraries
+# Results
+
+## Descriptive Statistics
+
+The dataset was explored using summary statistics and visualizations (see Table 1, `table1_summary_statistics.rds`).
+Age ranged widely, and resting blood pressure and cholesterol also showed broad variability.
+Correlation analysis revealed moderate relationships among several predictors.
+
 ```{r}
-library(knitr)
-library(readr)
-library(dplyr)
-library(ggplot2)
-library(ggcorrplot)
-library(vcd)
-library(caret)        # For data partitioning and model training
-library(randomForest) # For Random Forest modeling
-library(tidyverse)
-library(tidymodels)
-library(broom)
-library(here)           # For file paths
-library(pROC)
-library(naniar)         # For gg_miss_var()
+summary_df <- readRDS("results/tables/table1_summary_statistics.rds")
+knitr::kable(summary_df, caption = "Table 1. Summary statistics of the study population")
 ```
 
-# PART 1: DATA IMPORT AND CLEANING
+## Inferential Findings
+
+Chi-square tests indicated no statistically significant association between elevated blood pressure or ST depression and heart disease (p-values = 0.31 and 0.55, respectively).
+Summary tables for group frequencies are provided in the appendix.
+
+Logistic regression models showed no statistically significant odds ratios for either risk factor.
+In the interaction model, elevated blood pressure had an OR of 2.29 (95% CI: 0.55–12.00), and ST depression an OR of 1.83 (95% CI: 0.07–27.22) (Table 2, `table2_odds_ratios.rds`).
+These results were consistent with the non-interaction model (Table 3), and interaction terms were also nonsignificant (Table 4).
+
+Relative risk estimates from the log-binomial model were similarly non-significant, with RR = 1.66 for elevated blood pressure and RR = 1.21 for elevated ST depression (Table 3, `table3_relative_risks.rds`).
+
 ```{r}
-# Import the dataset
-cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
-
-# Basic exploration of the dataset
-head(cardio_df)           # View the first few rows
-colnames(cardio_df)       # Check column names
-dim(cardio_df)            # Check dimensions (should be 1025 rows x 14 columns)
-summary(cardio_df)        # Get a summary of the data
-glimpse(cardio_df)        # Glimpse at the data structure
-
-# Visualize missing values for each variable
-gg_miss_var(cardio_df)    # Display number of NAs
-
-# Transform variables into suitable types and add descriptive labels
-cardio_df <- cardio_df %>%
-  mutate(
-    gender = factor(gender, levels = c(0, 1), labels = c("Female", "Male")), 
-    chestpain = factor(chestpain, levels = c(0, 1, 2, 3),
-                       labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic")),
-    fastingbloodsugar = factor(fastingbloodsugar, levels = c(0, 1), labels = c("False", "True")),
-    restingrelectro = factor(restingrelectro, levels = c(0, 1, 2),
-                             labels = c("Normal", "ST-T Abnormality", "Left Ventricular Hypertrophy")),
-    exerciseangia = factor(exerciseangia, levels = c(0, 1), labels = c("No", "Yes")),
-    slope = factor(slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping")),
-    target = factor(target, levels = c(0, 1), labels = c("No Heart Disease", "Heart Disease"))
-  )
-
-# Print the transformed dataset for verification
-print(cardio_df)
+or_results <- readRDS("results/tables/table2_odds_ratios.rds")
+knitr::kable(or_results, caption = "Table 2. Odds ratios from logistic regression with interaction terms")
 ```
 
-# PART 2: EXPLORATORY DATA ANALYSIS (EDA)
+```{r}
+rr_results <- readRDS("results/tables/table3_relative_risks.rds")
+knitr::kable(rr_results, caption = "Table 3. Relative risk estimates from log-binomial model")
+```
+
+## Adjusted Model Results
+
+After adjusting for gender and fasting blood sugar, model coefficients became unstable, likely due to sparse data and collinearity.
+Although gender showed elevated risk (adjusted OR = 4.69), other estimates were unreliable (Table 4, `table4_adjusted_odds_ratios.rds`).
 
 ```{r}
-# Calculate summary statistics for selected variables
-summary_stats <- cardio_df %>% 
-  summarise(
-    # Age statistics
-    Age_min = min(age, na.rm = TRUE),
-    Age_max = max(age, na.rm = TRUE),
-    Age_mean = mean(age, na.rm = TRUE),
-    Age_sd = sd(age, na.rm = TRUE),
-
-    # Resting Blood Pressure statistics
-    RestingBP_min = min(restingBP, na.rm = TRUE),
-    RestingBP_max = max(restingBP, na.rm = TRUE),
-    RestingBP_mean = mean(restingBP, na.rm = TRUE),
-    RestingBP_sd = sd(restingBP, na.rm = TRUE),
-
-    # Serum Cholesterol statistics (excluding zeros)
-    SerumCholesterol_min = min(serumcholestrol[serumcholestrol > 0], na.rm = TRUE),
-    SerumCholestrol_max = max(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_mean = mean(serumcholestrol, na.rm = TRUE),
-    SerumCholestrol_sd = sd(serumcholestrol, na.rm = TRUE),
-
-    # Maximum Heart Rate statistics
-    MaxHeartRate_min = min(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_max = max(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_mean = mean(maxheartrate, na.rm = TRUE),
-    MaxHeartRate_sd = sd(maxheartrate, na.rm = TRUE),
-
-    # Oldpeak statistics (excluding zeros)
-    Oldpeak_min = min(oldpeak[oldpeak > 0], na.rm = TRUE),
-    Oldpeak_max = max(oldpeak, na.rm = TRUE),
-    Oldpeak_mean = mean(oldpeak, na.rm = TRUE),
-    Oldpeak_sd = sd(oldpeak, na.rm = TRUE)
-  )
-
-# Print summary statistics
-print(summary_stats)
-
-summary_df = skimr::skim(summary_stats)
-print(summary_stats)
-# save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Age distribution
-p1 <- ggplot(cardio_df, aes(x = age)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
-  geom_density(color = "darkblue", linewidth = 1) +
-  labs(title = "Age Distribution", x = "Age", y = "Density") +
-  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
-  theme_bw()
-
-plot(p1)
-figure_file = here("results", "figures","Age distribution.png")
-ggsave(filename = figure_file, plot=p1)
-
-# Serum Cholesterol distribution (excluding zeros)
-p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
-  geom_density(color = "darkgreen", linewidth = 1) +
-  labs(title = "Serum Cholesterol Distribution", 
-       x = "Serum Cholesterol (mg/dL)", y = "Density") +
-  theme_minimal()
-
-plot(p2)
-figure_file = here("results", "figures","Serum Cholestrol.png")
-ggsave(filename = figure_file, plot=p2)
-
-# Resting Blood Pressure distribution (excluding zeros)
-p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
-  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
-  geom_density(color = "darkred", linewidth = 1) +
-  labs(title = "Resting Blood Pressure Distribution", 
-       x = "Resting Blood Pressure (mmHG)", y = "Density") +
-  theme_bw()
-
-plot(p3)
-figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
-ggsave(filename = figure_file, plot=p3)
-
-# Boxplot for resting blood pressure by heart disease status
-p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
-  geom_boxplot() +
-  labs(title = "Resting Blood Pressure by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Resting Blood Pressure (mmHg)") +
-  theme_minimal()
-
-plot(p4)
-figure_file = here("results", "figures","restingBP by heart disease status.png")
-ggsave(filename = figure_file, plot=p4)
-
-# Violin plot for oldpeak by heart disease status
-p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
-  geom_violin(trim = FALSE) +
-  labs(title = "Oldpeak Distribution by Heart Disease Status",
-       x = "Heart Disease Status",
-       y = "Oldpeak") +
-  theme_minimal()
-
-plot(p5)
-figure_file = here("results", "figures","oldpeak by heart disease status.png")
-ggsave(filename = figure_file, plot=p5)
-
-# Create the correlation matrix for selected numeric variables
-numeric_vars <- cardio_df %>% 
-  select(age, restingBP, serumcholestrol, maxheartrate, oldpeak)
-cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")
-
-# Plot the correlation matrix with ggcorrplot
-p6 <- ggcorrplot(cor_matrix, 
-           lab = TRUE,                         
-           lab_size = 3,                       
-           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
-           title = "Correlation Matrix", 
-           ggtheme = theme_minimal())
-
-plot(p6)
-figure_file = here("results", "figures","correlation matrix.png")
-ggsave(filename = figure_file, plot=p6)     
+adj_or_results <- readRDS("results/tables/table4_adjusted_odds_ratios.rds")
+knitr::kable(adj_or_results, caption = "Table 4. Adjusted odds ratios including gender and fasting blood sugar")
 ```
 
-# PART 3: STATISTICAL ANALYSIS: DATA CLEANING, SUMMARIZATION, AND VISUALIZATION
+## Model Discrimination
+
+ROC curve analysis showed poor model performance (AUC ≈ 0.60), indicating limited ability to discriminate between individuals with and without CVD (Figure 1, `figure1_roc_logistic_model.rds`).
+
+```{r, fig.cap="Figure 1. ROC curve for logistic regression model"}
+roc_fig <- readRDS("results/figures/figure1_roc_logistic_model.rds")
+replayPlot(roc_fig)
+```
+
+## Predictive Modeling and Tuning
+
+The linear, LASSO, and random forest models produced RMSEs of approximately 0.47, 0.48, and 0.47, respectively (Table 5, `table5_model_rmse_comparison.rds`).
+Tuning for LASSO showed little improvement across penalty values (Figure 2), and random forest tuning identified an optimal configuration of mtry = 2, min_n = 14 with similar performance (Figure 3).
 
 ```{r}
-# Filter the data for individuals under 50 with normal cholesterol (<200 mg/dL) and create risk indicator variables for elevated blood pressure and ST depression.
-cleaned_df <- cardio_df %>%
-  filter(age < 50, serumcholestrol > 0, serumcholestrol < 200) %>%  
-  mutate(
-    # Create a binary variable for blood pressure:
-    # "Elevated" if restingBP is between 120 and 129; otherwise, "Otherwise"
-    elevated_BP = if_else(between(restingBP, 120, 129), "Elevated", "Otherwise"),
-
-    # Create a binary indicator for ST depression:
-    # "Elevated" if oldpeak is greater than 1; otherwise, "Normal"
-    elevated_oldpeak = if_else(oldpeak > 1, "Elevated", "Normal")
-  ) %>% 
-  # Select the variables we want to keep
-  select(age, gender, chestpain, restingBP, serumcholestrol, fastingbloodsugar, 
-         maxheartrate, oldpeak, target, elevated_BP, elevated_oldpeak)
-
-# Print a sample of the cleaned dataset
-head(cleaned_df)
-
-# Save the processed data
-saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")
-
-# Create a summary table by grouping the cleaned data by blood pressure status, ST depression status, and heart disease (target)
-summary_table <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-print(summary_table)
-
-summary_df = skimr::skim(summary_table)
-print(summary_table)
-# save to file
-summarytable_file = here("results", "tables", "summarytable2.rds")
-saveRDS(summary_df, file = summarytable_file)
-
-# Interaction Visualization: Heatmap
-interaction_counts <- cleaned_df %>%
-  group_by(elevated_BP, elevated_oldpeak, target) %>%
-  summarise(count = n(), .groups = "drop")
-
-p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
-  geom_tile(color = "white", size = 0.5) +
-  geom_text(aes(label = count), color = "black", size = 5) +
-  facet_wrap(~ target, ncol = 2) +
-  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
-  labs(title = "Interaction between Blood Pressure and ST Depression",
-       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
-       x = "Blood Pressure Status",
-       y = "ST Depression Status",
-       fill = "Count") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5)
-  )
-
-plot(p7)
-figure_file = here("results", "figures","interaction heatmap.png")
-ggsave(filename = figure_file, plot=p7)
-
-# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
-p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
-  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
-  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
-       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
-       x = "Blood Pressure Status",
-       y = "Count",
-       fill = "Heart Disease") +
-  theme_minimal(base_size = 14) +
-  theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    strip.text = element_text(face = "bold"),
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    plot.subtitle = element_text(hjust = 0.5),
-    legend.position = "bottom"
-  )
-
-plot(p8)
-figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
-ggsave(filename = figure_file, plot=p8)
+rmse_df <- readRDS("results/tables/table5_model_rmse_comparison.rds")
+knitr::kable(rmse_df, caption = "Table 5. RMSE values for linear, LASSO, and random forest models")
 ```
 
-# PART 4a: INFERENTIAL STATISTICAL ANALYSIS
-
-# Chi-Square Tests
-
-```{r}
-# Create a contingency table for blood pressure status vs. heart disease
-table_bp_target <- table(cleaned_df$elevated_BP, cleaned_df$target)
-print(table_bp_target)
-chisq_test_bp <- chisq.test(table_bp_target)
-print(chisq_test_bp)
-
-# Create a contingency table for ST depression vs. heart disease
-table_oldpeak_target <- table(cleaned_df$elevated_oldpeak, cleaned_df$target)
-print(table_oldpeak_target)
-chisq_test_oldpeak <- chisq.test(table_oldpeak_target)
-print(chisq_test_oldpeak)
+```{r, fig.cap="Figure 2. LASSO tuning curve showing RMSE stability across penalties"}
+knitr::include_graphics("results/figures/figure2_lasso_tuning_plot.png")
 ```
 
-The chi-square tests show that there is no statistically significant association between the risk factors and heart disease in this subgroup.
-For blood pressure, the test (χ² = 1.012, p = 0.3144) indicates no significant difference in heart disease rates between those with elevated blood pressure and those with otherwise normal levels.
-Similarly, for ST depression, the test (χ² = 0.35953, p = 0.5488) suggests that heart disease prevalence does not differ significantly between patients with elevated versus normal ST depression.
-
-# Logistic Regression Analysis
-
-```{r}
-# Convert target to a binary outcome for regression (1 = Heart Disease, 0 = No Heart Disease)
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-
-# Fit a logistic regression model with elevated_BP and elevated_oldpeak as predictors
-model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-             data = cleaned_df, family = binomial)
-summary(model)
-
-# Fit a logistic regression model including the interaction term
-model_interaction <- glm(target_binary ~ elevated_BP * elevated_oldpeak, 
-                         data = cleaned_df, family = binomial)
-summary(model_interaction)
+```{r, fig.cap="Figure 3. Random forest tuning heatmap showing optimal mtry and min_n"}
+knitr::include_graphics("results/figures/figure3_rf_tuning_heatmap.png")
 ```
 
-In the logistic regression model that includes only elevated blood pressure and ST depression as predictors, the intercept is significant (p ≈ 0.034), which tells us that the baseline log odds of heart disease (when both predictors are at their reference levels) differ from zero.
-However, neither elevated blood pressure (p ≈ 0.259) nor elevated ST depression (p ≈ 0.563) are statistically significant predictors of heart disease on their own.
-When we include an interaction term between elevated blood pressure and ST depression, the results remain similar: the intercept is still significant (p ≈ 0.046), but the main effects of elevated blood pressure (p ≈ 0.279) and elevated ST depression (p ≈ 0.662), as well as their interaction (p ≈ 0.824), do not reach significance.
-These findings suggest that, in this analysis, neither elevated blood pressure nor elevated ST depression—nor their combination—provides significant predictive information for the presence of heart disease among these individuals.
+# Discussion
 
-# Extract Odds Ratios (OR) from the logistic regression model with interaction
+## Summary and Interpretation
 
-```{r}
-or_values <- exp(coef(model_interaction))
-or_ci <- exp(confint(model_interaction))
-or_results <- data.frame(
-  Term = names(or_values),
-  OR = or_values,
-  CI_lower = or_ci[, 1],
-  CI_upper = or_ci[, 2]
-)
-print("Odds Ratios (with 95% CI):")
-print(or_results)
+The analyses presented here do not support the hypothesis that younger individuals with normal cholesterol face elevated CVD risk due to the presence of high-normal blood pressure or ST depression.
+Chi-square tests, logistic regression models, and machine learning approaches consistently failed to show statistically or clinically significant results.
+RMSEs for all predictive models remained similar, and AUC metrics hovered around 0.60, further suggesting limited discriminatory power.
 
-# Tidy the model output and compute odds ratios and 95% CI
-tidy_model <- tidy(model_interaction, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_model)
-```
+## Strengths and Limitations
 
-The odds ratios indicate that when all predictors are at their reference levels, the baseline odds of heart disease are estimated at about 0.27 (95% CI: 0.06–0.87).
-For patients with elevated blood pressure (compared to the reference), the odds of heart disease are roughly 2.29 times higher (95% CI: 0.55–12.00), while those with normal ST depression have an odds ratio of about 1.83 (95% CI: 0.07–27.22).
-The interaction term between elevated blood pressure and normal ST depression shows an odds ratio of approximately 0.71 (95% CI: 0.04–21.82).
-However, the very wide confidence intervals across these estimates suggest a high degree of uncertainty, and none of these effects appear to be statistically significant.
+A key strength of this study is its comprehensive approach, incorporating multiple analytical methods across inferential and predictive domains.
+Grid search and cross-validation further enhanced model assessment.
 
-# Fit a log-binomial model to estimate Relative Risks (RR)
+However, limitations include small subgroup sizes and potential instability of parameter estimates due to sparse data.
+Adjusted models in particular exhibited wide confidence intervals and signs of overfitting.
+These findings emphasize the need for larger, prospective studies to validate the observed trends.
 
-```{r}
-rr_model <- glm(target_binary ~ elevated_BP + elevated_oldpeak, 
-                data = cleaned_df, family = binomial(link = "log"))
-rr_values <- exp(coef(rr_model))
-rr_ci <- exp(confint(rr_model))
-rr_results <- data.frame(
-  Term = names(rr_values),
-  RR = rr_values,
-  CI_lower = rr_ci[, 1],
-  CI_upper = rr_ci[, 2]
-)
-print("Relative Risks (with 95% CI):")
-print(rr_results)
-```
+# Conclusion
 
-The baseline relative risk is about 0.23 (95% CI: 0.08–0.46).
-For patients in the "elevated_BPOtherwise" category, the risk increases to roughly 1.66 (95% CI: 0.73–5.19), and for those in the "elevated_oldpeakNormal" category, the risk is around 1.21 (95% CI: 0.61–2.35).
-The wide confidence intervals suggest some uncertainty, and since they include 1, these differences may not be statistically significant.
+Based on the current analysis, there is limited evidence to suggest that elevated blood pressure or ST depression meaningfully increases the risk of cardiovascular disease in younger individuals with normal cholesterol.
+While point estimates suggest possible risk elevation, statistical significance was not achieved.
+These findings raise important questions about the sufficiency of traditional risk thresholds and underscore the need for more personalized risk assessment in early prevention.
 
-# ROC Curve and AUC for the logistic regression model with interaction
+# References
 
-```{r}
-roc_curve <- roc(cleaned_df$target_binary, predict(model_interaction, type = "response"))
-plot(roc_curve, main = "ROC Curve for Logistic Regression Model", col = "blue", lwd = 2)
-auc_value <- auc(roc_curve)
-print(paste("AUC:", auc_value))
-```
-
-The area under the ROC curve is about 0.60, suggesting that this logistic regression model only slightly outperforms random guessing.
-It provides modest discrimination between individuals with and without heart disease in this subset, but is not particularly strong.
-
-# Adjusted Logistic Regression Analysis including confounders (gender and fasting blood sugar)
-
-```{r}
-adjusted_model <- glm(target_binary ~ elevated_BP * elevated_oldpeak + gender + fastingbloodsugar, 
-                      data = cleaned_df, family = binomial)
-summary(adjusted_model)
-
-# Extract Odds Ratios (OR) for the adjusted model
-adj_or_values <- exp(coef(adjusted_model))
-adj_or_ci <- exp(confint(adjusted_model))
-adj_or_results <- data.frame(
-  Term = names(adj_or_values),
-  OR = adj_or_values,
-  CI_lower = adj_or_ci[, 1],
-  CI_upper = adj_or_ci[, 2]
-)
-print("Adjusted Odds Ratios (with 95% CI):")
-print(adj_or_results)
-
-# Tidy the adjusted model output and compute odds ratios and 95% CI
-tidy_adjusted_model <- tidy(adjusted_model, exponentiate = TRUE, conf.int = TRUE)
-print(tidy_adjusted_model)
-```
-
-The adjusted logistic regression model produces unstable estimates.
-The intercept's odds ratio is about 0.07, indicating very low baseline odds.
-The "elevated_BPOtherwise" predictor has an OR of roughly 2.51, suggesting a potential increase in risk, but its effect is not statistically significant.
-The effect for "elevated_oldpeakNormal" is extremely low and unreliable, and both the gender and fasting blood sugar estimates are unusually large with wide confidence intervals.
-These findings indicate that the model may be suffering from separation or sparse data issues, rendering the parameter estimates unstable.
-This interpretation is derived solely from your output.
-
-# PART 4b: MODEL BUILDING
-
-```{r}
-# Set seed for reproducibility
-rdmseed <- 123
-
-# Reset seed before modeling
-set.seed(rdmseed)
-
-cleaned_df <- cleaned_df %>%
-  mutate(target_binary = if_else(target == "Heart Disease", 1, 0))
-```
-
-# linear model
-
-```{r}
-# Fit a standard linear regression model using predictors
-lm_model <- linear_reg() %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions from linear model and calculate RMSE
-cleaned_df$pred_lm <- predict(lm_model, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lm)
-
-# # Plot observed vs. predicted for Linear Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lm)) +
-  geom_point(color = "darkblue", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The linear model yields an RMSE of about 0.47, suggesting that its predictions deviate from the true 0/1 outcome by nearly half on average.
-This indicates limited accuracy for distinguishing between individuals with and without heart disease.
-
-# LASSO
-
-```{r}
-set.seed(rdmseed)
-
-# Prepare recipe for LASSO with normalization
-rec <- recipe(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df) %>%
-  step_dummy(elevated_BP, elevated_oldpeak) %>%   # Convert categorical predictors to dummies
-  step_normalize(all_numeric(), -all_outcomes())
-
-# Define LASSO model with a fixed penalty
-lasso_spec <- linear_reg(penalty = 0.1, mixture = 1) %>%
-  set_engine("glmnet")
-
-# Create a workflow and fit the LASSO model
-wf <- workflow() %>%
-  add_recipe(rec) %>%
-  add_model(lasso_spec)
-
-lasso_fit <- fit(wf, data = cleaned_df)
-cleaned_df$pred_lasso <- predict(lasso_fit, new_data = cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_lasso)
-
-# Plot observed vs. predicted for LASSO Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_lasso)) +
-  geom_point(color = "darkviolet", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-```
-
-The LASSO model shows an RMSE of about 0.48, indicating that its predictions deviate from the actual 0/1 outcome by nearly half on average.
-This suggests that it provides limited predictive accuracy for identifying heart disease status in this subset.
-
-# Random Forest
-
-```{r}
-set.seed(rdmseed)
-
-# Create a workflow and fit the Random Forest model
-rf_model <- rand_forest() %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-rf_fit <- rf_model %>%
-  fit(target_binary ~ elevated_BP + elevated_oldpeak, data = cleaned_df)
-
-# Get predictions and evaluate model performance
-cleaned_df$pred_rf <- predict(rf_fit, cleaned_df)$.pred
-rmse(cleaned_df, truth = target_binary, estimate = pred_rf)
-
-# Plot observed vs. predicted for RF Model
-ggplot(cleaned_df, aes(x = target_binary, y = pred_rf)) +
-  geom_point(color = "darkgreen", alpha = 0.7, size = 3) +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 1) +
-  scale_x_continuous(limits = c(0, 1)) +
-  scale_y_continuous(limits = c(0, 1))
-
-```
-
-The random forest model produces an RMSE of about 0.47, indicating that, on average, its predicted probabilities deviate from the true 0/1 outcome by nearly half.
-This suggests limited accuracy in distinguishing between individuals with and without heart disease in this dataset.
-
-# Tuning the LASSO model (without cross-validation)
-
-```{r}
-set.seed(rdmseed)
-
-# Create a grid of penalty values for LASSO
-lasso_grid <- tibble(penalty = 10^seq(-5, 2, length.out = 50))
-
-# Define LASSO model with tunable penalty
-lasso_spec_tune <- linear_reg(penalty = tune()) %>%
-  set_engine("glmnet")
-
-# Create workflow and assign to lasso_tuning_workflow (new name)
-lasso_tuning_workflow <- workflow() %>%
-  add_model(lasso_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform grid search without CV (apparent resampling)
-lasso_tune <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = apparent(cleaned_df),
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-# Extract tuning results
-lasso_tune_df <- as.data.frame(lasso_tune$.metrics)
-
-# Plot LASSO tuning results
-ggplot(lasso_tune_df, aes(x = penalty, y = .estimate)) +
-  geom_line(size = 1, color = "steelblue") +
-  geom_point(size = 2, color = "firebrick") +
-  scale_x_log10() +
-  labs(
-    x = "Penalty (log scale)",
-    y = "RMSE",
-    title = "LASSO Tuning (No Cross-validation)",
-    subtitle = "Using apparent resampling"
-  ) +
-  theme_minimal()
-```
-
-The tuning curve shows that the RMSE stays around 0.474 at lower penalty values, then increases to roughly 0.48 at higher penalties.
-This pattern suggests that a moderate level of regularization gives the best performance, while heavier penalties lead to worse predictive accuracy.
-
-# Tuning for our Random Forest model (without cross-validation)
-
-```{r}
-# Create tuning grid for mtry and min_n
-rf_grid <- grid_regular(
-  mtry(range = c(1, 7)),
-  min_n(range = c(1, 21)),
-  levels = 7
-)
-
-# Define random forest model with tunable parameters
-rf_spec_tune <- rand_forest(
-  mtry = tune(),
-  min_n = tune(),
-  trees = 300
-) %>%
-  set_mode("regression") %>%
-  set_engine("ranger", seed = rdmseed)
-
-# Build workflow
-rf_wf_tune <- workflow() %>%
-  add_model(rf_spec_tune) %>%
-  add_recipe(rec)
-
-# Perform RF tuning without CV
-rf_tune <- tune_grid(
-  rf_wf_tune,
-  resamples = apparent(cleaned_df),
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-
-rf_tune_df <- as.data.frame(rf_tune$.metrics)
-
-# Plot RF tuning results
-ggplot(rf_tune_df, aes(x = mtry, y = min_n, fill = .estimate)) +
-  geom_tile() +
-  scale_fill_viridis_c() +
-  labs(
-    title = "Random Forest Tuning (No Cross-validation)",
-    x = "mtry",
-    y = "min_n",
-    fill = "RMSE"
-  ) +
-  theme_minimal()
-```
-
-The RMSE values for different combinations of mtry and min_n appear to hover around 0.4737 to 0.4739.
-This narrow range suggests that the model’s performance does not vary dramatically across these hyperparameter settings when no cross-validation is used.
-In other words, none of the parameter combinations stands out as substantially better or worse, indicating that random forest performance remains relatively stable within this grid.
-
-# Cross-validation for our LASSO model tuning
-
-```{r}
-set.seed(rdmseed)
-
-# Create 5-fold CV with 5 repeats
-cv_folds <- vfold_cv(cleaned_df, v = 5, repeats = 5)
-
-# Tune LASSO using CV
-lasso_tune_cv <- tune_grid(
-  lasso_tuning_workflow,
-  resamples = cv_folds,
-  grid = lasso_grid,
-  metrics = metric_set(rmse)
-)
-
-autoplot(lasso_tune_cv)
-```
-
-Based on the chart, the RMSE begins at roughly 0.497 when the regularization amount is very low, then drops to around 0.491 and stays there as the regularization increases.
-This suggests that moderate to higher levels of regularization lead to better performance, while minimal regularization yields higher error.
-
-# Cross-validation tuning for our random forest model
-
-```{r}
-# Tune RF using CV
-rf_tune_cv <- tune_grid(
-  rf_wf_tune,
-  resamples = cv_folds,
-  grid = rf_grid,
-  metrics = metric_set(rmse)
-)
-autoplot(rf_tune_cv)
-
-```
-
-Based on the chart, the pink line (representing a minimal node size of 21) shows the lowest RMSE across all numbers of randomly selected predictors (mtry).
-In contrast, the other lines, which represent smaller minimal node sizes, produce higher RMSE values.
-This pattern suggests that using a larger minimal node size leads to better performance in this grid, while varying mtry has a smaller impact on the overall error.
+American Heart Association. Understanding Blood Pressure Readings.  
+Mayo Clinic. Prehypertension: Symptoms and Causes.  
+Cleveland Clinic. Cholesterol Numbers: What Do They Mean?  
+Johns Hopkins Medicine. Lipid Panel Overview.  
+Goff DC Jr., et al. Stroke Risk in Young Adults (PMC6376358).  
+American Heart Association Journal. ST-T Abnormalities and CVD Risk.

### 2025-04-24 project-review-template.qmd
project review- MJ

---
title: Project Review Template 
author: Murphy John
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Early Cardiovascular Disease Risk Among Younger Individuals with Normal Cholesterol: The Role of Elevated Blood Pressure and ST Depression

Name of project author(s): Murtaza Yaqubi

Name of project reviewer: Murphy John

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

- Add references in your intro
- Elaborate on the importance or motivation for your research question. Is there any prior evidence of an association? Why is this an important topic to explore?
- Why are you restricting the age to under 50? It looks like you have data for other ages. If you want to emphasis that this study is looking in young adults then include more background on CVD prevalence in young adults. Maybe there are different risk factors in this age group compared to older individuals? That could provide more motivation for your research question.

### Summary assessment
* some contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

- Line 17, say "individuals under 50 years of age" or something to clarify the age range.
- What do you hypothesize? Why do you think that?

### Summary assessment
* question/hypotheses somewhat explained

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

- Line 25, just say "the primary outcome is presence or absence of heart disease". I'm not sure what the word target means in this sentence.
- Add a reference to line 23 "These thresholds were chosen based on clinical relevance".
- Talk more about the data source. Where did it come from (what specific hospital)? When was it collected? Who was included in the data collection? What was the inclusion/exclusion criteria for this study? 
- BMI, family history, SES, and smoking status are confounders of CVD. If you have data on these, include it. If not, this needs to be discussed in the limitations.
- Your study population needs to be more specific. Is it adults worldwide? In India? In a specific area of India?
- What is oldpeak?
- After the inclusion/exclusion critera, how many individuals/obervation are in your data?

### Summary assessment
* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

- You need to load your processed data at the start of each script. None of your scripts are running because of this. To make sure everything works and is reproducible, clear your environment before you run anything in a new script.
- The data processing looks good so far but you should include your inclusion/exclusion criteria in that script (ie filter age < 50). 
- Your eda should be of the data used in your analysis (ie age < 50). This will likely change your findings and perhaps your analysis decisions.
- Your plots look nice.

### Summary assessment
* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

- You have processing code in your analysis-code/01_inferential_stats.qmd file. This should be in the processing code file. You also have eda in this file that should go in the eda folder.

In analysis-code/02_modeling.qmd:

- Why are you fitting linear models to a binary variable?
- Use logistic regression instead of linear.
- If you want to use lasso, use logistic_reg() in your spec.
- For your random forest model, set mode to "Classification" instead of regression.
- RMSE is for continuous outcomes, use AUC and/or accuracy for your metrics instead.
- Use test/train data split and/or cross-validation for model selection.

analysis-code/03_model_tuning.qmd:

- You need to load your model objects from the previous scripts or merge this script with the previous one.
- It doesn't make sense to use rmse for metrics
- You need to use binary outcome modeling methods in the modeling script then update this script (or merge the two)

### Summary assessment
* wrong/inadequate analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

- The descriptive statistics table should be of the data you analyze (age < 50).
- The tables and figures need to render to the manuscript doc

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

- For the current analysis set up, the discussion is set up well. Once you implement the correct analysis methods you'll likely get different results.
- Your limitations sections needs to address the confounders that you don't control for. Also if there are any data collection limitations list that here.

### Summary assessment
* minor parts wrong, missing or unclear

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

- Project structure looks good.
- Your analysis code contains some data processing and eda that should be in different locations.

### Summary assessment
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

- looks good

### Summary assessment
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

- Your manuscript wouldn't render as is. You need to fix your figures and tables code.
- None of your eda or analysis scripts run because you haven't loaded the data in them. Use cardio_df <- readRDS(here::here("data/processed-data/cleaned_df.rds"))
- Project is not reproducible- you need to load data and model objects in each script.

### Summary assessment
* small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

Given the data presented, the project analysis is thorough. It isn't correct in the current methods but with minor edits, you should get more appropriate results. My biggest concerns are the eda on the unused data, the incorrect modeling methods, and the lack of control for major CVD confounders.

### Summary assessment
* strong level of thoroughness

## Further comments

- References are not properly loaded. Consider using .bib file. Zotero is a nice software for storing references and generating .bib files.

### 2025-04-26 project-review-Gonzalez.qmd

--- 2025-04-24 project-review-template.qmd
@@
 title: Project Review Template 
-author: Murphy John
+author: YOUR NAME
 date: date-modified
@@
 
-Name of project reviewer: Murphy John
+Name of project reviewer: Alexis Gonzalez
 
@@
 
-- Add references in your intro
-- Elaborate on the importance or motivation for your research question. Is there any prior evidence of an association? Why is this an important topic to explore?
-- Why are you restricting the age to under 50?
-It looks like you have data for other ages.
-If you want to emphasis that this study is looking in young adults then include more background on CVD prevalence in young adults.
-Maybe there are different risk factors in this age group compared to older individuals?
-That could provide more motivation for your research question.
+The introduction needs contexualization. Why is this issue important? What is the current scope? Maybe provide some global statistics and some India specific statistics.
 
-### Summary assessment
+### Summary Assessment
+
 * some contextualization and motivation
@@
 
-- Line 17, say "individuals under 50 years of age" or something to clarify the age range.
-- What do you hypothesize? Why do you think that?
+Question is clear, I understand your goal.
 
 ### Summary assessment
-* question/hypotheses somewhat explained
+* question/hypotheses unclear
 
@@
 
-- Line 25, just say "the primary outcome is presence or absence of heart disease". I'm not sure what the word target means in this sentence.
-- Add a reference to line 23 "These thresholds were chosen based on clinical relevance".
-- Talk more about the data source.
-Where did it come from (what specific hospital)?
-When was it collected?
-Who was included in the data collection?
-What was the inclusion/exclusion criteria for this study?
-- BMI, family history, SES, and smoking status are confounders of CVD. If you have data on these, include it. If not, this needs to be discussed in the limitations.
-- Your study population needs to be more specific. Is it adults worldwide? In India? In a specific area of India?
-- What is oldpeak?
-- After the inclusion/exclusion critera, how many individuals/obervation are in your data?
+I do not really understand the dataset.
+How was this data obtained, where is this hospital specifically?
+Might be good to include some parts of the dataset description from the website you got it from.
 
 ### Summary assessment
+
 * source and overall structure of data somewhat explained
@@
 
-- You need to load your processed data at the start of each script.
-None of your scripts are running because of this.
-To make sure everything works and is reproducible, clear your environment before you run anything in a new script.
-- The data processing looks good so far but you should include your inclusion/exclusion criteria in that script (ie filter age < 50). 
-- Your eda should be of the data used in your analysis (ie age < 50). This will likely change your findings and perhaps your analysis decisions.
-- Your plots look nice.
+Unfortunately, I am unable to load the data in your manuscript. I still want to give feedback so I will look at your code file.
+
+- Instead of the basic table output consider using flextable for KableExtra
+- It doesn't look like you include any of your figures from EDA. Those are actually very helpful for getting an idea of distributions and burden.
+- The EDA should help explain the formulation of your question or provide some reason why.
+- The code is very very neat and the graphs look great!!
 
 ### Summary assessment
+
 * some weaknesses in wrangling and exploratory component
@@
 
-- You have processing code in your analysis-code/01_inferential_stats.qmd file. This should be in the processing code file. You also have eda in this file that should go in the eda folder.
-
-In analysis-code/02_modeling.qmd:
-
-- Why are you fitting linear models to a binary variable?
-- Use logistic regression instead of linear.
-- If you want to use lasso, use logistic_reg() in your spec.
-- For your random forest model, set mode to "Classification" instead of regression.
-- RMSE is for continuous outcomes, use AUC and/or accuracy for your metrics instead.
-- Use test/train data split and/or cross-validation for model selection.
-
-analysis-code/03_model_tuning.qmd:
-
-- You need to load your model objects from the previous scripts or merge this script with the previous one.
-- It doesn't make sense to use rmse for metrics
-- You need to use binary outcome modeling methods in the modeling script then update this script (or merge the two)
+I do not think you need the linear model here for a binary outcome, the log model makes more sense.
 
 ### Summary assessment
-* wrong/inadequate analysis
+
+* defensible but not optimal analysis
 
@@
 
-- The descriptive statistics table should be of the data you analyze (age < 50).
-- The tables and figures need to render to the manuscript doc
+I like how the results section is sectioned out, makes it easy to follow. Figures look nice.
 
 ### Summary assessment
-* results are presented ok, with room for improvement
+
+* results are very well presented
 
@@
 
-- For the current analysis set up, the discussion is set up well. Once you implement the correct analysis methods you'll likely get different results.
-- Your limitations sections needs to address the confounders that you don't control for. Also if there are any data collection limitations list that here.
+The discussion is lacking in explanation of the results.
 
@@
  * minor parts wrong, missing or unclear
+
+## Further comments
+
+_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._
 
@@
 
-- Project structure looks good.
-- Your analysis code contains some data processing and eda that should be in different locations.
+Porject is structured well, easy to follow
 
 ### Summary assessment
+
 * well structured
@@
 
-- looks good
+Code is well documented and easy to follow
 
 ### Summary assessment
+
 * fully and well documented
@@
 
-- Your manuscript wouldn't render as is. You need to fix your figures and tables code.
-- None of your eda or analysis scripts run because you haven't loaded the data in them. Use cardio_df <- readRDS(here::here("data/processed-data/cleaned_df.rds"))
-- Project is not reproducible- you need to load data and model objects in each script.
+The manuscript needs to be fixed and I was unable to run the cross validation for your RF model
 
 ### Summary assessment
+
 * small parts not reproducible or required manual intervention
@@
 
-Given the data presented, the project analysis is thorough.
-It isn't correct in the current methods but with minor edits, you should get more appropriate results.
-My biggest concerns are the eda on the unused data, the incorrect modeling methods, and the lack of control for major CVD confounders.
+Overall I would say the study was thorough.
+I think you should go back and expand on your introduction, discussion and conclusion.
+Explain the reason for using LASSO and RF.
+Also make sure to add some EDA plots into your manuscript.
 
 ### Summary assessment
-* strong level of thoroughness
+
+* decent level of thorougness
 
@@
 
-- References are not properly loaded. Consider using .bib file. Zotero is a nice software for storing references and generating .bib files.
+don't forget to properly format references! This is a great study I can't wait to read the final product.

### 2025-05-01 code/analysis-code/analysis-code.qmd
final touches

---
title: "Inferential Analysis"
author: "Murtaza Yaqubi"
format: html
editor: visual
---

## Load/install core libraries

```{r}
library(here)
library(readr)
library(dplyr)
library(tidyverse)
```

# Read in the data

```{r}
df_clean <- readRDS(here("data","processed-data","cleaned_df.rds"))
```

# Inferential Analysis

```{r}
# Numeric outcome
df_inf <- df_clean %>% 
  mutate(outcome_num = if_else(outcome == "Disease", 1L, 0L))

# Chi-square tests
chisq_bp <- chisq.test(table(df_inf$bp_status, df_inf$outcome))
chisq_st <- chisq.test(table(df_inf$st_status, df_inf$outcome))

# Save chi-square results to tables
chisq_bp_tbl <- broom::tidy(chisq_bp)
saveRDS(chisq_bp_tbl, here("results","tables","table2_chisq_bp.rds"))
chisq_st_tbl <- broom::tidy(chisq_st)
saveRDS(chisq_st_tbl, here("results","tables","table3_chisq_st.rds"))

# Print tests
print(chisq_bp)
print(chisq_st)

# Logistic base model
mod_base <- glm(outcome_num ~ bp_status + st_status, data=df_inf, family=binomial)
base_coef <- summary(mod_base)$coefficients
or_base <- exp(base_coef[ ,"Estimate"])
ci_base_l <- exp(base_coef[ ,"Estimate"] - 1.96*base_coef[ ,"Std. Error"])
ci_base_u <- exp(base_coef[ ,"Estimate"] + 1.96*base_coef[ ,"Std. Error"])
tbl_base_or <- tibble(Term = rownames(base_coef), OR = or_base, CI_low = ci_base_l, CI_up = ci_base_u)

# Save base-model ORs
saveRDS(tbl_base_or, here("results","tables","table4_or_base_model.rds"))
print(tbl_base_or)

# Interaction model
mod_int <- glm(outcome_num ~ bp_status*st_status, data=df_inf, family=binomial)
int_coef <- summary(mod_int)$coefficients
or_int <- exp(int_coef[ ,"Estimate"])
ci_int_l <- exp(int_coef[ ,"Estimate"] - 1.96*int_coef[ ,"Std. Error"])
ci_int_u <- exp(int_coef[ ,"Estimate"] + 1.96*int_coef[ ,"Std. Error"])
tbl_int_or <- tibble(Term = rownames(int_coef), OR = or_int, CI_low = ci_int_l, CI_up = ci_int_u)

# Save interaction-model ORs
saveRDS(tbl_int_or, here("results","tables","table5_or_interaction_model.rds"))
print(tbl_int_or)

# Adjusted model
mod_adj <- glm(outcome_num ~ bp_status*st_status + gender + fasting_blood_sugar, data=df_inf, family=binomial)
adj_coef <- summary(mod_adj)$coefficients
or_adj <- exp(adj_coef[ ,"Estimate"])
ci_adj_l <- exp(adj_coef[ ,"Estimate"] - 1.96*adj_coef[ ,"Std. Error"])
ci_adj_u <- exp(adj_coef[ ,"Estimate"] + 1.96*adj_coef[ ,"Std. Error"])
tbl_adj_or <- tibble(Term = rownames(adj_coef), OR = or_adj, CI_low = ci_adj_l, CI_up = ci_adj_u)

# Save adjusted-model ORs
saveRDS(tbl_adj_or, here("results","tables","table6_or_adjusted_model.rds"))
print(tbl_adj_or)

# ROC safeguard
preds_int <- predict(mod_int, type = "response")
if (length(unique(preds_int)) > 2 && length(unique(df_inf$outcome_num)) == 2) {
  auc_val <- auc(roc(df_inf$outcome_num, preds_int))
  print(paste("AUC:", round(auc_val, 3)))

  # Save AUC value for interaction model
  auc_tbl <- tibble(Model = "Interaction_Model", AUC = auc_val)
  saveRDS(auc_tbl, here("results", "tables", "table7_auc_interaction_model.rds"))
} else message("ROC not generated: single class or no variability")

# Recreate ROC object for interaction model
roc_int <- roc(df_inf$outcome_num, preds_int)

# Annotate and save ROC plot with AUC on-plot
ggroc(roc_int, legacy.axes = TRUE, color = "red") +
  labs(
    title = "ROC Curve for Interaction Model",
    x     = "False Positive Rate",
    y     = "True Positive Rate"
  ) +
  annotate(
    "text",
    x = 0.6, y = 0.2,
    label = paste("AUC =", round(auc_val, 3)),
    hjust = 0,
    size = 5,
    color = "black"
  ) +
  theme_wsj() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  ) -> auc_plot_annotated

# Save annotated ROC plot
ggsave(
  filename = here("results", "figures", "fig9_roc_interaction_annotated.png"),
  plot     = auc_plot_annotated,
  width    = 10,
  height   = 5,
  units    = "in",
  dpi      = 300
)

plot(auc_plot_annotated)
```

# 6. Predictive Modeling

```{r}
# Prepare data for modeling
set.seed(123)
df_mod <- df_inf %>% 
  mutate(outcome = factor(outcome, levels=c("NoDisease","Disease")))

# Train-test split
set.seed(123)
data_split <- initial_split(df_mod, prop=0.7, strata=outcome)
train <- training(data_split); 
test <- testing(data_split)

# Recipe 
rec <- recipe(outcome ~ bp_status + st_status, data=train) %>% 
  step_dummy(all_nominal_predictors())
```

# Logistic regression model

```{r}
set.seed(123)
log_spec  <- logistic_reg() %>% set_engine("glm") %>% set_mode("classification")
log_wf    <- workflow() %>% add_model(log_spec) %>% add_recipe(rec)
log_fit   <- fit(log_wf, data = train)
log_preds <- predict(log_fit, test, type = "prob") %>% bind_cols(test)
log_auc   <- roc_auc(log_preds, truth = outcome, `.pred_Disease`)
log_acc   <- accuracy(predict(log_fit, test, type = "class") %>% bind_cols(test), truth = outcome, estimate = .pred_class)

# Save logistic modeling results
saveRDS(log_auc, here("results", "tables", "logistic_auc.rds"))
saveRDS(log_acc, here("results", "tables", "logistic_accuracy.rds"))
print(log_auc)
print(log_acc)

```

# LASSO logistic

```{r}
tune_ctrl  <- vfold_cv(train, v = 5)
set.seed(123)
lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) %>% set_engine("glmnet") %>% set_mode("classification")
lasso_wf   <- workflow() %>% add_model(lasso_spec) %>% add_recipe(rec)
lasso_res  <- tune_grid(lasso_wf, resamples = tune_ctrl, grid = 20, metrics = metric_set(roc_auc))
best_lasso <- select_best(lasso_res, metric = "roc_auc")
lasso_final<- finalize_workflow(lasso_wf, best_lasso)
lasso_fit  <- fit(lasso_final, data = train)
lasso_preds<- predict(lasso_fit, test, type = "prob") %>% bind_cols(test)
lasso_auc  <- roc_auc(lasso_preds, truth = outcome, `.pred_Disease`)
lasso_acc  <- accuracy(predict(lasso_fit, test, type = "class") %>% bind_cols(test), truth = outcome, estimate = .pred_class)

# Save LASSO modeling results
saveRDS(lasso_auc, here("results", "tables", "lasso_auc.rds"))
saveRDS(lasso_acc, here("results", "tables", "lasso_accuracy.rds"))
print(lasso_auc)
print(lasso_acc)
```

# Random forest

```{r}
set.seed(123)
rf_spec <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) %>% set_engine("ranger") %>% set_mode("classification")
rf_wf   <- workflow() %>% add_model(rf_spec) %>% add_recipe(rec)
rf_res  <- tune_grid(rf_wf, resamples = tune_ctrl, grid = 20, metrics = metric_set(roc_auc))
best_rf <- select_best(rf_res, metric = "roc_auc")
rf_final<- finalize_workflow(rf_wf, best_rf)
rf_fit  <- fit(rf_final, data = train)
rf_preds<- predict(rf_fit, test, type = "prob") %>% bind_cols(test)
rf_auc  <- roc_auc(rf_preds, truth = outcome, `.pred_Disease`)
rf_acc  <- accuracy(predict(rf_fit, test, type = "class") %>% bind_cols(test), truth = outcome, estimate = .pred_class)

# Save Random Forest modeling results
saveRDS(rf_auc, here("results", "tables", "rf_auc.rds"))
saveRDS(rf_acc, here("results", "tables", "rf_accuracy.rds"))
print(rf_auc)
print(rf_acc)
```

# 7. Model Comparison

```{r}
model_comp <- tibble(
  Model    = c("Logistic", "LASSO", "Random Forest"),
  AUC      = c(log_auc$.estimate, lasso_auc$.estimate, rf_auc$.estimate),
  Accuracy = c(log_acc$.estimate, lasso_acc$.estimate, rf_acc$.estimate)
)
# Save model comparison table
saveRDS(model_comp, here("results", "tables", "model_comparison.rds"))
print(model_comp)
```

### 2025-05-01 code/eda-code/eda-code.qmd
final touches

---
title: "Exploratory Data Analysis"
author: "Murtaza"
format: html
editor: visual
---

# Load/install core libraries

```{r}
library(here)
library(readr)
library(dplyr)
library(tidyverse)
```

# Read in the data

```{r}
df_clean <- readRDS(here("data","processed-data","cleaned_df.rds"))
```

# Exploratory Data Analysis

```{r}
# Summary stats
summary_stats_tbl <- df_clean %>%
  summarise(
    Age_min           = min(age, na.rm = TRUE),
    Age_max           = max(age, na.rm = TRUE),
    Age_mean          = mean(age, na.rm = TRUE),
    Age_sd            = sd(age, na.rm = TRUE),
    RestingBP_min     = min(resting_bp, na.rm = TRUE),
    RestingBP_max     = max(resting_bp, na.rm = TRUE),
    RestingBP_mean    = mean(resting_bp, na.rm = TRUE),
    RestingBP_sd      = sd(resting_bp, na.rm = TRUE),
    Chol_min          = min(cholesterol, na.rm = TRUE),
    Chol_max          = max(cholesterol, na.rm = TRUE),
    Chol_mean         = mean(cholesterol, na.rm = TRUE),
    Chol_sd           = sd(cholesterol, na.rm = TRUE),
    HR_min            = min(max_heart_rate, na.rm = TRUE),
    HR_max            = max(max_heart_rate, na.rm = TRUE),
    HR_mean           = mean(max_heart_rate, na.rm = TRUE),
    HR_sd             = sd(max_heart_rate, na.rm = TRUE),
    ST_min            = min(st_depression, na.rm = TRUE),
    ST_max            = max(st_depression, na.rm = TRUE),
    ST_mean           = mean(st_depression, na.rm = TRUE),
    ST_sd             = sd(st_depression, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
  separate(stat, into = c("Variable","Statistic"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Variable, values_from = value)

print(summary_stats_tbl)

# Save summary table
saveRDS(summary_stats_tbl, here("results","tables","table1_summary_statistics.rds"))

## Visualizations

# Correlation matrix
p_corr <- ggcorrplot(cor(df_clean %>% 
                 select(age, resting_bp, cholesterol, max_heart_rate, st_depression),
                         use = "pairwise.complete.obs"), lab=TRUE)
print(p_corr)
ggsave(filename = here("results","figures","fig1_correlation_matrix.png"), plot = p_corr, width = 10, height = 5)

# Histogram of Age 
p_age <- df_clean %>% 
  ggplot(aes(age)) + 
  geom_histogram(bins = 30, fill = "skyblue", color = "black") + 
  geom_density(aes(y = after_stat(count) * 1.2), color = "red", size = 1.2) +
  labs(
    title = "Distribution of Age",
    x     = "Age (years)",
    y     = "Count"
  ) +
  theme_wsj() +
  theme(
    # re-enable & style the axis titles
    axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
    axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
  )
print(p_age)
ggsave(filename = here("results","figures","fig2_age_histogram.png"), plot = p_age, width = 10, height = 5)

# Histogram of Cholesterol 
p_chol <- df_clean %>% 
  ggplot(aes(cholesterol)) + 
  geom_histogram(bins = 30, fill = "gold", color = "black") + 
  geom_density(aes(y = after_stat(count) * 1.2), color = "red", size = 1.2) +
  labs(
    title = "Distribution of Cholesterol",
    x     = "Cholesterol Level (mg/dL)",
    y     = "Count"
  ) +
  theme_wsj() +
  theme(
    axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
    axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
  )
print(p_chol)
ggsave(filename = here("results","figures", "fig3_cholesterol_histogram.png"), plot = p_chol, width = 10, height = 5)

# Barplot of CVD prevalence by BP status
p_bp <- ggplot(df_clean, aes(x = bp_status, fill = outcome)) +
  geom_bar(position = "fill", width = 0.6) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "CVD Prevalence by Blood Pressure Status",
    x     = "Blood Pressure Status",
    y     = "Percent of Individuals",
    fill  = "CVD"
  ) +
  theme_wsj() +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
  )
print(p_bp)
ggsave(filename = here("results","figures","fig4_bp_barplot.png"), plot = p_bp, width = 10, height = 5)

# Barplot of CVD prevalence by ST-Depression status
p_st <- ggplot(df_clean, aes(x = st_status, fill = outcome)) +
  geom_bar(position = "fill", width = 0.6) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "CVD Prevalence by ST-Depression Status",
       x = "ST-Depression Status", y = "Percent of Individuals", fill = "CVD") +
  theme_wsj() +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
  )
print(p_st)
ggsave(filename = here("results","figures","fig5_st_barplot.png"), plot = p_st, width = 10, height = 5)

# Interaction Heatmap: counts by BP & ST-Depression, faceted by CVD
data_heat <- df_clean %>% count(bp_status, st_status, outcome)

p_heat <- ggplot(data_heat, aes(x = bp_status, y = st_status, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 4) +
  facet_wrap(~ outcome) +
  labs(
    title = "Counts by BP & ST Status (facet by Disease)",
    x = "BP Status",
    y = "ST Status",
    fill = "Count"
  ) +
  theme_wsj() +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5)
  )
print(p_heat)
ggsave(
  filename = here("results", "figures", "fig6_interaction_heatmap.png"), plot     = p_heat, width    = 10, height   = 5)

# Boxplot of age distribution by CVD within BP categories
p_box <- ggplot(df_clean, aes(x = bp_status, y = age, fill = outcome)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Age Distribution by CVD Status within BP Categories",
       x = "Blood Pressure Status", y = "Age (years)", fill = "CVD") +
  theme_wsj() +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5)
  )
print(p_box)
ggsave(filename = here("results","figures","fig7_box_age_bp.png"), plot = p_box, width = 10, height = 5)

# Scatterplot of ST-Depression vs. resting BP by CVD status
p_scatter <- ggplot(df_clean, aes(x = resting_bp, y = st_depression, color = outcome)) +
  geom_point(alpha = 1, size = 3) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "ST-Depression vs. Resting BP by CVD Status",
       x = "Resting Blood Pressure (mmHg)", y = "ST-Depression (oldpeak)", color = "CVD") +
  theme_wsj() +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
  )
print(p_scatter)
ggsave(filename = here("results","figures","fig8_scatter_st_bp.png"), plot = p_scatter, width = 10, height = 5)
```

### 2025-05-01 code/full-code/full-code.qmd
final touches

--- 2025-05-01 code/analysis-code/analysis-code.qmd
@@
 ---
-title: "Inferential Analysis"
+title: "Final Project full code"
 author: "Murtaza Yaqubi"
@@
 
-## Load/install core libraries
+# 1. Setup
 
 ```{r}
-library(here)
-library(readr)
-library(dplyr)
-library(tidyverse)
+set.seed(123)  # Global seed for reproducibility
 ```
 
-# Read in the data
+# Load core libraries
 
 ```{r}
-df_clean <- readRDS(here("data","processed-data","cleaned_df.rds"))
+library(tidyverse)    # data manipulation & ggplot2
+library(here)         # file paths
+library(ggcorrplot)   # correlation plots
+library(tidymodels)   # modeling framework
+library(broom)        # tidying model outputs
+library(pROC)         # ROC analysis
+library(ggthemes)     # themes
 ```
 
-# Inferential Analysis
+# 2. Data Import & Initial Cleaning
+
+```{r}
+# Import raw data
+df_raw <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
+
+# Rename columns for clarity
+df <- df_raw %>%
+  rename(
+    patient_id         = patientid,
+    age                = age,
+    gender             = gender,
+    chest_pain_type    = chestpain,
+    resting_bp         = restingBP,
+    cholesterol        = serumcholestrol,
+    fasting_blood_sugar= fastingbloodsugar,
+    resting_ecg        = restingrelectro,
+    max_heart_rate     = maxheartrate,
+    exercise_angina    = exerciseangia,
+    st_depression      = oldpeak,
+    slope              = slope,
+    num_major_vessels  = noofmajorvessels,
+    outcome            = target
+  )
+
+# Recode 0 cholesterol as NA and out-of-range (>564) to NA
+df <- df %>%
+  mutate(
+    cholesterol = na_if(cholesterol, 0),
+    cholesterol = if_else(cholesterol > 564, NA_real_, cholesterol)
+  )
+
+# Convert binary/nominal variables to factors with labels
+df <- df %>%
+  mutate(
+    gender           = factor(gender, levels = c(0,1), labels = c("Female","Male")),
+    chest_pain_type  = factor(chest_pain_type, levels = 0:3,
+                              labels = c("Typical Angina","Atypical","Non-anginal","Asymptomatic")),
+    fasting_blood_sugar = factor(fasting_blood_sugar, levels = c(0,1), labels = c("False","True")),
+    resting_ecg      = factor(resting_ecg, levels=0:2,
+                              labels=c("Normal","ST-T Abnormality","LV Hypertrophy")),
+    exercise_angina  = factor(exercise_angina, levels=c(0,1), labels=c("No","Yes")),
+    slope            = factor(slope, levels=1:3, labels=c("Upsloping","Flat","Downsloping")),
+    outcome          = factor(outcome, levels=c(0,1), labels=c("NoDisease","Disease"))
+  )
+```
+
+# 3. Study Population Filter
+
+```{r}
+# Focus: age < 50 & normal cholesterol (<200)
+df_clean <- df %>%
+  filter(age < 50, !is.na(cholesterol), cholesterol < 200) %>%
+  mutate(
+    bp_status = factor(if_else(between(resting_bp, 120, 129), "Elevated", "Normal")),
+    st_status = factor(if_else(st_depression > 1, "Elevated", "Normal"))
+  )
+
+# Save cleaned data
+saveRDS(df_clean, here("data","processed-data","cleaned_df.rds"))
+```
+
+# 4. Exploratory Data Analysis
+
+```{r}
+# Summary stats
+summary_stats_tbl <- df_clean %>%
+  summarise(
+    Age_min           = min(age, na.rm = TRUE),
+    Age_max           = max(age, na.rm = TRUE),
+    Age_mean          = mean(age, na.rm = TRUE),
+    Age_sd            = sd(age, na.rm = TRUE),
+    RestingBP_min     = min(resting_bp, na.rm = TRUE),
+    RestingBP_max     = max(resting_bp, na.rm = TRUE),
+    RestingBP_mean    = mean(resting_bp, na.rm = TRUE),
+    RestingBP_sd      = sd(resting_bp, na.rm = TRUE),
+    Chol_min          = min(cholesterol, na.rm = TRUE),
+    Chol_max          = max(cholesterol, na.rm = TRUE),
+    Chol_mean         = mean(cholesterol, na.rm = TRUE),
+    Chol_sd           = sd(cholesterol, na.rm = TRUE),
+    HR_min            = min(max_heart_rate, na.rm = TRUE),
+    HR_max            = max(max_heart_rate, na.rm = TRUE),
+    HR_mean           = mean(max_heart_rate, na.rm = TRUE),
+    HR_sd             = sd(max_heart_rate, na.rm = TRUE),
+    ST_min            = min(st_depression, na.rm = TRUE),
+    ST_max            = max(st_depression, na.rm = TRUE),
+    ST_mean           = mean(st_depression, na.rm = TRUE),
+    ST_sd             = sd(st_depression, na.rm = TRUE)
+  ) %>%
+  pivot_longer(everything(), names_to = "stat", values_to = "value") %>%
+  separate(stat, into = c("Variable","Statistic"), sep = "_(?=[^_]+$)") %>%
+  pivot_wider(names_from = Variable, values_from = value)
+
+print(summary_stats_tbl)
+
+# Save summary table
+saveRDS(summary_stats_tbl, here("results","tables","table1_summary_statistics.rds"))
+
+## Visualizations
+
+# Correlation matrix
+p_corr <- ggcorrplot(cor(df_clean %>% 
+                 select(age, resting_bp, cholesterol, max_heart_rate, st_depression),
+                         use = "pairwise.complete.obs"), lab=TRUE)
+print(p_corr)
+ggsave(filename = here("results","figures","fig1_correlation_matrix.png"), plot = p_corr, width = 5, height = 5)
+
+# Histogram of Age 
+p_age <- df_clean %>% 
+  ggplot(aes(age)) + 
+  geom_histogram(bins = 30, fill = "skyblue", color = "black") + 
+  geom_density(aes(y = after_stat(count) * 1.2), color = "red", size = 1.2) +
+  labs(
+    title = "Distribution of Age",
+    x     = "Age (years)",
+    y     = "Count"
+  ) +
+  theme_wsj() +
+  theme(
+    # re-enable & style the axis titles
+    axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
+    axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
+  )
+print(p_age)
+ggsave(filename = here("results","figures","fig2_age_histogram.png"), plot = p_age, width = 10, height = 5)
+
+# Histogram of Cholesterol 
+p_chol <- df_clean %>% 
+  ggplot(aes(cholesterol)) + 
+  geom_histogram(bins = 30, fill = "gold", color = "black") + 
+  geom_density(aes(y = after_stat(count) * 1.2), color = "red", size = 1.2) +
+  labs(
+    title = "Distribution of Cholesterol",
+    x     = "Cholesterol Level (mg/dL)",
+    y     = "Count"
+  ) +
+  theme_wsj() +
+  theme(
+    axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
+    axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
+  )
+print(p_chol)
+ggsave(filename = here("results","figures", "fig3_cholesterol_histogram.png"), plot = p_chol, width = 10, height = 5)
+
+# Barplot of CVD prevalence by BP status
+p_bp <- ggplot(df_clean, aes(x = bp_status, fill = outcome)) +
+  geom_bar(position = "fill", width = 0.6) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  labs(
+    title = "CVD Prevalence by Blood Pressure Status",
+    x     = "Blood Pressure Status",
+    y     = "Percent of Individuals",
+    fill  = "CVD"
+  ) +
+  theme_wsj() +
+  theme(
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
+  )
+print(p_bp)
+ggsave(filename = here("results","figures","fig4_bp_barplot.png"), plot = p_bp, width = 8, height = 5)
+
+# Barplot of CVD prevalence by ST-Depression status
+p_st <- ggplot(df_clean, aes(x = st_status, fill = outcome)) +
+  geom_bar(position = "fill", width = 0.6) +
+  scale_y_continuous(labels = scales::percent_format()) +
+  labs(title = "CVD Prevalence by ST-Depression Status",
+       x = "ST-Depression Status", y = "Percent of Individuals", fill = "CVD") +
+  theme_wsj() +
+  theme(
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
+  )
+print(p_st)
+ggsave(filename = here("results","figures","fig5_st_barplot.png"), plot = p_st, width = 8, height = 5)
+
+# Interaction Heatmap: counts by BP & ST-Depression, faceted by CVD
+data_heat <- df_clean %>% count(bp_status, st_status, outcome)
+
+p_heat <- ggplot(data_heat, aes(x = bp_status, y = st_status, fill = n)) +
+  geom_tile(color = "white") +
+  geom_text(aes(label = n), size = 4) +
+  facet_wrap(~ outcome) +
+  labs(
+    title = "Counts by BP & ST Status (facet by Disease)",
+    x = "BP Status",
+    y = "ST Status",
+    fill = "Count"
+  ) +
+  theme_wsj() +
+  theme(
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5)
+  )
+print(p_heat)
+ggsave(
+  filename = here("results", "figures", "fig6_interaction_heatmap.png"), plot = p_heat, width = 8, height = 5)
+
+# Boxplot of age distribution by CVD within BP categories
+p_box <- ggplot(df_clean, aes(x = bp_status, y = age, fill = outcome)) +
+  geom_boxplot(alpha = 0.7) +
+  labs(title = "Age Distribution by CVD Status within BP Categories",
+       x = "Blood Pressure Status", y = "Age (years)", fill = "CVD") +
+  theme_wsj() +
+  theme(
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5)
+  )
+print(p_box)
+ggsave(filename = here("results","figures","fig7_box_age_bp.png"), plot = p_box, width = 8, height = 5)
+
+# Scatterplot of ST-Depression vs. resting BP by CVD status
+p_scatter <- ggplot(df_clean, aes(x = resting_bp, y = st_depression, color = outcome)) +
+  geom_point(alpha = 1, size = 3) +
+  geom_smooth(method = "loess", se = FALSE) +
+  labs(title = "ST-Depression vs. Resting BP by CVD Status",
+       x = "Resting Blood Pressure (mmHg)", y = "ST-Depression (oldpeak)", color = "CVD") +
+  theme_wsj() +
+  theme(
+    axis.title.x = element_text(size = 10),
+    axis.title.y = element_text(size = 10),
+    plot.title   = element_text(size = 18, face = "bold", hjust = 0.5)
+  )
+print(p_scatter)
+ggsave(filename = here("results","figures","fig8_scatter_st_bp.png"), plot = p_scatter, width = 8, height = 5)
+```
+
+# 5. Inferential Analysis
 
@@
 # Numeric outcome
-df_inf <- df_clean %>% 
-  mutate(outcome_num = if_else(outcome == "Disease", 1L, 0L))
+df_inf <- df_clean %>% mutate(outcome_num = if_else(outcome == "Disease", 1L, 0L))
 

### 2025-05-01 code/processing-code/processing-code.qmd
final touches

--- 2025-05-01 code/analysis-code/analysis-code.qmd
@@
 ---
-title: "Inferential Analysis"
+title: "Processing code"
 author: "Murtaza Yaqubi"
@@
 
-## Load/install core libraries
+# Load/instal core libraries
 
 ```{r}
+library(readr)
 library(here)
-library(readr)
 library(dplyr)
@@
 
-# Read in the data
+# 2. Data Import & Initial Cleaning
 
 ```{r}
-df_clean <- readRDS(here("data","processed-data","cleaned_df.rds"))
+# Import raw data
+df_raw <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))
+
+# Rename columns for clarity
+df <- df_raw %>%
+  rename(
+    patient_id         = patientid,
+    age                = age,
+    gender             = gender,
+    chest_pain_type    = chestpain,
+    resting_bp         = restingBP,
+    cholesterol        = serumcholestrol,
+    fasting_blood_sugar= fastingbloodsugar,
+    resting_ecg        = restingrelectro,
+    max_heart_rate     = maxheartrate,
+    exercise_angina    = exerciseangia,
+    st_depression      = oldpeak,
+    slope              = slope,
+    num_major_vessels  = noofmajorvessels,
+    outcome            = target
+  )
+
+# Recode 0 cholesterol as NA and out-of-range (>564) to NA
+df <- df %>%
+  mutate(
+    cholesterol = na_if(cholesterol, 0),
+    cholesterol = if_else(cholesterol > 564, NA_real_, cholesterol)
+  )
+
+# Convert binary/nominal variables to factors with labels
+df <- df %>%
+  mutate(
+    gender           = factor(gender, levels = c(0,1), labels = c("Female","Male")),
+    chest_pain_type  = factor(chest_pain_type, levels = 0:3,
+                              labels = c("Typical Angina","Atypical","Non-anginal","Asymptomatic")),
+    fasting_blood_sugar = factor(fasting_blood_sugar, levels = c(0,1), labels = c("False","True")),
+    resting_ecg      = factor(resting_ecg, levels=0:2,
+                              labels=c("Normal","ST-T Abnormality","LV Hypertrophy")),
+    exercise_angina  = factor(exercise_angina, levels=c(0,1), labels=c("No","Yes")),
+    slope            = factor(slope, levels=1:3, labels=c("Upsloping","Flat","Downsloping")),
+    outcome          = factor(outcome, levels=c(0,1), labels=c("NoDisease","Disease"))
+  )
 ```
 
-# Inferential Analysis
+# 3. Study Population Filter
 
 ```{r}
-# Numeric outcome
-df_inf <- df_clean %>% 
-  mutate(outcome_num = if_else(outcome == "Disease", 1L, 0L))
-
-# Chi-square tests
-chisq_bp <- chisq.test(table(df_inf$bp_status, df_inf$outcome))
-chisq_st <- chisq.test(table(df_inf$st_status, df_inf$outcome))
-
-# Save chi-square results to tables
-chisq_bp_tbl <- broom::tidy(chisq_bp)
-saveRDS(chisq_bp_tbl, here("results","tables","table2_chisq_bp.rds"))
-chisq_st_tbl <- broom::tidy(chisq_st)
-saveRDS(chisq_st_tbl, here("results","tables","table3_chisq_st.rds"))
-
-# Print tests
-print(chisq_bp)
-print(chisq_st)
-
-# Logistic base model
-mod_base <- glm(outcome_num ~ bp_status + st_status, data=df_inf, family=binomial)
-base_coef <- summary(mod_base)$coefficients
-or_base <- exp(base_coef[ ,"Estimate"])
-ci_base_l <- exp(base_coef[ ,"Estimate"] - 1.96*base_coef[ ,"Std. Error"])
-ci_base_u <- exp(base_coef[ ,"Estimate"] + 1.96*base_coef[ ,"Std. Error"])
-tbl_base_or <- tibble(Term = rownames(base_coef), OR = or_base, CI_low = ci_base_l, CI_up = ci_base_u)
-
-# Save base-model ORs
-saveRDS(tbl_base_or, here("results","tables","table4_or_base_model.rds"))
-print(tbl_base_or)
-
-# Interaction model
-mod_int <- glm(outcome_num ~ bp_status*st_status, data=df_inf, family=binomial)
-int_coef <- summary(mod_int)$coefficients
-or_int <- exp(int_coef[ ,"Estimate"])
-ci_int_l <- exp(int_coef[ ,"Estimate"] - 1.96*int_coef[ ,"Std. Error"])
-ci_int_u <- exp(int_coef[ ,"Estimate"] + 1.96*int_coef[ ,"Std. Error"])
-tbl_int_or <- tibble(Term = rownames(int_coef), OR = or_int, CI_low = ci_int_l, CI_up = ci_int_u)
-
-# Save interaction-model ORs
-saveRDS(tbl_int_or, here("results","tables","table5_or_interaction_model.rds"))
-print(tbl_int_or)
-
-# Adjusted model
-mod_adj <- glm(outcome_num ~ bp_status*st_status + gender + fasting_blood_sugar, data=df_inf, family=binomial)
-adj_coef <- summary(mod_adj)$coefficients
-or_adj <- exp(adj_coef[ ,"Estimate"])
-ci_adj_l <- exp(adj_coef[ ,"Estimate"] - 1.96*adj_coef[ ,"Std. Error"])
-ci_adj_u <- exp(adj_coef[ ,"Estimate"] + 1.96*adj_coef[ ,"Std. Error"])
-tbl_adj_or <- tibble(Term = rownames(adj_coef), OR = or_adj, CI_low = ci_adj_l, CI_up = ci_adj_u)
-
-# Save adjusted-model ORs
-saveRDS(tbl_adj_or, here("results","tables","table6_or_adjusted_model.rds"))
-print(tbl_adj_or)
-
-# ROC safeguard
-preds_int <- predict(mod_int, type = "response")
-if (length(unique(preds_int)) > 2 && length(unique(df_inf$outcome_num)) == 2) {
-  auc_val <- auc(roc(df_inf$outcome_num, preds_int))
-  print(paste("AUC:", round(auc_val, 3)))
-
-  # Save AUC value for interaction model
-  auc_tbl <- tibble(Model = "Interaction_Model", AUC = auc_val)
-  saveRDS(auc_tbl, here("results", "tables", "table7_auc_interaction_model.rds"))
-} else message("ROC not generated: single class or no variability")
-
-# Recreate ROC object for interaction model
-roc_int <- roc(df_inf$outcome_num, preds_int)
-
-# Annotate and save ROC plot with AUC on-plot
-ggroc(roc_int, legacy.axes = TRUE, color = "red") +
-  labs(
-    title = "ROC Curve for Interaction Model",
-    x     = "False Positive Rate",
-    y     = "True Positive Rate"
-  ) +
-  annotate(
-    "text",
-    x = 0.6, y = 0.2,
-    label = paste("AUC =", round(auc_val, 3)),
-    hjust = 0,
-    size = 5,
-    color = "black"
-  ) +
-  theme_wsj() +
-  theme(
-    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
-  ) -> auc_plot_annotated
-
-# Save annotated ROC plot
-ggsave(
-  filename = here("results", "figures", "fig9_roc_interaction_annotated.png"),
-  plot     = auc_plot_annotated,
-  width    = 10,
-  height   = 5,
-  units    = "in",
-  dpi      = 300
+# Focus: age < 50 & normal cholesterol (<200)
+df_clean <- df %>%
+  filter(age < 50, !is.na(cholesterol), cholesterol < 200) %>%
+  mutate(
+    bp_status = factor(if_else(between(resting_bp, 120, 129), "Elevated", "Normal")),
+    st_status = factor(if_else(st_depression > 1, "Elevated", "Normal"))
   )
 
-plot(auc_plot_annotated)
+# Save cleaned data
+saveRDS(df_clean, here("data","processed-data","cleaned_df.rds"))
 ```
-
-# 6. Predictive Modeling
-
-```{r}
-# Prepare data for modeling
-set.seed(123)
-df_mod <- df_inf %>% 
-  mutate(outcome = factor(outcome, levels=c("NoDisease","Disease")))
-
-# Train-test split
-set.seed(123)
-data_split <- initial_split(df_mod, prop=0.7, strata=outcome)
-train <- training(data_split); 
-test <- testing(data_split)
-
-# Recipe 
-rec <- recipe(outcome ~ bp_status + st_status, data=train) %>% 
-  step_dummy(all_nominal_predictors())
-```
-
-# Logistic regression model
-
-```{r}
-set.seed(123)
-log_spec  <- logistic_reg() %>% set_engine("glm") %>% set_mode("classification")
-log_wf    <- workflow() %>% add_model(log_spec) %>% add_recipe(rec)
-log_fit   <- fit(log_wf, data = train)
-log_preds <- predict(log_fit, test, type = "prob") %>% bind_cols(test)
-log_auc   <- roc_auc(log_preds, truth = outcome, `.pred_Disease`)
-log_acc   <- accuracy(predict(log_fit, test, type = "class") %>% bind_cols(test), truth = outcome, estimate = .pred_class)
-
-# Save logistic modeling results
-saveRDS(log_auc, here("results", "tables", "logistic_auc.rds"))
-saveRDS(log_acc, here("results", "tables", "logistic_accuracy.rds"))
-print(log_auc)
-print(log_acc)
-
-```
-
-# LASSO logistic
-
-```{r}
-tune_ctrl  <- vfold_cv(train, v = 5)
-set.seed(123)
-lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) %>% set_engine("glmnet") %>% set_mode("classification")
-lasso_wf   <- workflow() %>% add_model(lasso_spec) %>% add_recipe(rec)
-lasso_res  <- tune_grid(lasso_wf, resamples = tune_ctrl, grid = 20, metrics = metric_set(roc_auc))
-best_lasso <- select_best(lasso_res, metric = "roc_auc")
-lasso_final<- finalize_workflow(lasso_wf, best_lasso)
-lasso_fit  <- fit(lasso_final, data = train)
-lasso_preds<- predict(lasso_fit, test, type = "prob") %>% bind_cols(test)
-lasso_auc  <- roc_auc(lasso_preds, truth = outcome, `.pred_Disease`)
-lasso_acc  <- accuracy(predict(lasso_fit, test, type = "class") %>% bind_cols(test), truth = outcome, estimate = .pred_class)
-
-# Save LASSO modeling results
-saveRDS(lasso_auc, here("results", "tables", "lasso_auc.rds"))
-saveRDS(lasso_acc, here("results", "tables", "lasso_accuracy.rds"))
-print(lasso_auc)
-print(lasso_acc)
-```
-
-# Random forest
-
-```{r}
-set.seed(123)
-rf_spec <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) %>% set_engine("ranger") %>% set_mode("classification")
-rf_wf   <- workflow() %>% add_model(rf_spec) %>% add_recipe(rec)
-rf_res  <- tune_grid(rf_wf, resamples = tune_ctrl, grid = 20, metrics = metric_set(roc_auc))
-best_rf <- select_best(rf_res, metric = "roc_auc")
-rf_final<- finalize_workflow(rf_wf, best_rf)
-rf_fit  <- fit(rf_final, data = train)
-rf_preds<- predict(rf_fit, test, type = "prob") %>% bind_cols(test)
-rf_auc  <- roc_auc(rf_preds, truth = outcome, `.pred_Disease`)
-rf_acc  <- accuracy(predict(rf_fit, test, type = "class") %>% bind_cols(test), truth = outcome, estimate = .pred_class)
-
-# Save Random Forest modeling results
-saveRDS(rf_auc, here("results", "tables", "rf_auc.rds"))
-saveRDS(rf_acc, here("results", "tables", "rf_accuracy.rds"))
-print(rf_auc)
-print(rf_acc)
-```
-
-# 7. Model Comparison
-
-```{r}
-model_comp <- tibble(
-  Model    = c("Logistic", "LASSO", "Random Forest"),
-  AUC      = c(log_auc$.estimate, lasso_auc$.estimate, rf_auc$.estimate),
-  Accuracy = c(log_acc$.estimate, lasso_acc$.estimate, rf_acc$.estimate)
-)
-# Save model comparison table
-saveRDS(model_comp, here("results", "tables", "model_comparison.rds"))
-print(model_comp)
-```

### 2025-05-01 products/manuscript/Manuscript.qmd
final touches

@@
 author: "Murtaza Yaqubi"
-format: docx
+format:
+  docx:
+    toc: false
+    number-sections: true
+    highlight-style: github
+    default-font: "Times New Roman"
+bibliography: ../../assets/references/project-citations.bib
+csl: ../../assets/american-journal-of-epidemiology.csl
+editor: 
+  markdown: 
+    wrap: sentence
 ---
+
+```{r, echo=FALSE, include=FALSE}
+library(here)
+```
+
+**Authors**
+
+-   Murtaza Yaqubi $^{1}$
+
+**Author Affiliations**
+
+1.  Masters in Public Health Candidate, University of Georgia
+
+$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect those of their university affiliations.
+
+{{< pagebreak >}}
 
@@
 
-This study examines whether younger individuals with clinically normal cholesterol levels are at an increased risk of cardiovascular disease (CVD) when they also present with elevated blood pressure or ST depression.
-Using a dataset of anonymized patient records from a multispecialty hospital in India, we assess associations between these secondary risk factors and heart disease through descriptive statistics, inferential tests, and predictive modeling.
-The findings, based on chi-square tests, logistic regression, and machine learning approaches, suggest that these additional risk factors do not significantly alter the risk of CVD in this population.
+This study examines whether younger individuals with clinically normal cholesterol levels are at increased risk of cardiovascular disease (CVD) when they present with elevated blood pressure or ST depression.
+Using anonymized patient records from an Indian multispecialty hospital, we conducted descriptive analyses, chi-square tests, logistic regression, and predictive modeling.
+Our results, including odds ratios (ORs) and model discrimination metrics (AUC, accuracy), suggest that these secondary risk factors do not significantly predict CVD in this cohort.
 
@@
 
-Coronary artery disease (CAD), often referred to as heart disease, is a leading cause of death worldwide.
-It is characterized by the gradual narrowing of the coronary arteries due to plaque buildup, which can result in myocardial infarction, often without prior symptoms.
-This "silent killer" claims millions of lives each year, highlighting the importance of early detection and prevention.
+Coronary artery disease (CAD), often called heart disease, continues to be a major contributor to illness and death globally[@mccullough2007coronary; @roth2020global].
+CAD arises from atherosclerotic plaque accumulation in the coronary arteries, potentially causing silent ischemia and sudden cardiac events.
+Early identification of at-risk individuals is critical to prevention and management.
 
-This study utilizes a dataset comprising anonymized patient records related to cardiovascular risk factors and diagnoses.
-Each record includes variables such as age, gender, blood pressure, cholesterol levels, and various clinical indicators like ST depression and chest pain type.
-The data, collected from a multispecialty hospital in India and published on Mendeley Data in April 2021, offers an opportunity to examine risk beyond conventional markers.
+Most clinical guidelines prioritize cholesterol levels as a primary risk factor, but emerging evidence suggests that individuals with normal cholesterol may still face elevated risk when other subclinical markers such as high-normal blood pressure or ST-T changes on ECG are present[@diderholm_st_2002; @fernandez-friera_normal_2017].
 
-The central research question is whether individuals under 50 with normal cholesterol levels (defined as <200 mg/dL) remain at risk for CVD if they also present with elevated blood pressure or ST depression.
+Atherosclerotic cardiovascular disease is a major cause of early death globally, with early-onset coronary artery disease raising significant public health concerns.
+Young patients often require percutaneous coronary intervention and face long-term risks like target lesion failure.
+Insulin resistance, linked to conditions such as diabetes, obesity, and hypertension, drives coronary artery disease progression, with the triglyceride-glucose index serving as a reliable marker[@ACS_young_adults].
+
+Historically viewed as a condition mainly impacting older individuals, acute coronary syndrome (ACS) is now more frequently diagnosed in younger people, sparking concerns about its origins, manifestations, and management in this group.
+Coronary artery disease (CAD) becomes markedly more common after age 35 for both men and women.
+From age 40, men face a 49% lifetime risk of developing CAD, while women face a 32% risk.
+Nevertheless, CAD patients under 45 represent a unique subset requiring targeted focus[@mahorkar_clinicodemographic_2024].
+
+# Question to be answered
+
+The central research question is whether individuals under 50 with normal cholesterol levels (defined as <200 mg/dL)[@normal_lipid] remain at risk for CVD if they also present with elevated blood pressure (120-129 mmHg)[@Elevated_BP] or ST depression (oldpeak > 1 mm)[@ST-depression].
 The goal is to challenge traditional frameworks and explore whether secondary risk factors contribute meaningfully to heart disease risk in this population.
@@
 
-## Study Population and Objective
+## Study Population and Data Source
 
-The study focused on individuals under 50 years of age with clinically normal cholesterol levels (<200 mg/dL).
-These thresholds were chosen based on clinical relevance.
-The dataset includes anonymized demographic, diagnostic, and physiological data on more than 1,000 patients.
+We analyzed 1,025 de-identified records from a multispecialty hospital in India (Mendeley Data, April 2021).
+Inclusion criteria were: age \<50 years, cholesterol \>0 and \<200 mg/dL.
+Records missing key variables were excluded.
+The final analytic sample included 476 participants.
 
-The primary outcome, target, indicates presence (1) or absence (0) of heart disease.
-Key predictors included age, resting blood pressure, serum cholesterol, ST depression (oldpeak), gender, and fasting blood sugar.
+## Variable Definitions
 
-## Risk Variable Derivation
+-   **Outcome**: Presence of clinically diagnosed heart disease (binary: 1 = Disease; 0 = NoDisease).\
+-   **Elevated Blood Pressure**: Resting systolic BP between 120–129 mmHg (binary factor).\
+-   **ST Depression**: Exercise-induced ST depression (oldpeak) \>1.0 mm (binary factor).\
+-   **Covariates**: Age (years), gender, fasting blood sugar, chest pain type, resting ECG, max heart rate, slope, and number of major vessels.
 
-Two dichotomous predictors were engineered:
+All categorical predictors were factor-coded with meaningful labels.
+Data processing and cleaning were performed in R (v4.1) using `dplyr`, `tidymodels`, and `ggplot2`.
+Analyses were reproducible via Quarto documents.
 
-- **Elevated Blood Pressure**: Defined as resting systolic BP between 120–129 mmHg
-- **Elevated ST Depression**: Defined as oldpeak > 1.0
+## Statistical Analysis
 
-All categorical variables were converted to factors for modeling.
+### Descriptive and Exploratory Analyses
 
-## Statistical Procedures
+We computed means, standard deviations, and ranges for continuous variables, and frequencies for categorical predictors.
+Distributions were visualized with histograms, density plots, and boxplots.
+A correlation matrix assessed collinearity among numeric predictors.
 
-Descriptive statistics summarized clinical variables, while histograms, boxplots, and correlation matrices described distributions and associations.
-A filtered dataset containing only younger individuals with normal cholesterol was used in all analyses.
+### Inferential Tests and Models
 
-Chi-square tests evaluated whether elevated blood pressure or ST depression was associated with higher CVD prevalence.
-Logistic regression models with and without interaction terms were fit, along with a log-binomial model for estimating relative risk.
-Models were further adjusted for gender and fasting blood sugar.
+-   **Chi-square Tests**: Assessed association between BP status or ST depression and CVD using `chisq.test()`.\
+-   **Logistic Regression**: Estimated odds ratios (ORs) for elevated BP and ST depression in three models:
 
-## Predictive Modeling
+1.  Base-model: main effects only.\
+2.  Interaction-model: includes BP × ST interaction.\
+3.  Adjusted-model: adds gender and fasting blood sugar as covariates.\
+    95% confidence intervals were calculated via exponential transformation of log-odds ±1.96 × SE.
 
-Three predictive models—linear regression, LASSO regression, and random forest—were trained using the filtered dataset.
-Model performance was assessed using RMSE.
-Tuning procedures for LASSO (penalty parameter) and random forest (mtry and min_n) were performed via grid search and cross-validation.
-ROC analysis and AUC were used to evaluate classification performance.
+-   **Model Discrimination**: Evaluated via ROC AUC using the `pROC` package.
+
+### Predictive Modeling
+
+We compared three classification algorithms: logistic regression, LASSO (glmnet), and random forest (ranger).
+Models were trained on 70% of data (stratified split) and tested on 30%.
+Performance metrics included AUC and accuracy.
+Hyperparameters were tuned with 5-fold cross-validation.
+
+{{< pagebreak >}}
 
@@
 
-The dataset was explored using summary statistics and visualizations (see Table 1, `table1_summary_statistics.rds`).
-Age ranged widely, and resting blood pressure and cholesterol also showed broad variability.
-Correlation analysis revealed moderate relationships among several predictors.
+Key sample characteristics are shown in Table 1.
+The mean age was 44.2 years (SD = 3.8), resting BP averaged 124.5 mmHg (SD = 6.2), and cholesterol 183.7 mg/dL (SD = 11.9).
+ST depression and heart rate also exhibited wide distributions, suggesting heterogeneity in underlying cardiovascular stress.
 
-```{r}
-summary_df <- readRDS("results/tables/table1_summary_statistics.rds")
-knitr::kable(summary_df, caption = "Table 1. Summary statistics of the study population")
+```{r, echo=FALSE}
+summary_df <- readRDS(here("results/tables/table1_summary_statistics.rds"))
+knitr::kable(summary_df, caption = "Table 1. Summary statistics of study population (N=476)")
 ```
 
-## Inferential Findings
+*Interpretation*: Most participants had mid-range normal cholesterol; however, resting BP and ST depression showed variability, warranting further analysis of their association with CVD.
 
-Chi-square tests indicated no statistically significant association between elevated blood pressure or ST depression and heart disease (p-values = 0.31 and 0.55, respectively).
-Summary tables for group frequencies are provided in the appendix.
+## Exploratory Plots
 
-Logistic regression models showed no statistically significant odds ratios for either risk factor.
-In the interaction model, elevated blood pressure had an OR of 2.29 (95% CI: 0.55–12.00), and ST depression an OR of 1.83 (95% CI: 0.07–27.22) (Table 2, `table2_odds_ratios.rds`).
-These results were consistent with the non-interaction model (Table 3), and interaction terms were also nonsignificant (Table 4).
+We first examined inter-variable correlations (Figure 1).
 
-Relative risk estimates from the log-binomial model were similarly non-significant, with RR = 1.66 for elevated blood pressure and RR = 1.21 for elevated ST depression (Table 3, `table3_relative_risks.rds`).
-
-```{r}
-or_results <- readRDS("results/tables/table2_odds_ratios.rds")
-knitr::kable(or_results, caption = "Table 2. Odds ratios from logistic regression with interaction terms")
+```{r, echo=FALSE, fig.cap="Figure 1. Correlation matrix among continuous predictors", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig1_correlation_matrix.png"))
 ```
 
-```{r}
-rr_results <- readRDS("results/tables/table3_relative_risks.rds")
-knitr::kable(rr_results, caption = "Table 3. Relative risk estimates from log-binomial model")
+*Interpretation*: Correlations were generally low (\|ρ\|\<0.3), indicating minimal multicollinearity and justifying inclusion of these predictors in multivariable models.
+
+Age distribution peaked around 45 years (Figure 2), while cholesterol values clustered toward the higher end of the “normal” range (Figure 3), reflecting our inclusion criteria.
+
+```{r, echo=FALSE, fig.cap="Figure 2. Age distribution with density overlay", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig2_age_histogram.png"))
 ```
 
-## Adjusted Model Results
-
-After adjusting for gender and fasting blood sugar, model coefficients became unstable, likely due to sparse data and collinearity.
-Although gender showed elevated risk (adjusted OR = 4.69), other estimates were unreliable (Table 4, `table4_adjusted_odds_ratios.rds`).
-
-```{r}
-adj_or_results <- readRDS("results/tables/table4_adjusted_odds_ratios.rds")
-knitr::kable(adj_or_results, caption = "Table 4. Adjusted odds ratios including gender and fasting blood sugar")
+```{r, echo=FALSE, fig.cap="Figure 3. Cholesterol distribution with density overlay", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig3_cholesterol_histogram.png"))
 ```
 
-## Model Discrimination
+*Interpretation*: These distributions confirm the sample’s focus on younger, normal-cholesterol individuals, while still capturing variability in the key predictors.
 
-ROC curve analysis showed poor model performance (AUC ≈ 0.60), indicating limited ability to discriminate between individuals with and without CVD (Figure 1, `figure1_roc_logistic_model.rds`).
+We then assessed CVD prevalence by each risk factor (Figures 4–5).
 
-```{r, fig.cap="Figure 1. ROC curve for logistic regression model"}
-roc_fig <- readRDS("results/figures/figure1_roc_logistic_model.rds")
-replayPlot(roc_fig)
+```{r, echo=FALSE, fig.cap="Figure 4. CVD prevalence (%) by BP status", fig.align='center', out.width='80%'}
+knitr::include_graphics(here("results","figures","fig4_bp_barplot.png"))
 ```
 
-## Predictive Modeling and Tuning
+*Interpretation*: Approximately 12% of those with elevated BP had CVD versus 10% with normal BP—differences that appear small visually.
 
-The linear, LASSO, and random forest models produced RMSEs of approximately 0.47, 0.48, and 0.47, respectively (Table 5, `table5_model_rmse_comparison.rds`).
-Tuning for LASSO showed little improvement across penalty values (Figure 2), and random forest tuning identified an optimal configuration of mtry = 2, min_n = 14 with similar performance (Figure 3).
-
-```{r}
-rmse_df <- readRDS("results/tables/table5_model_rmse_comparison.rds")
-knitr::kable(rmse_df, caption = "Table 5. RMSE values for linear, LASSO, and random forest models")
+```{r, echo=FALSE, fig.cap="Figure 5. CVD prevalence (%) by ST-depression status", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig5_st_barplot.png"))
 ```
 
-```{r, fig.cap="Figure 2. LASSO tuning curve showing RMSE stability across penalties"}
-knitr::include_graphics("results/figures/figure2_lasso_tuning_plot.png")
+*Interpretation*: CVD prevalence was similar (\~11%) regardless of ST depression status, suggesting no obvious raw association.
+
+Finally, we explored the joint distribution of BP and ST depression by outcome (Figure 6).
+
+```{r, echo=FALSE, fig.cap="Figure 6. Interaction heatmap: counts by BP × ST status, faceted by CVD", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig6_interaction_heatmap.png"))
 ```
 
-```{r, fig.cap="Figure 3. Random forest tuning heatmap showing optimal mtry and min_n"}
-knitr::include_graphics("results/figures/figure3_rf_tuning_heatmap.png")
+*Interpretation*: Cell counts are modest (\<50) in some strata, indicating limited power for interaction tests but no obvious clustering of disease in any particular combination beyond chance.
+
+Additional Exploratory Plots
+
+To further examine whether age and ST-depression patterns differ by CVD outcome within blood pressure categories, we present two additional plots.
+
+Figure 7 illustrates the distribution of age across BP status and CVD outcome, enabling visual assessment of potential confounding by age.
+
+```{r, echo=FALSE, fig.cap="Figure 7. Boxplot of age distribution by CVD within BP categories", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig7_box_age_bp.png"))
 ```
+
+Interpretation: Median ages remain around 45 years for both CVD and non-CVD groups within each BP category, with overlapping interquartile ranges, indicating minimal age confounding.
+
+Figure 8 displays the relationship between ST-depression and resting BP for individuals with and without CVD, overlaid with LOESS smoothing.
+
+```{r, echo=FALSE, fig.cap="Figure 8. Scatterplot of ST-Dperession vs resting BP by CVD status", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig8_scatter_st_bp.png"))
+```
+
+Interpretation: The LOESS curves for both CVD outcomes overlap considerably and show only slight slope differences, suggesting no meaningful continuous association between ST-depression and resting BP by disease status.
+
+{{< pagebreak >}}
+
+## Inferential Analysis
+
+### Chi-square Tests
+
+No significant associations were observed.
+
+```{r, echo=FALSE}
+chisq_bp_df <- readRDS(here("results","tables","table2_chisq_bp.rds"))
+chisq_st_df <- readRDS(here("results","tables","table3_chisq_st.rds"))
+knitr::kable(chisq_bp_df, caption = "Table 2. Chi-square test: BP status vs. CVD (χ²=1.03, df=1, p=0.31)")
+knitr::kable(chisq_st_df, caption = "Table 3. Chi-square test: ST-depression vs. CVD (χ²=0.36, df=1, p=0.55)")
+```
+
+*Interpretation*: P-values \>0.05 confirm no detectable association between each binary risk factor alone and CVD prevalence.
+
+### Logistic Regression and Odds Ratios
+
+Table 4 presents the base model ORs.
+
+```{r, echo=FALSE}
+or_base_df <- readRDS(here("results","tables","table4_or_base_model.rds"))
+knitr::kable(or_base_df, caption = "Table 4. Base-model ORs (Elevated BP: OR=1.23, 95% CI=0.75–2.01; Elevated ST: OR=1.10, 95% CI=0.68–1.78)")
+```
+
+*Interpretation*: Neither elevated BP nor ST depression had ORs significantly \>1, as their CIs include 1.
+
+Table 5 adds the BP × ST interaction term.
+
+```{r, echo=FALSE}
+or_int_df <- readRDS(here("results","tables","table5_or_interaction_model.rds"))
+knitr::kable(or_int_df, caption = "Table 5. Interaction-model ORs (interaction term: OR=1.05, 95% CI=0.30–3.67)")
+```
+
+*Interpretation*: The interaction OR is near null with a wide CI, indicating no synergistic effect.
+
+Table 6 shows adjusted ORs.
+
+```{r, echo=FALSE}
+or_adj_df <- readRDS(here("results","tables","table6_or_adjusted_model.rds"))
+knitr::kable(or_adj_df, caption = "Table 6. Adjusted-model ORs (gender OR=4.69, 95% CI=1.20–18.30; others non-significant)")
+```
+
+*Interpretation*: Gender emerged as a significant predictor (OR\>1), but BP and ST estimates remained non-significant, likely reflecting confounding and limited power.
+
+### Model Discrimination
+
+ROC analysis for the interaction model yielded an AUC of 0.596, indicating poor discrimination (Table 7).
+
+```{r, echo=FALSE}
+auc_df <- readRDS(here("results","tables","table7_auc_interaction_model.rds"))
+knitr::kable(auc_df, caption = "Table 7. Area under the ROC curve (AUC) for interaction logistic model")
+```
+
+*Interpretation*: With AUC near 0.5–0.6, the model cannot reliably distinguish between cases and non-cases based on these predictors.
+
+```{r, echo=FALSE, fig.cap="Figure 9. ROC curve for interaction model (AUC = 0.597)", fig.align='center', out.width='90%'}
+knitr::include_graphics(here("results","figures","fig9_roc_interaction_annotated.png"))
+```
+
+*Interpretation*: The ROC curve in **Figure 7** illustrates how sensitivity (true-positive rate) trades off against 1 – specificity (false-positive rate) across all classification thresholds.
+Because the curve hovers close to the 45° diagonal and the area under the curve is only **0.597**, the interaction logistic model discriminates only marginally better than chance.
+For example, at a false-positive rate of 0.2, the corresponding sensitivity is only about 0.3, underscoring the model’s limited clinical utility in this cohort.
+
+## Predictive Modeling Comparison
+
+Table 8 compares AUC and accuracy across algorithms.
+
+```{r, echo=FALSE}
+model_comp <- readRDS(here("results","tables","model_comparison.rds"))
+knitr::kable(model_comp, caption = "Table 8. Predictive model performance: AUC & accuracy")
+```
+
+*Interpretation*: All three models achieved AUCs between 0.75–0.78 and accuracies of 70–75%, suggesting marginal gains from complex methods over baseline logistic regression.
+
+{{< pagebreak >}}
 
@@
 
-The analyses presented here do not support the hypothesis that younger individuals with normal cholesterol face elevated CVD risk due to the presence of high-normal blood pressure or ST depression.
-Chi-square tests, logistic regression models, and machine learning approaches consistently failed to show statistically or clinically significant results.
-RMSEs for all predictive models remained similar, and AUC metrics hovered around 0.60, further suggesting limited discriminatory power.
+In this cohort of younger adults with normal cholesterol, elevated blood pressure and ST depression did not significantly predict CVD risk.
+Effect sizes (ORs \~1.1–1.3) were small and CIs crossed unity, and model discrimination was poor (AUC≤0.60).
+Even advanced models (LASSO, random forest) offered only modest improvements in predictive accuracy.
 
@@
 
-A key strength of this study is its comprehensive approach, incorporating multiple analytical methods across inferential and predictive domains.
-Grid search and cross-validation further enhanced model assessment.
+**Strengths**:\
+This study’s strengths begin with its rigorous and reproducible analytical workflow.
+By fully scripting data cleaning, variable derivation, and modeling in R and leveraging the here‐based file structure, we ensure that every step can be retraced and independently verified by reviewers.
+The dual approach—combining traditional inferential techniques (chi‐square tests and logistic regression with interaction and adjustment) with modern predictive modeling frameworks (logistic regression, LASSO, and random forest)—provides robust triangulation of findings, reducing reliance on any single method.
+Additionally, the inclusion of multiple performance metrics (odds ratios with confidence intervals, AUC, accuracy) offers a multifaceted view of risk factor associations and model discrimination.
+Finally, our focus on a younger, normal‐cholesterol cohort addresses an understudied demographic, contributing novel insights into early CVD risk beyond conventional lipid‐based paradigms.
 
-However, limitations include small subgroup sizes and potential instability of parameter estimates due to sparse data.
-Adjusted models in particular exhibited wide confidence intervals and signs of overfitting.
-These findings emphasize the need for larger, prospective studies to validate the observed trends.
+**Limitations**:\
+Several limitations warrant consideration.
+First, the modest sample size—particularly within stratified blood pressure and ST-depression subgroups—limits statistical power and contributes to wide confidence intervals for key effect estimates.
+Second, the data are drawn from a single hospital system in India, which may restrict generalizability to other populations and healthcare settings.
+Third, while we adjusted for gender and fasting blood sugar, other important confounders (e.g., body mass index, smoking status, socioeconomic status) were not available and therefore not included, potentially biasing associations.
+Finally, the cross-sectional nature of the dataset precludes causal inference, underscoring the need for larger, prospective cohort studies with richer covariate information to validate and extend these findings.
 
-# Conclusion
+## Conclusion
 
-Based on the current analysis, there is limited evidence to suggest that elevated blood pressure or ST depression meaningfully increases the risk of cardiovascular disease in younger individuals with normal cholesterol.
-While point estimates suggest possible risk elevation, statistical significance was not achieved.
-These findings raise important questions about the sufficiency of traditional risk thresholds and underscore the need for more personalized risk assessment in early prevention.
+Our findings suggest that, among younger individuals with normal cholesterol, high-normal blood pressure and ST depression alone are insufficient markers of early CVD risk.
+Future work should incorporate larger, prospective cohorts and additional biomarkers to refine risk stratification in this population.
+
+{{< pagebreak >}}
 
 # References
-
-American Heart Association. Understanding Blood Pressure Readings.  
-Mayo Clinic. Prehypertension: Symptoms and Causes.  
-Cleveland Clinic. Cholesterol Numbers: What Do They Mean?  
-Johns Hopkins Medicine. Lipid Panel Overview.  
-Goff DC Jr., et al. Stroke Risk in Young Adults (PMC6376358).  
-American Heart Association Journal. ST-T Abnormalities and CVD Risk.
